### 2015-04-15 manuscript/manuscript.Rmd
Manuscript dated Feb 17

---
title: Crowdsourced geometric morphometrics enable rapid large-scale collection and
  analysis of phenotypic data
author:
- affiliation: ucla
  email: jonathan.chang\@ucla.edu
  footnote: Author for correspondence
  name: Jonathan Chang
- affiliation: ucla
  name: Michael E. Alfaro
bibliography:
- ../../Mendeley/library.bib
- knitcitations.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: xelatex
    template: inc/mstemplate.tex
    toc: yes
    toc_depth: 1
  html_document:
    toc: yes
  word_document: default
documentclass: inc/msarticle
fontsize: 11pt
geometry: margin=1.2in
layout: preprint
lineno: yes
linestretch: 1.05
csl: inc/evolution.csl
address:
- address: |
    Department of Ecology and Evolutionary Biology, University of California, Los Angeles, CA, USA
  code: ucla
---

```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
library(MASS)
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(magrittr)
library(tidyr)
library(reshape2)
library(scales)
library(grid)
library(geomorph)
library(rjson)
library(ipred)
library(directlabels)
library(BAMMtools)
library(knitcitations)
library(kfigr)
options("citation_format"="pandoc")
cleanbib()
```

# Abstract

1. Advances in genomics and informatics have enabled the production of large phylogenetic trees. However, the ability to collect large phenotypic datasets has not kept pace. Here, we present a method to quickly gather geometric morphometric data using crowdsourced image-based landmarking.

2. We show that crowdsourced workers perform equally well as experienced morphologists on the same tasks. We also demonstrate the speed of our method on 7 families of Actinopterygii fishes.

3. Our methods are publicly available as open source software.

# Introduction

The integration of phenotypic data, such as anatomy, behavior, physiology, and other traits, with phylogenies has been a powerful tool for investigating the patterns of biological evolution. Recent advances in next-generation sequencing [@Shendure2008; @Meyer2008] and sequence capture technologies [@Faircloth2012b; @Lemmon2012] have made phylogenetic inference of large radiations of organisms possible [@McCormack2012; @McCormack2013; @Faircloth2013; @Faircloth2014].
These large phylogenies have been successfully used in conjunction with existing comprehensive phenotypic datasets to answer key questions about the mode and tempo of diversifiaction and morphological evolution [@Rabosky2013].

However, similar breakthroughs for generating new phenotypic datasets have been comparatively uncommon, likely due to the high expense and effort required [reviewed in @Burleigh2013]. Creating these large phenotypic datasets has generally required an extended dedicated effort of measuring and describing morphological or behavioral traits that are then coded into a comprehensive data matrix. One such example is the Phenoscaping project [<http://kb.phenoscape.org>; @Deans2015], and its complementary efforts in the Vertebrate Taxonomy Ontogeny [@Midford2013] and Hymenoptera Anatomy Ontology [@Yoder2010], which requires large amounts of researcher effort to compile. Other approaches include using machine learning [@Mckinney2013a] or natural language processing [@Cui2012a] to identify or infer phenotypes. These statistical techniques function ideally with either a large training dataset or a complex model [@Brill2003; @Halevy2009; @Hastie2009], both of which also require intensive researcher effort to build and validate. Finally, methods such as high-throughput infrared imaging, mass spectrometry, and chromatography have been successfully used in plant physiology [@Furbank2011]
 and microbiology [@Skelly2013a], but these methods may not be applicable for zoological researchers.
These bottlenecks in the collection of phenotypic data has limited the taxonomic breadth and depth of research in comparative biology.

Here we present a method and toolkit to efficiently collect two-dimensional geometric morphometric phenotypic data at a "phylogenomic" scale. 
We developed a novel web browser-based image landmarking application, and use Amazon Mechanical Turk (<https://www.mturk.com>) to distribute digitization tasks to remote workers (hereafter *turkers*) over the Internet, who are paid for their contributions.
We also evaluate the accuracy and precision of turkers by assigning identical image sets and digitization protocols to users who are experienced with fish morphology (hereafter *experts*), and compare the inter- and intra-observer differences between turkers and experts.
Finally we demonstrate the utility of this new method by constructing a phylogenetic analysis pipeline, and use this pipeline to download fish photographs from the web, collect Mechanical Turk shape results, and analyze the rate of body shape evolution using BAMM [@Rabosky2014a].

# Methods

## Amazon Mechanical Turk

Amazon Mechanical Turk ("MTurk") is a web-based service where Requesters can request work, known as Human Intelligence Tasks ("HITs") to be performed by Workers. Workers work from home and submit the tasks over the Internet, where Requesters review it, and, if they are satisfied with the results, accept the work and pay the Worker. We use MTurk as a platform to distribute our geometric morphometric tasks and financially compensate the worker accordingly. Scientific collection of data over MTurk and similar services has generally been limited to the fields of psychology and computer science, and there have been few attempts to crowdsource biological trait data [but see @Burleigh2013].

## Web-based geometric morphometrics

![A screenshot of the web app that turkers used to digitize images. A live demonstration is available at <https://jonchang.github.io/eol-mturk-landmark/>](interface.png)

We developed a cross-browser web application that runs completely on the client. This simplifies the infrastructure challenge of needing to serve many crowdsdourced workers simultaneously. The web app is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task. Below is an example JSON file that demonstrates the utility of our web app:

```json
{
    "C": {
        "kind": "point",
        "help": "Click the center of the eye."
    },
    "D": {
        "kind": "line",
        "help": "Click and drag from the left edge of the eye
                 to the right edge of the eye."
    },
    "O": {
        "kind": "curve",
        "help": "Click and drag over the outline of the eye,
                 starting from the leftmost point of the eye."
    }
}
```

Each digitization task to be completed is given a short abbreviation to aid task identification ("C" for center, "D" for diameter, and "O" for outline), and the type of task, "point", "line" or "curve", for homologous landmarks, linear measurements, and sliding semilandmarks can be specified. There is also an optional short help snippet that is automatically displayed to inform users what to do. These inline help texts complement a larger and more detailed protocol document that workers are required to read before beginning work (Supplementary Material).

## Server-controlled experiment

Although digitizing and landmarking a single image [microtasks *sensu* @Good2013] is most effective for high-throughput work on MTurk, it is not suited towards conducting controlled experiments. To solve this issue we created a server-side front end that mimiced an official Amazon Mechanical Turk endpoint, but would also automatically distribute tasks in the order required by the experiment. The MTurk facsimile checkpoints each completed image and distributes another from a Requester-defined set of images. A configuration file allows the experimenter to vary the number of replicates required for the given experiment. There is also functionality for non-MTurk workers to contribute data using the same mechanism. The software is open source and uses PHP and SQLite, and can be installed on nearly all web hosting services (<https://github.com/jonchang/fake-mechanical-turk>).

## Reliability analysis

Collecting landmark-based geometric morphometric data at scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007]. To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk were significantly better or worse than traditionally-collected data, we asked both turkers and experts to landmark a set of 5 fish images, 5 times each. All participants used the same protocol and same software to digitize the same set of fishes.

The landmarks were chosen according to previously-published literature concerning fish shape [@Chakrabarty2005; @Frederich2008], and included a range of landmarks that would be relatively easy to digitize (e.g., the anterior margin of the eye) and ones that would be challenging (the opercular series). A discussion of the performance of turkers versus morphlogists on a per-landmark basis is included in the supplemental material.

We used multiple measurements per worker per image to quantify measurement error within and among observers [@Palmer1986] using *geomorph* `r citep(citation("geomorph"))`. To determine whether turkers and experts were statistically distinguishable, we used the machine learning technique linear discriminant analysis [LDA, @Ripley1996] as implemented in *MASS* `r citep(citation("MASS"))`. A common error in assessing the efficacy of machine learning methods is training the model and then testing the model using the same data set for both. The researcher can make the model increasingly sophisticated to better fit the training dataset, but be unable to make accurate inferences to real-world data, due to overfitting. Conversely, withholding too much data for validation results in underfitting and the inability of the model to generalize its predictions to real data.

To assess the ability of LDA to distinguish between these two groups, we used $K$-fold cross-validation (CV), which splits our data into $K$ equally-sized groups, using $K-1$ for training and $1$ for validation, and testing the model for all possible permutations of these groups. The special case of $K=N$ groups is also known as leave-one-out CV. We used a value of $K=10$ to validate the LDA model. The statistical literature typically recommends $K=5$ or $K=10$ for its balanced approach to the bias-variance tradeoff [@Kohavi1995; @Hastie2009]. All analyses were performed in R `r citep(citation())`. As an alternate model we also used quadratic discriminant analysis (QDA).

We calculated the per-digitizer median shape for each image used, as well as the consensus turker and morphologist landmarks, and projected these shapes into Procrustes space. We calculated the 95% confidence ellipse to estimate the uncertainty around these shape estimates.

## Phylogenetic analysis on 7 families

To demonstrate the utility and how quickly we can obtain comparative data using this method, we downloaded previously-published phylogenies for 7 fish families (Acanthurdiae, Apogonidae, Balistidae, Chaetodontidae, Labridae, Pomacentridae, and Tetraodontidae) from TreeBASE and Dryad (XXX: CITE TREES) and match them to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using the Reol interface to the EOL API `r citep(citation("Reol"))`. If there were multiple images for a given species we manually selected the best one. Crowdsourced workers collected body shape data following a standard protocol (Supplementary Material). We then determined the major components of shape variation using a principal components analysis (PCA) as implemented in the R package *geomorph* `r citep(citation("geomorph"))`.

We used Bayesian Analysis of Macroevolutionary Mixtures [*BAMM*, @Rabosky2014a] to estimate the diversification rate as well as the rate of morphological evolution, using the first six principal components for each of the 7 families. BAMM estimates the location of rate shifts in either diversification or character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method, where the number of rate regimes on a phylogeny is allowed to vary. We assessed convergence and mixing using CODA `r citep(citation("coda"))`, BAMMtools `r citep(citation("BAMMtools"))` and Tracer [@rambaut2007tracer]. We then used BAMMtools to identify groups in each family that experienced increased rates of morphological evolution.

# Results

## Reliability analysis

```{r cv_table, echo=F, results="asis", anchor="table"}
ip.lda <- function(object, newdata) predict(object, newdata=newdata)$class

no_st <- fread("anon.csv") %>% tbl_df() %>% filter(role != "student") %>% mutate(role=factor(role), mark=factor(mark))
subsamp <- group_by(no_st, fn, role) %>% sample_n(size=400) %>% ungroup()

CV <- subsamp %>% group_by(fn) %>% do(estimator = errorest(role ~ mark:x + mark:y, data=., model=lda, estimator="cv", est.para = control.errorest(k=10), predict=ip.lda))

CV_sum <- CV %>% summarise(fn=fn, error=estimator[["error"]]) %>% separate("fn", "fn", sep="_", extra="drop")

# format for display
CV_sum %>% transmute(`Family`=fn, `Misprediction rate`=error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis with 10-fold cross validation for each fish image.")
```

The linear discriminant analysis (LDA) with 10-fold cross validation (Table `r figr("cv_table")`) was unable to reliably distinguish between morphologists and crowdsourced workers. The quadratic estimator (QDA) performed similarly poorly (results not shown).

```{r turk_morph_procrustes_space, echo=FALSE, fig.cap="Hello", fig.keep="last", anchor="figure", message=FALSE}

# Load data from the anonymized results file, remove students, drop bad marks
df <- fread("anon.csv") %>% tbl_df() %>%
	filter(role != "student") %>% 
	separate(col=mark, into=c("cat", "cat_seq"), sep=1, remove=F, convert=T) %>%
	mutate(round=floor(sequence/5), fn=factor(str_extract(fn, ".{8}"))) %>%
	filter(!(fn %in% c("Balistid") & mark %in% c("D1","D2")), !mark %in% c("O1","O2","O3","O4","O5", "A3", "D3", "D4"))

# grabs the consensus sequence by person and fish
df <- df %>% group_by(fn, mark, role, sha1mac) %>% summarise(x=median(x), y=median(y))

format_for_gpagen <- function(df) {
	splut <- gather(df, variable, val, x:y) %>% unite(col=id, fn, role, sha1mac)
	xtabs(val ~ mark + variable + id, data=splut)
}

coords <- gpagen(format_for_gpagen(df), ShowPlot=F)$coords

real_plot_tangent_space <- function(coords, axis1="PC1", axis2="PC2") {
  x <- two.d.array(coords)
  pc.res <- prcomp(x)
  pcdata <- pc.res$x

	res <- as.data.frame(pcdata)
	res$id <- rownames(res)
	res2 <- res %>% separate(id, into=c("fn", "role","sha"), sep="_") %>% tbl_df()

	# consensus in Procrustes space for downstream highlighting
	dots <- list(
		x=lazyeval::interp(~median(axis1, na.rm=T), axis1=as.name(axis1)),
		y=lazyeval::interp(~median(axis2, na.rm=T), axis2=as.name(axis2)))
	consensus <- res2 %>% group_by(fn, role) %>% summarise_(.dots=dots)

	p <- ggplot(res2, aes_string(axis1, axis2, group="fn")) + geom_point(aes(col=fn, shape=role)) + geom_point(aes(x, y), data=consensus, alpha=0.5, size=5, col="black") + geom_point(aes(x, y, col=fn, shape=role), data=consensus) + stat_ellipse(aes(color=fn), linetype=2, type="norm", level=0.95) + theme_minimal() + theme(legend.position="none")

	direct.label(p, method="smart.grid")
}

real_plot_tangent_space(coords, "PC1", "PC2")
```

Projecting consensus shapes into Procrustes space (Figure `r figr("turk_morph_procrustes_space")`) shows that within-group variation exceeds that of among group variation. The consensus turker shape and consensus expert shape are nearly identical for 5 of 7 families on PC1 and PC6 (Figure `r figr("turk_morph_procrustes_space")`, S1 and S2).

## Phylogenetic analysis on 7 families

The full data analysis pipeline end-to-end took 2.2 hours of wall-clock time. The BAMMtools analysis identified areas of each phylogeny that were experiencing faster rates of body shape evolution. XXX: Talk about these in more detail.

```{r bamm_plots, echo=F, message=F, warning=F, anchor="figure", cache=T}
results <- fread("fishturk.results") %>% tbl_df() %>% filter(Answer.marks != "") %>% mutate(tip=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 3], family=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 2], Answer.marks=str_replace_all(Answer.marks, '""', '"'))

get_marks <- function(blob, ...) {
  fromJSON(blob) %>% as.data.frame %>% t %>% as.data.frame %>% set_colnames(c("x", "y")) %>% mutate(mark=rownames(.)) %>% cbind(...)
}

allmarks <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=.$workerid, family=.$family, tip=.$tip))

df <- allmarks %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))

splut <- gather(df, variable, val, x:y) %>% unite(id, family, tip, sep="|")

res <- xtabs(val ~ mark + variable + id, data=splut)

groups <- str_match(dimnames(res)$id, "(.*)\\|")[, 2] %>% factor %>% as.numeric

full_coords <- gpagen(res, ShowPlot=F)$coords

res <- plotTangentSpace(coords, groups=groups, axis1=1, axis2=2, verbose=T, warpgrids=F)$pc.scores

res <- as.data.frame(res)
res$id <- rownames(res)
res2<-res %>% separate(id, into=c("family", "name"), sep="\\|", extra="drop") %>% tbl_df()

p <- ggplot(res2, aes(x=PC1, y=PC2, col=family)) + geom_point() + theme(legend.position="none")

res2 %>% select(name, PC1) %>% write.table(file="lol.txt", quote=F, row.names=F, col.names=F, sep="\t")
```

# Discussion

Previous work in generating large phenotypic datasets typically relied on one or more of expertise, effort, and computational power.
Collecting phenotypic data remains a large bottleneck in our ability to comprehensively test hypotheses about large radiations of organisms. We developed a method to gather phenotypic data using Amazon Mechanical Turk.
Our results show that the quality of turkers' data are not significantly different from results collected from experts, and that these data can be gathered extremely quickly.

We attempted to distinguish between turker-generated landmarks and expert-generated landmarks using linear discriminant analysis. Even the image with the lowest misprediction rate, from the family `r filter(CV_sum, min(error) == error)$fn`, still incorrectly predicted the digitizer's identity `r filter(CV_sum, min(error) == error)$error %>% as.numeric() %>% scales::percent()` of the time, suggesting that, given a set of landmarks, it is difficult to reliably predict whether a turker or a morphologist created them.

Our novel pipeline to download images, upload them to Amazon MTurk, and process them using BAMM and BAMMtools showcases the ability to rapidly collect phenotypic data. The majority of the 2.2 hours of wall-clock time to collect these data were spent on waiting for worker results; however, a majority of the data had already been collected at the 1-hour mark. An online methodology could conceivably improve on this analysis time, by iteratively refining its results as new data streamed in from Amazon's servers. (BAMM does not support adding new data mid-analysis). Other speedups could be found, particularly in identifying the best image for each species to use for analysis. This could easily be replaced with a protocol for image selection and uploaded to MTurk for categorization.

The results of this pipeline also serve to test previous predictions of the rates of cladogenic speciation and mophological evolution [@Rabosky2013]. Body shape and body size are usually correlated (XXX: find the Cite), and the first PC axis of body shape analyses is typically associated with features related to size. (XXX: make an extensive comparision between the results of the nat comm paper and the 7 families here)

The traditional way of collecting phenotypic data involves enormous researcher effort and significant morphological expertise. For example, @Brusatte2014 used a 853 character discrete character matrix for 150 taxa to estimate the rate of morphological evolution in the transition from theropod dinosaurs to modern birds. These data were collected over the course of 20 years as part of the Therapod Working Group [@Norell2001; @Clark2002; @Hwang2002; @Hwang2004; @Makovicky2005; @Norell2006; @Turner2007; @Turner2012; @dryad_84t75]. @OLeary2013 combined the work of MorphoBank contributors [@OLeary2011] with literature review to generate 4,541 characters for 86 species. @Rabosky2013 examined 7,822 species of ray-finned fish but used a single quantitative measure (body size) to evaluate correlated speciation and morphological evolution.

Newer approaches, typically involving statistical techniques such as machine vision or natural language processing, still require significant researcher expertise in either gathering large training datasets, or to develop new and more sophisticated machine learning models [@Brill2003; @Halevy2009].

Our approach hits a "sweet spot" on the three axes of expertise, effort, and computation. We use researcher expertise to identify a comparative hypothesis, and design a data collection protocol to specifically test this hypothesis. Users on Amazon Mechanical Turk supply a large source of worker effort that collects data per our protocol. Finally, computational statistical techniques validate the accuracy of our data and identify outliers and other errors in data collection. Researchers do not have to spend time digitizing collections, workers need not generate biological hypotheses, and biologists will not have to solve open questions in the fields of machine vision and natural language processing in order to answer questions in comparative biology. The task of phenomic-scale data collection is split up and efficiently allocated according to the strengths of each role, without overly relying on any one axis to carry out the entire task.

We stress that our work does not attempt to, and will likely never displace traditional morphological research and data collection. The role of these traditional techinques is critical for generating preliminary hypotheses. Our method instead performs high-throughput collection of phenotypic data to match the current speeds of molecular data collection. Collecting this data is not useful without having in hand specific hypothesis to test. Our work fills the niche of gathering phenotypic data across large radiations, which is still a challenging open research question. Even seemingly obvious phenotypes, such as the woodiness of plant species, are incomplete and sampled in a biased manner, potentially misleading inference on a global scale [@Fitzjohn2014]. We suggest that our method can play a crucial role in increasing the amount of data that can be brought to bear against comparative hypotheses.

# Acknowledgements

We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frederich, Vanson Liu, Rosalia Agular, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the hundreds of Amazon Mechanical Turk workers for contributing their time and efforts. We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for useful discussions about fish morphology, and Katie Staab, Tonya Kane, and their respective undergraduate students for serving as guinea pigs during initial evaluation for this study. This work was supported by an Encyclopedia of Life David M. Rubenstein Fellowship (EOL-33066-13) and partially supported by a Stephen and Ruth Wainwright Fellowship to JC. (XXX: Mike's grants?). Travel support to present this research was also provided by the Society for Study of Evolution to JC.

## Author contributions

Conceived and designed the experiments: JC MEA. Performed the experiments: JC. Analyzed the data: JC. Contributed reagents/materials/analysis tools: JC MEA. Wrote the paper: JC MEA.

## Supplementary material

```{r turker_morph_procrustes_other_pc34, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
real_plot_tangent_space(coords, "PC3", "PC4")
```

```{r turker_morph_procrustes_other_pc56, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC5 and PC6 for turker / morphologist Procrustes space"}
real_plot_tangent_space(coords, "PC5", "PC6")
```

- Timing figures for turkers
- Detailed comparison between turkers and morphologists

# References

```{r knitcitations_write, echo=FALSE, message=FALSE}
write.bibtex(file="knitcitations.bib")
```

### 2015-04-15 manuscript/manuscript.Rmd
Manuscript Apr 15

@@
   email: jonathan.chang\@ucla.edu
-  footnote: Author for correspondence
+  footnote: Author for correspondence jonathan.chang\@ucla.edu
   name: Jonathan Chang
@@
     template: inc/mstemplate.tex
-    toc: yes
-    toc_depth: 1
   html_document:
@@
 linestretch: 1.05
-csl: inc/evolution.csl
+csl: inc/methods-in-ecology-and-evolution.csl
+running_head: Fast crowdsourced phenotypic data collection
 address:
@@
   code: ucla
+word_count: 5000?
 ---
@@
 library(MASS)
-library(data.table)
+library(readr)
 library(dplyr)
@@
 library(ipred)
+library(vegan)
+library(lubridate)
 library(directlabels)
@@
 cleanbib()
+set.seed(111) # reproducible and stable sampling
+source("functions.R")
 ```
+
+<!--
+Title page: Provide a concise and informative title (do not include the authorities for any taxonomic names);
+            a short running title (45 characters maximum);
+            the word count (including references, tables and figure legends);
+            a list of all authors' names and addresses;
+            and the full contact details of the corresponding author.
+-->
 
@@
 
-1.
-Advances in genomics and informatics have enabled the production of large phylogenetic trees.
-However, the ability to collect large phenotypic datasets has not kept pace.
-Here, we present a method to quickly gather geometric morphometric data using crowdsourced image-based landmarking.
+<!--
+Abstract (maximum 350 words):
+Point 1: set the context and purpose for the work;
+Point 2: indicate the approach and methods used;
+Point 3: outline the main results;
+Point 4: identify the conclusions, the wider implications and the relevance to management or policy. 
+-->
 
-2. We show that crowdsourced workers perform equally well as experienced morphologists on the same tasks. We also demonstrate the speed of our method on 7 families of Actinopterygii fishes.
+1. Advances in genomics and informatics have enabled the production of large phylogenetic trees. However, the ability to collect large phenotypic datasets has not kept pace.
 
-3. Our methods are publicly available as open source software.
+2. Here, we present a method to quickly and accurately gather morphometric data using crowdsourced image-based landmarking.
+
+3.
+We find that crowdsourced workers perform similarly to experienced morphologists on the same digitization tasks.
+We also demonstrate the speed and accuracy of our method on 7 families of ray-finned fishes (Actinopterygii), and find that rates of body shape evolution are correlated with diversification rates.
+
+4.
+Crowdsourcing will enable the collection of morphological data across vast radiations of organisms, and will facilitate richer inference on the macroevolutionary processes that shape phenotypic diversity across the tree of life.
 
@@
 
-The integration of phenotypic data, such as anatomy, behavior, physiology, and other traits, with phylogenies has been a powerful tool for investigating the patterns of biological evolution.
+<!--
+Introduction: State the reason for the work, the context and the hypotheses being tested.
+-->
+
+XXX: Use r package captioner for captions
+
+Integrating phenotypic data, such as anatomy, behavior, physiology, and other traits, with phylogenies is powerful workflow for investigating the patterns of biological evolution.
 Recent advances in next-generation sequencing [@Shendure2008; @Meyer2008] and sequence capture technologies [@Faircloth2012b; @Lemmon2012] have made phylogenetic inference of large radiations of organisms possible [@McCormack2012; @McCormack2013; @Faircloth2013; @Faircloth2014].
-These large phylogenies have been successfully used in conjunction with existing comprehensive phenotypic datasets to answer key questions about the mode and tempo of diversifiaction and morphological evolution [@Rabosky2013].
+These large phylogenies have been successfully used in conjunction with existing comprehensive phenotypic datasets to answer key questions about the mode and tempo of diversification and morphological evolution [@Rabosky2013].
+However, similar breakthroughs for generating new phenotypic datasets have been comparatively uncommon, likely due to the high expense and effort required [reviewed in @Burleigh2013].
 
-However, similar breakthroughs for generating new phenotypic datasets have been comparatively uncommon, likely due to the high expense and effort required [reviewed in @Burleigh2013].
 Creating these large phenotypic datasets has generally required an extended dedicated effort of measuring and describing morphological or behavioral traits that are then coded into a comprehensive data matrix.
@@
 Finally, methods such as high-throughput infrared imaging, mass spectrometry, and chromatography have been successfully used in plant physiology [@Furbank2011]
- and microbiology [@Skelly2013a], but these methods may not be applicable for zoological researchers.
-These bottlenecks in the collection of phenotypic data has limited the taxonomic breadth and depth of research in comparative biology.
+ and microbiology [@Skelly2013], but these methods may not be applicable for zoological researchers.
+These approaches all share a similar goal of collecting large comparative datasets, but also require large investments in researcher effort.
+This bottleneck has limited the taxonomic breadth and depth of research in comparative biology.
 
@@
 We developed a novel web browser-based image landmarking application, and use Amazon Mechanical Turk (<https://www.mturk.com>) to distribute digitization tasks to remote workers (hereafter *turkers*) over the Internet, who are paid for their contributions.
-We also evaluate the accuracy and precision of turkers by assigning identical image sets and digitization protocols to users who are experienced with fish morphology (hereafter *experts*), and compare the inter- and intra-observer differences between turkers and experts.
-Finally we demonstrate the utility of this new method by constructing a phylogenetic analysis pipeline, and use this pipeline to download fish photographs from the web, collect Mechanical Turk shape results, and analyze the rate of body shape evolution using BAMM [@Rabosky2014a].
+We evaluate the accuracy and precision of turkers by assigning identical image sets and digitization protocols to users who are experienced with fish morphology (hereafter *experts*), and compare the inter- and intra-observer differences between turkers and experts.
+As an example of how this new method may be used, we construct a phylogenetic analysis pipeline to download photographs and phylogenies of 7 actinopterygiian families from the web, collect Mechanical Turk shape results, and analyze the rate of diversification and body shape evolution using BAMM [@Rabosky2014a].
 
-# Methods
+# Materials and methods
+
+<!--
+Materials and methods: Include sufficient details for the work to be repeated.
+-->
 
@@
 
-We developed a cross-browser web application that runs completely on the client.
+We developed an application that runs completely on the user's local web browser.
 This simplifies the infrastructure challenge of needing to serve many crowdsdourced workers simultaneously.
-The web app is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task.
+The web application is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task.
 Below is an example JSON file that demonstrates the utility of our web app:
@@
 Each digitization task to be completed is given a short abbreviation to aid task identification ("C" for center, "D" for diameter, and "O" for outline), and the type of task, "point", "line" or "curve", for homologous landmarks, linear measurements, and sliding semilandmarks can be specified.
-There is also an optional short help snippet that is automatically displayed to inform users what to do.
-These inline help texts complement a larger and more detailed protocol document that workers are required to read before beginning work (Supplementary Material).
+There is also an optional short help snippet displayed inline, which serve as a brief reminder for each landmark and complements a larger and more detailed protocol document that workers are required to read before beginning work (Supplementary Material).
 
-## Server-controlled experiment
-
-Although digitizing and landmarking a single image [microtasks *sensu* @Good2013] is most effective for high-throughput work on MTurk, it is not suited towards conducting controlled experiments.
-To solve this issue we created a server-side front end that mimiced an official Amazon Mechanical Turk endpoint, but would also automatically distribute tasks in the order required by the experiment.
-The MTurk facsimile checkpoints each completed image and distributes another from a Requester-defined set of images.
-A configuration file allows the experimenter to vary the number of replicates required for the given experiment.
-There is also functionality for non-MTurk workers to contribute data using the same mechanism.
+Although digitizing and landmarking a single image [microtasks *sensu* @Good2013] is most effective for high-throughput work on MTurk, it is unsuitable for conducting controlled experiments.
+To solve this issue we created a server-side application backend that automatically distribute tasks according to a configurable set of images and experimental protocol.
+This application also mimics an official Amazon Mechanical Turk endpoint, to decouple the experimental backend from the image landmarking frontend.
+External non-MTurk workers can also participate in the same experiment, ensuring effective comparisons across these groups.
 The software is open source and uses PHP and SQLite, and can be installed on nearly all web hosting services (<https://github.com/jonchang/fake-mechanical-turk>).
@@
 Collecting landmark-based geometric morphometric data at scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007].
-To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk were significantly better or worse than traditionally-collected data, we asked both turkers and experts to landmark a set of 5 fish images, 5 times each.
+To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked both turkers and experts to landmark a set of 5 fish images, 5 times each.
 All participants used the same protocol and same software to digitize the same set of fishes.
 
-The landmarks were chosen according to previously-published literature concerning fish shape [@Chakrabarty2005; @Frederich2008], and included a range of landmarks that would be relatively easy to digitize (e.g., the anterior margin of the eye) and ones that would be challenging (the opercular series).
-A discussion of the performance of turkers versus morphlogists on a per-landmark basis is included in the supplemental material.
+The landmarks were chosen according to previously-published literature concerning fish shape [@Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Thacker2014], and included landmarks both easy to digitize, such as the anterior margin of the eye, and challenging to digitize, such as the opercular series.
+We used 5 replicate measurements per worker per image to quantify measurement error within and among observers [@Palmer1986] using *geomorph* `r citep(citation("geomorph"))`.
 
-We used multiple measurements per worker per image to quantify measurement error within and among observers [@Palmer1986] using *geomorph* `r citep(citation("geomorph"))`.
+We used the median shape of the combined experts' data as a baseline for all comparisons, as the "true" shape.
+To assess within-individual variation, we compared each worker's shape replicate to their own median shape.
+To assess within-group and between-group variation, we also compare those individual worker median shapes to the median shape of all other turkers and all other experts.
+We also measured the median absolute deviation (MAD) as a dispersion statistic for within-worker, within-group, and between-group comparisons.
+
+To test whether turkers and experts were statistically distinguishable,
+
+The curse of dimensionality is a common problem in machine learning, where $p$, the number of predictors, greatly exceeds $n$, the number of observations.
+Geometric morphometric datasets are typically extremely highly dimensional, with $2 * k - 1$ observations per 2D landmark.
+To reduce the dimensionality of our dataset, we orthogonalize our features used a generalized Procrustes superimposition as implement in the R package geomorph.
+We then use logistic regression on the PC axes, starting with PC1 and adding each subsequent axis in a stepwise fashion.
+We validate each model using $K$-fold cross-validation, which splits the data into $K$ equally sized groups, using $K-1$ groups to train the model and the last group to validate the model.
+This process is repeated for each subset of the groups.
+We set $K=10$ to validate the model, for its balanced statistical properties [@Kohavi1995; @Hastie2009].
+Using 10-fold CV, we calculated the confusion matrix for each model in turn, and plot the sensitivity vs specificify (also known as the receiver operator characteristic curve).
+
+To reduce the effect of the curse of dimensionality ($p >> n$, where the number of predictors  exceeds the number of observations), we first
+
+<!-- (XXX: Talk about setting the consensus morphologist shape as the "true" shape in terms of determining accuracy / precision of individual workers and the consensus turker shape) -->
+
 To determine whether turkers and experts were statistically distinguishable, we used the machine learning technique linear discriminant analysis [LDA, @Ripley1996] as implemented in *MASS* `r citep(citation("MASS"))`.
@@
 Conversely, withholding too much data for validation results in underfitting and the inability of the model to generalize its predictions to real data.
-
-To assess the ability of LDA to distinguish between these two groups, we used $K$-fold cross-validation (CV), which splits our data into $K$ equally-sized groups, using $K-1$ for training and $1$ for validation, and testing the model for all possible permutations of these groups.
-The special case of $K=N$ groups is also known as leave-one-out CV.
-We used a value of $K=10$ to validate the LDA model.
-The statistical literature typically recommends $K=5$ or $K=10$ for its balanced approach to the bias-variance tradeoff [@Kohavi1995; @Hastie2009].
+To guard against these potentially misleading inferences, we used $K$-fold cross-validation (CV), which splits our data into $K$ equally-sized groups, using $K-1$ for training and $1$ for validation, and tests the model for all possible permutations of these groups.
+We used a value of $K=10$ to validate the LDA model, for its balanced approach to the bias-variance tradeoff [@Kohavi1995; @Hastie2009].
+Using 10-fold CV, we assessed the LDA model's misprediction rate.
+Acceptable misprediction rates vary depending upon the application, here we judge 20% misprediction rate to sufficently distinguish turkers and experts, while a 50% or higher rate is no better than a random guess.
+We also used quadratic discriminant analysis (QDA), to test the robustness of our results with a different model.
 All analyses were performed in R `r citep(citation())`.
-As an alternate model we also used quadratic discriminant analysis (QDA).
+(XXX: How to interpret the result of this method.
+Use R package caret for easier comparisons among different machine learning techniques.
+Try out support vector machines.)
 
@@
 We calculated the 95% confidence ellipse to estimate the uncertainty around these shape estimates.
+(XXX: How to interpret the result of this method)
 
-## Phylogenetic analysis on 7 families
+## Example: a phenomic pipeline for comparative phylogenetic analysis
 
-To demonstrate the utility and how quickly we can obtain comparative data using this method, we downloaded previously-published phylogenies for 7 fish families (Acanthurdiae, Apogonidae, Balistidae, Chaetodontidae, Labridae, Pomacentridae, and Tetraodontidae) from TreeBASE and Dryad (XXX: CITE TREES) and match them to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using the Reol interface to the EOL API `r citep(citation("Reol"))`.
-If there were multiple images for a given species we manually selected the best one.
+A common strategy in fish comparative studies is to examine evolutionary dynamics within a single family (XXX: cite more papers that do this) [@Frederich2013; @Thacker2014], potentially due to the extensive amount of time necessary to collect data.
+
+To demonstrate the utility of obtaining comparative data using this method, we downloaded previously-published phylogenies for 7 fish families (Acanthurdiae, Apogonidae, Balistidae, Chaetodontidae, Labridae, Pomacentridae, and Tetraodontidae) from TreeBASE and Dryad (XXX: CITE TREES -- need to get info from Mike) and matched them to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using the EOL API [@Parr2014].
+If there were multiple images for a single species, we manually selected the best one.
 Crowdsourced workers collected body shape data following a standard protocol (Supplementary Material).
-We then determined the major components of shape variation using a principal components analysis (PCA) as implemented in the R package *geomorph* `r citep(citation("geomorph"))`.
+We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] as implemented in the R package *geomorph* `r citep(citation("geomorph"))`.
 
-We used Bayesian Analysis of Macroevolutionary Mixtures [*BAMM*, @Rabosky2014a] to estimate the diversification rate as well as the rate of morphological evolution, using the first six principal components for each of the 7 families.
+To test whether our method can improve on the case where the data collection method is geometric morphometrics, we use the average time it took an expert to measure a single fish image and predict the time it would take for a single individual expert to measure our 7 fish families with 5 replicates, and compare it to the time it actually took turkers to collect these measurements at the same replication level.
+If the turkers in aggregate annotated images more quickly than a single expert would have, this suggests that the parallelization afforded by crowdsourcing is effective at reducing the time required for data collection.
+
+We used Bayesian Analysis of Macroevolutionary Mixtures [*BAMM*, @Rabosky2014a] to estimate the diversification rate as well as the rate of morphological evolution, retaining the PC axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998], for each of the 7 families.
 BAMM estimates the location of rate shifts in either diversification or character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method, where the number of rate regimes on a phylogeny is allowed to vary.
-We assessed convergence and mixing using CODA `r citep(citation("coda"))`, BAMMtools `r citep(citation("BAMMtools"))` and Tracer [@rambaut2007tracer].
-We then used BAMMtools to identify groups in each family that experienced increased rates of morphological evolution.
+We assessed convergence and mixing using Tracer [@rambaut2007tracer].
+We then used BAMMtools `r citep(citation("BAMMtools"))` to detect whether body shape affects change in the speciation or extinction rate.
+State-dependent diversification has often been invoked  to explain the high species richness enjoyed by certain lineages [@Mitter1988; @Barraclough1998]; this pipeline demonstrates a largely-automated way to detect these.
+`r citep(bibentry(bibtype="Article", title="A robust semi-parametric test for detecting trait-dependent diversification", author=c("Daniel L.
+Rabosky", "Huateng Huang"), year="in.review", journal="Systematic Biology"))`
 
@@
 
+<!--
+State the results, drawing attention to important details in tables and figures
+-->
+
 ## Reliability analysis
 
-```{r cv_table, echo=F, results="asis", anchor="table"}
-ip.lda <- function(object, newdata) predict(object, newdata=newdata)$class
+```{r sum_table}
+anon_csv <- read_csv("anon.csv")
+no_st <- anon_csv %>% filter(role != "student") %>% mutate(role=factor(role), mark=factor(mark), Family = str_replace(fn, "(.{3}).*", "\\1")) %>% select(-fn)
 
-no_st <- fread("anon.csv") %>% tbl_df() %>% filter(role != "student") %>% mutate(role=factor(role), mark=factor(mark))
+pcaed <- no_st %>% group_by(Family, mark) %>% do(as_df(prcomp(cbind(.$x, .$y), scale=F, center=T)$x, role=as.character(.$role), ident=as.character(.$sha1mac))) %>% mutate(PC1 = as.numeric(PC1), PC2 = as.numeric(PC2))
+
+meds <- pcaed %>% group_by(Family, role, mark) %>% summarise(med=median(PC1)) %>% ungroup() %>% spread(role, med) %>% transmute(Family, mark, value = log(abs(turker - morphologist)), stat="median")
+
+mads <- pcaed %>% group_by(Family, role, mark) %>% summarise(mad=mad(PC1)) %>% ungroup() %>% spread(role, mad) %>% transmute(Family, mark, value = log(turker / morphologist), stat="mad")
+
+all_stats <- bind_rows(meds, mads) %>% spread(stat, value)
+```
+
+```{r geomorph_lm}
+fmted <- format_for_gpagen(no_st)
+allsplit <- str_split_fixed(dimnames(fmted)$id, "_", 3)
+family_vec <- allsplit[, 1]
+role_vec <- allsplit[, 2]
+ind_vec <- allsplit[, 3]
+gpaed <- gpagen(fmted, ShowPlot=F)
+
+role_by_ind <- procD.lm(gpaed$coords ~ family_vec * role_vec * ind_vec, iter=999, RRPP=T)
+ind_only <- procD.lm(gpaed$coords ~ family_vec:ind_vec, iter=999, RRPP=T)
+
+advanced.procD.lm(gpaed$coords ~ ind_vec,  ~ role_vec * ind_vec, ~ family_vec, iter=499)
+```
+
+<!--
+```{r median_tbl, echo=F, results="asis", anchor="table", fig.cap="Median table"}
+med_wide <- meds %>% spread(key=Family, value=value)
+med_wide %>% knitr::kable(digits = 2, caption = "Median table")
+```
+
+```{r mad_tbl, echo=F, results="asis", anchor="table", fig.cap="MAD table"}
+mad_wide <- mads %>% spread(key=Family, value=value)
+mad_wide %>% knitr::kable(digits = 2, caption="MAD table")
+```
+-->
+
+```{r turker_v_expert} 
+ggplot(all_stats, aes(median, mad, label=mark)) + geom_text() + labs(x="Accuracy", y="Precision")
+```
+
+We compare the difference in accuracy and precision between turkers and experts in Figure `r figr("turker_v_expert")`.
+
+The most accurate and precise points are those that are related to the position of the eye (landmarks E1 and E2).
+The least accurate are those in the opercular series (O1-O5), particularly the ones related to the preopercle (O1-O3) as well as in species where the preopercle is difficult to visualize from external morphology alone (e.g., Tetraodontidae).
+We also plot the median difference and the
+
+```{r cv_table, echo=F, results="asis", anchor="table", message=FALSE}
+
 subsamp <- group_by(no_st, fn, role) %>% sample_n(size=400) %>% ungroup()
@@
 
+CV <- subsamp %>% group_by(fn) %>% do(estimator = errorest(role ~ mark:x + mark:y, data=., model=qda, estimator="cv", est.para = control.errorest(k=10), predict=ip.lda))
+
+CV_sum2 <- CV %>% summarise(fn=fn, qda_error=estimator[["error"]]) %>% separate("fn", "fn", sep="_", extra="drop")
+
+CV_sum <- left_join(CV_sum, CV_sum2)
+
 # format for display
-CV_sum %>% transmute(`Family`=fn, `Misprediction rate`=error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis with 10-fold cross validation for each fish image.")
+CV_sum %>% transmute(`Family`=fn, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.")
 ```
 
-The linear discriminant analysis (LDA) with 10-fold cross validation (Table `r figr("cv_table")`) was unable to reliably distinguish between morphologists and crowdsourced workers.
-The quadratic estimator (QDA) performed similarly poorly (results not shown).
+Both linear and quadratic discriminant analysis with 10-fold cross validation (Table `r figr("cv_table")`) was unable to reliably distinguish between morphologists and crowdsourced workers using our acceptable misprediction threshold of 20%.
 
@@
 # Load data from the anonymized results file, remove students, drop bad marks
-df <- fread("anon.csv") %>% tbl_df() %>%
+df <- read_csv("anon.csv") %>%
 	filter(role != "student") %>% 
@@
 df <- df %>% group_by(fn, mark, role, sha1mac) %>% summarise(x=median(x), y=median(y))
-
-format_for_gpagen <- function(df) {
-	splut <- gather(df, variable, val, x:y) %>% unite(col=id, fn, role, sha1mac)
-	xtabs(val ~ mark + variable + id, data=splut)
-}
 
@@
 
+```{r, echo=FALSE}
+# use broken stick method to get the principal components stopping point
+
+pc_res <- prcomp(two.d.array(coords))
+eigen_res <- pc_res$sdev^2
+bstick_res <- bstick(pc_res)
+
+stopping_pc <- which(eigen_res < bstick_res)[1] %>% as.numeric
+```
+
+We analyzed PC 1 through PC `r stopping_pc`, based on the broken-stick method of determining a PCA stopping point.
 Projecting consensus shapes into Procrustes space (Figure `r figr("turk_morph_procrustes_space")`) shows that within-group variation exceeds that of among group variation.
@@
 
-## Phylogenetic analysis on 7 families
+## Phenomic pipeline for comparative phylogenetic analysis
 
-The full data analysis pipeline end-to-end took 2.2 hours of wall-clock time.
-The BAMMtools analysis identified areas of each phylogeny that were experiencing faster rates of body shape evolution.
-XXX: Talk about these in more detail.
+XXX: Include BAMM speciation / morphological evolution analysis.
 
 ```{r bamm_plots, echo=F, message=F, warning=F, anchor="figure", cache=T}
-results <- fread("fishturk.results") %>% tbl_df() %>% filter(Answer.marks != "") %>% mutate(tip=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 3], family=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 2], Answer.marks=str_replace_all(Answer.marks, '""', '"'))
+results <- read_tsv("fishturk.results") %>% filter(Answer.marks != "") %>% mutate(tip=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 3], family=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 2], Answer.marks=str_replace_all(Answer.marks, '""', '"'))
 
@@
 
-res <- plotTangentSpace(coords, groups=groups, axis1=1, axis2=2, verbose=T, warpgrids=F)$pc.scores
+res <- plotTangentSpace(full_coords, groups=groups, axis1=1, axis2=2, verbose=T, warpgrids=F)$pc.scores
 
@@
 res$id <- rownames(res)
-res2<-res %>% separate(id, into=c("family", "name"), sep="\\|", extra="drop") %>% tbl_df()
+res2<- res %>% separate(id, into=c("family", "name"), sep="\\|", extra="drop") %>% tbl_df()
 
@@
 
+```{r timing_plot, echo=FALSE, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
+Landmarks were first received 6 minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
+# median time (in minutes) it takes an expert to measure these fishes
+timings <- (no_st %>% group_by(role) %>% summarise(duration=median(duration))) %>% mutate(total_time=duration * nrow(res2))
+
+morpho_per_image <- timings[1, 2] %>% as.numeric() %>% dminutes() %>% round(1)
+
+# data collection time plot
+results %>% transmute(hitid, start=parse_date_time(creationtime, "bdHMSY"), end=parse_date_time(assignmentsubmittime, "bdHMSY"), diff = end - start) %>% group_by(hitid) %>% summarise(time=min(diff)) %>% arrange(time) %>% mutate(seq = 1:n(), frac = seq / n()) %>% ggplot(aes(time / eminutes(), frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
+
+total_turker_time <- results %>% transmute(hitid, start=parse_date_time(creationtime, "bdHMSY"), end=parse_date_time(assignmentsubmittime, "bdHMSY"), diff = end - start) %>% summarise(max_time = max(diff)) %>% as.numeric %>% dminutes %>% round(1)
+```
+
+Using a median-expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(res2))` to landmark all `r nrow(res2)` images.
+At 5x replication, this would take `r as.character(morpho_per_image * nrow(res2) * 5)`.
+By comparison, turkers took a total of `r as.character(total_turker_time)` to complete all images at 5x replication.
+
+The BAMMtools analysis identified areas of each phylogeny that were experiencing faster rates of body shape evolution. XXX: Talk about these in more detail.
+
 # Discussion
+
+<!--
+Point out the importance of the results and place them in the context of previous studies and in relation to the application of the work (expanding on the Synthesis and applications section of the Summary).
+Where appropriate, set out recommendations for management or policy.
+-->
+
+XXX: Discuss what people currently do to estimate error in geometric morphometrics. Make the point that this method is really good at getting at an estimate of error.
+
+We found that for certain landmarks there was significant between and within group disagreement.
+Based on median average deviation, points belonging to the opercular series and those that concerned the distal margin of the dorsal and anal fins were particularly challenging for turkers to digitize, compared to the experts.
+Typically, geometric morphometric studies only use a single digitizer in efforts to reduce interdigitizer error.
+We argue that this common technique is flawed, as it papers over potential differences in how observers measure the same homologous landmark on the same organism.
+Instead, explicitly accounting for interobsever error is critical to determine the efficacy of each individual landmark and the replicability of the study as a whole.
+Examining interobserver error is a critical process that signals which landmarks can be relied on and which can be discarded, as we have done in this analysis.
+Despite the reduction in the amounts of landmark data collected, even a simplified protocol can produce meaningful biological results that are concordant with previous hypotheses in these groups.
 
@@
 We attempted to distinguish between turker-generated landmarks and expert-generated landmarks using linear discriminant analysis.
-Even the image with the lowest misprediction rate, from the family `r filter(CV_sum, min(error) == error)$fn`, still incorrectly predicted the digitizer's identity `r filter(CV_sum, min(error) == error)$error %>% as.numeric() %>% scales::percent()` of the time, suggesting that, given a set of landmarks, it is difficult to reliably predict whether a turker or a morphologist created them.
+Even the image with the lowest misprediction rate, from the family `r filter(CV_sum, min(error) == error)$fn`, still incorrectly predicted the digitizer's identity `r filter(CV_sum, min(error) == error)$error %>% as.numeric() %>% scales::percent()` of the time, suggesting that, given a set of landmarks, it is difficult to predict whether a turker or a morphologist created them.
+The quadratic discriminant analysis did not reliably improve prediction of worker identity over the LDA.
+
+The traditional way of collecting phenotypic data involves enormous researcher effort and significant morphological expertise.
+For example, @Brusatte2014 used a 853 character discrete character matrix for 150 taxa to estimate the rate of morphological evolution in the transition from theropod dinosaurs to modern birds.
+These data were collected over the course of 20 years as part of the Therapod Working Group [@Norell2001; @Clark2002; @Hwang2002; @Hwang2004; @Makovicky2005; @Norell2006; @Turner2007; @Turner2012; @dryad_84t75].
+@OLeary2013 combined the work of MorphoBank contributors [@OLeary2011] with literature review to generate 4,541 characters for 86 species.
+@Rabosky2013 examined 7,822 species of ray-finned fish but used a single quantitative measure (body size) to evaluate correlated speciation and morphological evolution.
+Newer approaches, typically involving statistical techniques such as machine vision or natural language processing, still require significant researcher expertise in either gathering large training datasets, or to develop new and more sophisticated machine learning models [@Brill2003; @Halevy2009].
 
 Our novel pipeline to download images, upload them to Amazon MTurk, and process them using BAMM and BAMMtools showcases the ability to rapidly collect phenotypic data.
-The majority of the 2.2 hours of wall-clock time to collect these data were spent on waiting for worker results; however, a majority of the data had already been collected at the 1-hour mark.
+Most of the time taken to collect these data were spent on waiting for worker results; however, a majority of the data had already been collected at the 1-hour mark.
 An online methodology could conceivably improve on this analysis time, by iteratively refining its results as new data streamed in from Amazon's servers.
@@
 
-The traditional way of collecting phenotypic data involves enormous researcher effort and significant morphological expertise.
-For example, @Brusatte2014 used a 853 character discrete character matrix for 150 taxa to estimate the rate of morphological evolution in the transition from theropod dinosaurs to modern birds.
-These data were collected over the course of 20 years as part of the Therapod Working Group [@Norell2001; @Clark2002; @Hwang2002; @Hwang2004; @Makovicky2005; @Norell2006; @Turner2007; @Turner2012; @dryad_84t75].
-@OLeary2013 combined the work of MorphoBank contributors [@OLeary2011] with literature review to generate 4,541 characters for 86 species.
-@Rabosky2013 examined 7,822 species of ray-finned fish but used a single quantitative measure (body size) to evaluate correlated speciation and morphological evolution.
-
-Newer approaches, typically involving statistical techniques such as machine vision or natural language processing, still require significant researcher expertise in either gathering large training datasets, or to develop new and more sophisticated machine learning models [@Brill2003; @Halevy2009].
-
-Our approach hits a "sweet spot" on the three axes of expertise, effort, and computation.
+Our approach hits a "sweet spot" on the three axes of expertise, effort, and computational complexity.
 We use researcher expertise to identify a comparative hypothesis, and design a data collection protocol to specifically test this hypothesis.
@@
 Our method instead performs high-throughput collection of phenotypic data to match the current speeds of molecular data collection.
-Collecting this data is not useful without having in hand specific hypothesis to test.
+Collecting this data is not useful without having in mind a specific hypothesis to test.
 Our work fills the niche of gathering phenotypic data across large radiations, which is still a challenging open research question.
-Even seemingly obvious phenotypes, such as the woodiness of plant species, are incomplete and sampled in a biased manner, potentially misleading inference on a global scale [@Fitzjohn2014].
+Even seemingly obvious phenotypes, such as the woodiness of plant species, are incomplete and sampled in a biased manner, potentially misleading inference on a global scale [@FitzJohn2014].
 We suggest that our method can play a crucial role in increasing the amount of data that can be brought to bear against comparative hypotheses.
@@
 
-We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frederich, Vanson Liu, Rosalia Agular, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the hundreds of Amazon Mechanical Turk workers for contributing their time and efforts.
-We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for useful discussions about fish morphology, and Katie Staab, Tonya Kane, and their respective undergraduate students for serving as guinea pigs during initial evaluation for this study.
+We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frédérich, Vanson Liu, Rosalia Aguilar, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the `r (no_st %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (results %>% select(workerid) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
+We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for insightful discussions about fish morphology and digitization protocols.
+Katie Staab, Tonya Kane, and their `r anon_csv %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students served as beta testers for the methods in this study.
 This work was supported by an Encyclopedia of Life David M.
-Rubenstein Fellowship (EOL-33066-13) and partially supported by a Stephen and Ruth Wainwright Fellowship to JC.
+Rubenstein Fellowship (EOL-33066-13), a Stephen and Ruth Wainwright Fellowship, and a UCLA Research and Conference Award to JC.
 (XXX: Mike's grants?).
@@
 
+XXX: Best practices for turking
+
 ```{r turker_morph_procrustes_other_pc34, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
@@
 
-- Timing figures for turkers
-- Detailed comparison between turkers and morphologists
-
 # References

### 2015-04-16 manuscript/manuscript.Rmd
Move functions into package

@@
 ```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
-library(MASS)
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(stringr)
-library(magrittr)
-library(tidyr)
-library(reshape2)
-library(scales)
-library(grid)
-library(geomorph)
-library(rjson)
-library(ipred)
-library(vegan)
-library(lubridate)
-library(directlabels)
-library(BAMMtools)
-library(knitcitations)
-library(kfigr)
+library("MASS")
+library("readr")
+library("dplyr")
+library("ggplot2")
+library("stringr")
+library("magrittr")
+library("tidyr")
+library("reshape2")
+library("scales")
+library("grid")
+library("geomorph")
+library("rjson")
+library("ipred")
+library("vegan")
+library("lubridate")
+library("directlabels")
+library("BAMMtools")
+library("knitcitations")
+library("kfigr")
+library("fish.reliability")
 options("citation_format"="pandoc")
@@
 set.seed(111) # reproducible and stable sampling
-source("functions.R")
 ```

### 2015-04-16 R/functions.R

@@ -1,0 +1,79 @@
+#' Coerce to a data frame
+#'
+#' @param x A rectangular object, with or without rownames
+#' @param ... Additional columns to include
+#' @return A data frame. Row names, if present in `x`, are included as a `.rownames` column
+#' @export
+as_df <- function(x, ...) {
+  if (!is.null(rownames(x))) {
+    x <- cbind(.rownames=rownames(x), x)
+  }
+  x <- cbind(x, ...)
+  as.data.frame(x, stringsAsFactors=F)
+}
+
+#' Prediction object for `ipred`
+#' @export
+ip.lda <- function(object, newdata) {
+  predict(object, newdata=newdata)$class
+}
+
+#' Formats a data frame for `geomorph::gpagen`
+#'
+#' @param df a data frame, with certain columns
+#' @return A 3D matrix
+#' @importFrom tidyr gather unite
+#' @importFrom magrittr %>%
+#' @export
+format_for_gpagen <- function(df) {
+  splut <- gather(df, variable, val, x:y) %>% unite(col=id, Family, role, sha1mac)
+  xtabs(val ~ mark + variable + id, data=splut)
+}
+
+
+#' Hack the shit out of geomorph's slow procD.lm function
+#' @export
+my.procD.lm <- function (f1, iter = 999, RRPP = FALSE, int.first = FALSE, verbose = FALSE)
+{
+  form.in <- formula(f1)
+  Y <- eval(form.in[[2]], parent.frame())
+  if (length(dim(Y)) == 3)
+    Y <- two.d.array(Y)
+  else Y <- as.matrix(Y)
+  if (nrow(Y) != nrow(model.frame(form.in[-2])))
+    stop("Different numbers of specimens in dependent and independent variables")
+  form.new <- as.formula(paste(c("Y", form.in[[3]]), collapse = "~"))
+  if (int.first == TRUE)
+    ko = TRUE
+  else ko = FALSE
+  Terms <- terms(form.new, keep.order = ko)
+  mod.mf <- model.frame(Terms)
+  if (any(is.na(Y)) == T)
+    stop("Response data matrix (shape) contains missing values. Estimate these first (see 'estimate.missing').")
+  anova.parts.obs <- geomorph:::anova.parts(f1 = form.new, Yalt = "observed",
+                                 keep.order = ko)
+  anova.tab <- anova.parts.obs$table
[…]

### 2015-04-16 R/functions.R
Fix description, imports

@@ -33,6 +33,7 @@ format_for_gpagen <- function(df) {
 
 #' Hack the shit out of geomorph's slow procD.lm function
 #' @export
+#' @import geomorph
 my.procD.lm <- function (f1, iter = 999, RRPP = FALSE, int.first = FALSE, verbose = FALSE)
 {
   form.in <- formula(f1)

### 2015-04-16 manuscript/manuscript.Rmd
Add data to package

--- 2015-04-15 manuscript/manuscript.Rmd
@@
 set.seed(111) # reproducible and stable sampling
-source("functions.R")
+data(fish_landmarks)
 ```
@@
 ```{r sum_table}
-anon_csv <- read_csv("anon.csv")
-no_st <- anon_csv %>% filter(role != "student") %>% mutate(role=factor(role), mark=factor(mark), Family = str_replace(fn, "(.{3}).*", "\\1")) %>% select(-fn)
+no_st <- fish_landmarks %>% filter(role != "student")
 
-pcaed <- no_st %>% group_by(Family, mark) %>% do(as_df(prcomp(cbind(.$x, .$y), scale=F, center=T)$x, role=as.character(.$role), ident=as.character(.$sha1mac))) %>% mutate(PC1 = as.numeric(PC1), PC2 = as.numeric(PC2))
+pcaed <- no_st %>% group_by(family, mark) %>% do(as_df(prcomp(cbind(.$x, .$y), scale=F, center=T)$x, role=as.character(.$role), ident=as.character(.$sha1mac))) %>% mutate(PC1 = as.numeric(PC1), PC2 = as.numeric(PC2))
 
-meds <- pcaed %>% group_by(Family, role, mark) %>% summarise(med=median(PC1)) %>% ungroup() %>% spread(role, med) %>% transmute(Family, mark, value = log(abs(turker - morphologist)), stat="median")
+meds <- pcaed %>% group_by(family, role, mark) %>% summarise(med=median(PC1)) %>% ungroup() %>% spread(role, med) %>% transmute(family, mark, value = log(abs(turker - morphologist)), stat="median")
 
-mads <- pcaed %>% group_by(Family, role, mark) %>% summarise(mad=mad(PC1)) %>% ungroup() %>% spread(role, mad) %>% transmute(Family, mark, value = log(turker / morphologist), stat="mad")
+mads <- pcaed %>% group_by(family, role, mark) %>% summarise(mad=mad(PC1)) %>% ungroup() %>% spread(role, mad) %>% transmute(family, mark, value = log(turker / morphologist), stat="mad")
 
@@
 ```{r median_tbl, echo=F, results="asis", anchor="table", fig.cap="Median table"}
-med_wide <- meds %>% spread(key=Family, value=value)
+med_wide <- meds %>% spread(key=family, value=value)
 med_wide %>% knitr::kable(digits = 2, caption = "Median table")
@@
 ```{r mad_tbl, echo=F, results="asis", anchor="table", fig.cap="MAD table"}
-mad_wide <- mads %>% spread(key=Family, value=value)
+mad_wide <- mads %>% spread(key=family, value=value)
 mad_wide %>% knitr::kable(digits = 2, caption="MAD table")
@@
 
-subsamp <- group_by(no_st, fn, role) %>% sample_n(size=400) %>% ungroup()
+subsamp <- group_by(no_st, family, role) %>% sample_n(size=400) %>% ungroup()
 
-CV <- subsamp %>% group_by(fn) %>% do(estimator = errorest(role ~ mark:x + mark:y, data=., model=lda, estimator="cv", est.para = control.errorest(k=10), predict=ip.lda))
+CV <- subsamp %>% group_by(family) %>% do(estimator = errorest(role ~ mark:x + mark:y, data=., model=lda, estimator="cv", est.para = control.errorest(k=10), predict=ip.lda))
 
-CV_sum <- CV %>% summarise(fn=fn, error=estimator[["error"]]) %>% separate("fn", "fn", sep="_", extra="drop")
+CV_sum <- CV %>% summarise(family=family, error=estimator[["error"]]) %>% separate("family", "family", sep="_", extra="drop")
 
-CV <- subsamp %>% group_by(fn) %>% do(estimator = errorest(role ~ mark:x + mark:y, data=., model=qda, estimator="cv", est.para = control.errorest(k=10), predict=ip.lda))
+CV <- subsamp %>% group_by(family) %>% do(estimator = errorest(role ~ mark:x + mark:y, data=., model=qda, estimator="cv", est.para = control.errorest(k=10), predict=ip.lda))
 
-CV_sum2 <- CV %>% summarise(fn=fn, qda_error=estimator[["error"]]) %>% separate("fn", "fn", sep="_", extra="drop")
+CV_sum2 <- CV %>% summarise(family=family, qda_error=estimator[["error"]]) %>% separate("family", "family", sep="_", extra="drop")
 
@@
 # format for display
-CV_sum %>% transmute(`Family`=fn, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.")
+CV_sum %>% transmute(`Family`=family, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.")
 ```
@@
 # Load data from the anonymized results file, remove students, drop bad marks
-df <- read_csv("anon.csv") %>%
+df <- fish_landmarks %>%
 	filter(role != "student") %>% 
 	separate(col=mark, into=c("cat", "cat_seq"), sep=1, remove=F, convert=T) %>%
-	mutate(round=floor(sequence/5), fn=factor(str_extract(fn, ".{8}"))) %>%
-	filter(!(fn %in% c("Balistid") & mark %in% c("D1","D2")), !mark %in% c("O1","O2","O3","O4","O5", "A3", "D3", "D4"))
+	filter(!(family %in% c("Bal") & mark %in% c("D1","D2")), !mark %in% c("O1","O2","O3","O4","O5", "A3", "D3", "D4"))
 
 # grabs the consensus sequence by person and fish
-df <- df %>% group_by(fn, mark, role, sha1mac) %>% summarise(x=median(x), y=median(y))
+df <- df %>% group_by(family, mark, role, sha1mac) %>% summarise(x=median(x), y=median(y))
 
@@
 	res$id <- rownames(res)
-	res2 <- res %>% separate(id, into=c("fn", "role","sha"), sep="_") %>% tbl_df()
+	res2 <- res %>% separate(id, into=c("family", "role","sha"), sep="_") %>% tbl_df()
 
@@
 		y=lazyeval::interp(~median(axis2, na.rm=T), axis2=as.name(axis2)))
-	consensus <- res2 %>% group_by(fn, role) %>% summarise_(.dots=dots)
+	consensus <- res2 %>% group_by(family, role) %>% summarise_(.dots=dots)
 
-	p <- ggplot(res2, aes_string(axis1, axis2, group="fn")) + geom_point(aes(col=fn, shape=role)) + geom_point(aes(x, y), data=consensus, alpha=0.5, size=5, col="black") + geom_point(aes(x, y, col=fn, shape=role), data=consensus) + stat_ellipse(aes(color=fn), linetype=2, type="norm", level=0.95) + theme_minimal() + theme(legend.position="none")
+	p <- ggplot(res2, aes_string(axis1, axis2, group="family")) + geom_point(aes(col=family, shape=role)) + geom_point(aes(x, y), data=consensus, alpha=0.5, size=5, col="black") + geom_point(aes(x, y, col=family, shape=role), data=consensus) + stat_ellipse(aes(color=family), linetype=2, type="norm", level=0.95) + theme_minimal() + theme(legend.position="none")
 
@@
 We attempted to distinguish between turker-generated landmarks and expert-generated landmarks using linear discriminant analysis.
-Even the image with the lowest misprediction rate, from the family `r filter(CV_sum, min(error) == error)$fn`, still incorrectly predicted the digitizer's identity `r filter(CV_sum, min(error) == error)$error %>% as.numeric() %>% scales::percent()` of the time, suggesting that, given a set of landmarks, it is difficult to predict whether a turker or a morphologist created them.
+Even the image with the lowest misprediction rate, from the family `r filter(CV_sum, min(error) == error)$family`, still incorrectly predicted the digitizer's identity `r filter(CV_sum, min(error) == error)$error %>% as.numeric() %>% scales::percent()` of the time, suggesting that, given a set of landmarks, it is difficult to predict whether a turker or a morphologist created them.
 The quadratic discriminant analysis did not reliably improve prediction of worker identity over the LDA.

### 2015-04-16 R/fish_landmarks.R

@@ -1,0 +1,18 @@
+#' Shape landmarks for 7 fish families
+#'
+#' A dataset containing landmarks for 7 fish families, marked up by turkers
+#' and experienced morphologists alike.
+#'
+#' @format A data frame with 124875 rows and 8 variables:
+#' \describe{
+#'   \item{sha1mac}{hash identifying each digitizer}
+#'   \item{sequence}{an integer describing the order this image was landmarked}
+#'   \item{round}{an integer describing the nth time this image was landmarked}
+#'   \item{role}{a factor describing whether the digitizer was a student, morphologist, or turker}
+#'   \item{mark}{the name of the landmark in question}
+#'   \item{x}{the x-coordinate of this landmark}
+#'   \item{y}{the y-coordinate of this landmark}
+#'   \item{duration}{how long, in seconds, this image took to landmark}
+#'   \item{family}{the family that this specimen belonged to}
+#' }
+"fish_landmarks"

### 2015-04-16 R/functions.R

@@ -26,7 +26,7 @@ ip.lda <- function(object, newdata) {
 #' @importFrom magrittr %>%
 #' @export
 format_for_gpagen <- function(df) {
-  splut <- gather(df, variable, val, x:y) %>% unite(col=id, Family, role, sha1mac)
+  splut <- gather(df, variable, val, x:y) %>% unite(col=id, family, role, sha1mac)
   xtabs(val ~ mark + variable + id, data=splut)
 }
 

### 2015-04-16 manuscript/manuscript.Rmd
Fix manuscript for packaging

@@
 ```{r sum_table}
-no_st <- fish_landmarks %>% filter(role != "student")
+no_st <- fish_landmarks %>% filter(role != "student") %>% mutate(role = factor(role))
 
@@
 
-```{r geomorph_lm}
+```{r geomorph_lm, cache=TRUE}
 fmted <- format_for_gpagen(no_st)
@@
 
-```{r turker_v_expert} 
+```{r turker_v_expert, anchor="figure"} 
 ggplot(all_stats, aes(median, mad, label=mark)) + geom_text() + labs(x="Accuracy", y="Precision")
@@
 We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for insightful discussions about fish morphology and digitization protocols.
-Katie Staab, Tonya Kane, and their `r anon_csv %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students served as beta testers for the methods in this study.
+Katie Staab, Tonya Kane, and their `r fish_landmarks %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students served as beta testers for the methods in this study.
 This work was supported by an Encyclopedia of Life David M.

### 2015-04-16 manuscript/manuscript.Rmd
Fix variables for packaging

@@
 bibliography:
-- ../../Mendeley/library.bib
+- ../../../Mendeley/library.bib
 - knitcitations.bib

### 2015-04-16 manuscript/manuscript.Rmd
Download bibliography from Dropbox

--- 2015-04-16 manuscript/manuscript.Rmd
@@
 bibliography:
-- ../../Mendeley/library.bib
+- library.bib
 - knitcitations.bib
@@
 write.bibtex(file="knitcitations.bib")
+# Also download Mendeley bibliography
+download.file("https://www.dropbox.com/s/ehwm3c945k7rh0n/library.bib?dl=0", destfile="library.bib", method="curl", extra="-L")
 ```

### 2015-04-16 manuscript/manuscript.Rmd
Geomorph LM seems broken

@@
 
+<!--
 ```{r geomorph_lm, cache=TRUE}
@@
 ```
+-->
 

### 2015-04-16 manuscript/manuscript.Rmd
Really disable geomorph LM

--- 2015-04-16 manuscript/manuscript.Rmd
@@
 
-```{r geomorph_lm, cache=TRUE}
+```{r geomorph_lm, eval=FALSE, cache=TRUE}
 fmted <- format_for_gpagen(no_st)
@@
 
-<!--
 ```{r median_tbl, echo=F, results="asis", anchor="table", fig.cap="Median table"}
@@
 ```
--->
 

### 2015-04-16 R/fish_landmarks.R
Add species data

@@ -1,18 +1,0 @@
-#' Shape landmarks for 7 fish families
-#'
-#' A dataset containing landmarks for 7 fish families, marked up by turkers
-#' and experienced morphologists alike.
-#'
-#' @format A data frame with 124875 rows and 8 variables:
-#' \describe{
-#'   \item{sha1mac}{hash identifying each digitizer}
-#'   \item{sequence}{an integer describing the order this image was landmarked}
-#'   \item{round}{an integer describing the nth time this image was landmarked}
-#'   \item{role}{a factor describing whether the digitizer was a student, morphologist, or turker}
-#'   \item{mark}{the name of the landmark in question}
-#'   \item{x}{the x-coordinate of this landmark}
-#'   \item{y}{the y-coordinate of this landmark}
-#'   \item{duration}{how long, in seconds, this image took to landmark}
-#'   \item{family}{the family that this specimen belonged to}
-#' }
-"fish_landmarks"

### 2015-04-16 R/zzzdata.R

@@ -1,0 +1,38 @@
+#' Replicated shape landmarks for 7 fish families
+#'
+#' A dataset containing landmarks for 7 fish families, marked up by turkers
+#' and experienced morphologists alike. Each digitizer analyzed 5 images 5
+#' times each over the course of approximately an hour.
+#'
+#' @format A data frame with 124875 rows and 8 variables:
+#' \describe{
+#'   \item{sha1mac}{hash identifying each digitizer}
+#'   \item{sequence}{an integer describing the order this image was landmarked}
+#'   \item{round}{an integer describing the nth time this image was landmarked}
+#'   \item{role}{a factor describing whether the digitizer was a student, morphologist, or turker}
+#'   \item{mark}{the name of the landmark in question}
+#'   \item{x}{the x-coordinate of this landmark}
+#'   \item{y}{the y-coordinate of this landmark}
+#'   \item{duration}{how long, in seconds, this image took to landmark}
+#'   \item{family}{the family that this specimen belonged to}
+#' }
+"fish_landmarks"
+
+
+
+
+#' Shape landmarks for 7 fish families
+#'
+#' A dataset containing landmarks for 147 species across 7 fish families.
+#'
+#' @format A data frame with 10,360 rows and 6 variables:
+#' \describe{
+#'   \item{worker}{hash identifying each digitizer}
+#'   \item{mark}{the name of the landmark in question}
+#'   \item{x}{the x-coordinate of this landmark}
+#'   \item{y}{the y-coordinate of this landmark}
+#'   \item{family}{the family that this specimen belonged to}
+#'   \item{tip}{the species digitized}
+#'   \item{time_taken}{the amount of time taken by this worker to digitize this image}
+#' }
+"fish_families"

### 2015-04-16 manuscript/manuscript.Rmd
Update MS for data changes

@@
 data(fish_landmarks)
+data(fish_families)
 ```
@@
 ```{r bamm_plots, echo=F, message=F, warning=F, anchor="figure", cache=T}
-results <- read_tsv("fishturk.results") %>% filter(Answer.marks != "") %>% mutate(tip=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 3], family=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 2], Answer.marks=str_replace_all(Answer.marks, '""', '"'))
-
-get_marks <- function(blob, ...) {
-  fromJSON(blob) %>% as.data.frame %>% t %>% as.data.frame %>% set_colnames(c("x", "y")) %>% mutate(mark=rownames(.)) %>% cbind(...)
-}
-
-allmarks <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=.$workerid, family=.$family, tip=.$tip))
-
-df <- allmarks %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))
+df <- fish_families %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))
 
@@
 # data collection time plot
-results %>% transmute(hitid, start=parse_date_time(creationtime, "bdHMSY"), end=parse_date_time(assignmentsubmittime, "bdHMSY"), diff = end - start) %>% group_by(hitid) %>% summarise(time=min(diff)) %>% arrange(time) %>% mutate(seq = 1:n(), frac = seq / n()) %>% ggplot(aes(time / eminutes(), frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
+fish_families %>% group_by(worker, tip) %>% summarise(time=min(time_taken)) %>% arrange(time) %>% mutate(seq = 1:n(), frac = seq / n()) %>% ggplot(aes(dseconds(time) / eminutes(), frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
 
-total_turker_time <- results %>% transmute(hitid, start=parse_date_time(creationtime, "bdHMSY"), end=parse_date_time(assignmentsubmittime, "bdHMSY"), diff = end - start) %>% summarise(max_time = max(diff)) %>% as.numeric %>% dminutes %>% round(1)
+total_turker_time <- max(fish_families$time_taken) %>% as.numeric %>% dminutes %>% round(1)
 ```
@@
 
-We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frédérich, Vanson Liu, Rosalia Aguilar, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the `r (no_st %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (results %>% select(workerid) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
+We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frédérich, Vanson Liu, Rosalia Aguilar, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the `r (fish_landmarks %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (fish_families %>% select(worker) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
 We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for insightful discussions about fish morphology and digitization protocols.

### 2015-04-22 manuscript/manuscript.Rmd
Use RCurl and R 3.2.0

@@
 # Also download Mendeley bibliography
-download.file("https://www.dropbox.com/s/ehwm3c945k7rh0n/library.bib?dl=0", destfile="library.bib", method="curl", extra="-L")
+mendbib <- RCurl::getURL("https://www.dropbox.com/s/ehwm3c945k7rh0n/library.bib?dl=1", .opts=RCurl::curlOptions(followlocation=TRUE))
+cat(mendbib, file="library.bib")
 ```

### 2015-04-22 manuscript/manuscript.Rmd
Move plot tangent space function into package

@@
 coords <- gpagen(format_for_gpagen(df), ShowPlot=F)$coords
+```
 
-real_plot_tangent_space <- function(coords, axis1="PC1", axis2="PC2") {
-  x <- two.d.array(coords)
-  pc.res <- prcomp(x)
-  pcdata <- pc.res$x
+Text text
 
-	res <- as.data.frame(pcdata)
-	res$id <- rownames(res)
-	res2 <- res %>% separate(id, into=c("family", "role","sha"), sep="_") %>% tbl_df()
+```{r turk_morph_procrustes_space, echo=FALSE, fig.cap="Testing my figure", anchor="figure"}
+plot_tangent_space(coords, "PC1", "PC2")
+```
 
-	# consensus in Procrustes space for downstream highlighting
-	dots <- list(
-		x=lazyeval::interp(~median(axis1, na.rm=T), axis1=as.name(axis1)),
-		y=lazyeval::interp(~median(axis2, na.rm=T), axis2=as.name(axis2)))
-	consensus <- res2 %>% group_by(family, role) %>% summarise_(.dots=dots)
-
-	p <- ggplot(res2, aes_string(axis1, axis2, group="family")) + geom_point(aes(col=family, shape=role)) + geom_point(aes(x, y), data=consensus, alpha=0.5, size=5, col="black") + geom_point(aes(x, y, col=family, shape=role), data=consensus) + stat_ellipse(aes(color=family), linetype=2, type="norm", level=0.95) + theme_minimal() + theme(legend.position="none")
-
-	direct.label(p, method="smart.grid")
-}
-
-real_plot_tangent_space(coords, "PC1", "PC2")
-```
+Post text
 
@@
 
-```{r turker_morph_procrustes_other_pc34, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
-real_plot_tangent_space(coords, "PC3", "PC4")
+```{r turker_morph_procrustes_other_pc23, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC2 and PC3 for turker / morphologist Procrustes space"}
+plot_tangent_space(coords, "PC2", "PC3")
 ```
 
-```{r turker_morph_procrustes_other_pc56, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC5 and PC6 for turker / morphologist Procrustes space"}
-real_plot_tangent_space(coords, "PC5", "PC6")
+```{r turker_morph_procrustes_other_pc34, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
+plot_tangent_space(coords, "PC3", "PC4")
+```
+
+```{r turker_morph_procrustes_other_pc45, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC4 and PC5 for turker / morphologist Procrustes space"}
+plot_tangent_space(coords, "PC4", "PC5")
 ```

### 2015-04-22 R/plotting.R

@@ -1,0 +1,30 @@
+#' Plot specimens in tangent space
+#'
+#' This function plots a set of Procrustes-aligned specimens in tangent space along their principal axes. This is a custom replacement for `geomorph::plotTangentSpace` tailored for the fish reliability project.
+#'
+#' @param coords An array (p x k x n) containing landmark coordinates for a set of aligned specimens, possibly from `geomorph::gpagen`. Array labels must be in the format `group1_group2_individual`.
+#' @param axis1 The name of the first axis to plot
+#' @param axis2 The name of the second axis to plot
+#' @return A `ggplot` object
+#' @export
+#' @import dplyr
+#' @import ggplot2
+plot_tangent_space <- function(coords, axis1="PC1", axis2="PC2") {
+  x <- geomorph::two.d.array(coords)
+  pc.res <- stats::prcomp(x)
+  pcdata <- pc.res$x
+
+  res <- as.data.frame(pcdata)
+  res$id <- rownames(res)
+  res2 <- res %>% tidyr::separate(id, into=c("family", "role","sha"), sep="_") %>% left_join(full_family, by="family")
+
+  # consensus in Procrustes space for downstream highlighting
+  dots <- list(
+    x=lazyeval::interp(~median(axis1, na.rm=T), axis1=as.name(axis1)),
+    y=lazyeval::interp(~median(axis2, na.rm=T), axis2=as.name(axis2)))
+  consensus <- res2 %>% group_by(family, role) %>% summarise_(.dots=dots)
+
+  chulls <- res2 %>% group_by(family) %>% do(.[chull(.[[axis1]], .[[axis2]]), ])
+
+  ggplot(res2, aes_string(axis1, axis2)) + geom_polygon(aes_string(axis1, axis2, fill="family", color="family"), data=chulls, alpha=0.5) + geom_point(aes(color=role)) + geom_path(aes(x, y, group=family), data=consensus) + geom_point(aes(x, y), data=consensus, size=3, color="black") + geom_point(aes(x, y, color=role), data=consensus, size=2)	+ geom_dl(aes(label=family), data=res2, list("smart.grid")) + theme_minimal() + theme(legend.position="none")
+}

### 2015-04-22 manuscript/manuscript.Rmd
Update MS with new plots

@@
   code: ucla
-word_count: 5000?
+word_count: 6000?
 ---
@@
 
-```{r sum_table}
+```{r sum_table, echo=FALSE}
 no_st <- fish_landmarks %>% filter(role != "student") %>% mutate(role = factor(role))
@@
 
-meds <- pcaed %>% group_by(family, role, mark) %>% summarise(med=median(PC1)) %>% ungroup() %>% spread(role, med) %>% transmute(family, mark, value = log(abs(turker - morphologist)), stat="median")
+meds <- pcaed %>% group_by(family, role, mark) %>%
+  summarise(med=median(PC1)) %>%
+  ungroup() %>%
+  spread(role, med) %>%
+  transmute(family, mark, value = (abs(turker - morphologist)), stat="median")
 
@@
 
-all_stats <- bind_rows(meds, mads) %>% spread(stat, value)
+all_stats <- bind_rows(meds, mads) %>% spread(stat, value) %>% group_by(family) %>% mutate(lowerq=quantile(median)[2], upperq=quantile(median)[4], iqr=IQR(median), outlier = median > iqr * 2 + upperq) %>% select(-lowerq, -upperq, -iqr)
+
+full_family <- data_frame(
+  family=c("Aca", "Apo", "Bal", "Cha", "Gob", "Lab", "Pom", "Sco", "Tet"),
+  full_family=c("Acanthuridae", "Apogonidae", "Balistidae", "Chaetodontidae", "Gobiidae", "Labridae", "Pomacentridae", "Scorpaenidae", "Tetraodontidae")
+)
+
+all_stats %<>% left_join(full_family) %>% mutate(outlier_label = ifelse(outlier, as.character(mark), NA))
 ```
 
-```{r geomorph_lm, eval=FALSE, cache=TRUE}
+```{r turker_v_expert, fig.cap="Test figure caption", anchor="figure"}
+ggplot(all_stats, aes(median, mad, color=outlier)) + geom_point() + labs(x="Accuracy", y="Precision") + facet_wrap(~full_family) + geom_dl(aes(label=outlier_label), size=1, list("top.bumptwice", dl.trans(y=y+0.1), cex=0.75)) + scale_color_manual(values=c("black", "red")) + ylim(-max(abs(all_stats$mad) + 0.25), max(abs(all_stats$mad) + 0.25)) + theme("legend.position" = "none")
+```
+
+```{r geomorph_lm, eval=FALSE, echo=FALSE, cache=TRUE}
 fmted <- format_for_gpagen(no_st)
@@
 
-```{r median_tbl, echo=F, results="asis", anchor="table", fig.cap="Median table"}
+```{r median_tbl, eval=F, echo=F, results="asis", anchor="table", fig.cap="Median table"}
 med_wide <- meds %>% spread(key=family, value=value)
@@
 
-```{r mad_tbl, echo=F, results="asis", anchor="table", fig.cap="MAD table"}
+```{r mad_tbl, eval=F, echo=F, results="asis", anchor="table", fig.cap="MAD table"}
 mad_wide <- mads %>% spread(key=family, value=value)
 mad_wide %>% knitr::kable(digits = 2, caption="MAD table")
-```
-
-```{r turker_v_expert, anchor="figure"} 
-ggplot(all_stats, aes(median, mad, label=mark)) + geom_text() + labs(x="Accuracy", y="Precision")
 ```
@@
 
-CV_sum <- left_join(CV_sum, CV_sum2)
+CV_sum <- left_join(CV_sum, CV_sum2) %>% left_join(full_family, by="family")
 
 # format for display
-CV_sum %>% transmute(`Family`=family, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.")
+CV_sum %>% transmute(`Family`=full_family, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.")
 ```
@@
 
-```{r turk_morph_procrustes_space, echo=FALSE, fig.cap="Hello", fig.keep="last", anchor="figure", message=FALSE}
+```{r, echo=FALSE}
 
@@
 ```{r timing_plot, echo=FALSE, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
-Landmarks were first received 6 minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
+Landmarks were first received `r timing_data[[1, 'time']] %>% dminutes()` after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
 # median time (in minutes) it takes an expert to measure these fishes
@@
 # data collection time plot
-fish_families %>% group_by(worker, tip) %>% summarise(time=min(time_taken)) %>% arrange(time) %>% mutate(seq = 1:n(), frac = seq / n()) %>% ggplot(aes(dseconds(time) / eminutes(), frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
+timing_data <- fish_families %>% ungroup() %>% group_by(tip) %>% summarise(time=min(time_taken)) %>% arrange(time) %>% mutate(frac=1:n()/n())
+
+ggplot(timing_data, aes(time, frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
 

### 2015-04-22 R/plotting.R
Silence import masking warnings

@@ -7,7 +7,7 @@
 #' @param axis2 The name of the second axis to plot
 #' @return A `ggplot` object
 #' @export
-#' @import dplyr
+#' @importFrom dplyr left_join group_by summarise_ do
 #' @import ggplot2
 plot_tangent_space <- function(coords, axis1="PC1", axis2="PC2") {
   x <- geomorph::two.d.array(coords)

### 2015-04-22 R/plotting.R
Fix reference to undefined variable

@@ -16,7 +16,7 @@ plot_tangent_space <- function(coords, axis1="PC1", axis2="PC2") {
 
   res <- as.data.frame(pcdata)
   res$id <- rownames(res)
-  res2 <- res %>% tidyr::separate(id, into=c("family", "role","sha"), sep="_") %>% left_join(full_family, by="family")
+  res2 <- res %>% tidyr::separate(id, into=c("family", "role","sha"), sep="_")
 
   # consensus in Procrustes space for downstream highlighting
   dots <- list(

### 2015-04-22 manuscript/manuscript.Rmd
Try to cache most chunks, and don't echo anything

--- 2015-04-22 manuscript/manuscript.Rmd
@@
 data(fish_families)
+knitr::opts_chunk$set(echo=FALSE, cache=1)
 ```
@@
 
-```{r geomorph_lm, eval=FALSE, cache=TRUE}
+```{r geomorph_lm, eval=FALSE}
 fmted <- format_for_gpagen(no_st)
@@
 
-```{r median_tbl, echo=F, results="asis", anchor="table", fig.cap="Median table"}
+```{r median_tbl, results="asis", anchor="table", fig.cap="Median table"}
 med_wide <- meds %>% spread(key=family, value=value)
@@
 
-```{r mad_tbl, echo=F, results="asis", anchor="table", fig.cap="MAD table"}
+```{r mad_tbl, results="asis", anchor="table", fig.cap="MAD table"}
 mad_wide <- mads %>% spread(key=family, value=value)
@@
 
-```{r cv_table, echo=F, results="asis", anchor="table", message=FALSE}
+```{r cv_table, results="asis", anchor="table", message=FALSE}
 
@@
 
-```{r turk_morph_procrustes_space, echo=FALSE, fig.cap="Hello", fig.keep="last", anchor="figure", message=FALSE}
-
+```{r turk_morph_procrustes_space, fig.cap="Hello", anchor="figure", message=FALSE}
 # Load data from the anonymized results file, remove students, drop bad marks
@@
 coords <- gpagen(format_for_gpagen(df), ShowPlot=F)$coords
-```
 
-Text text
-
-```{r turk_morph_procrustes_space, echo=FALSE, fig.cap="Testing my figure", anchor="figure"}
 plot_tangent_space(coords, "PC1", "PC2")
@@
 
-```{r, echo=FALSE}
+```{r brokenstick}
 # use broken stick method to get the principal components stopping point
@@
 
-```{r bamm_plots, echo=F, message=F, warning=F, anchor="figure", cache=T}
+```{r bamm_plots, message=F, warning=F, anchor="figure"}
 df <- fish_families %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))
@@
 
-```{r timing_plot, echo=FALSE, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
+```{r timing_plot, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
 Landmarks were first received 6 minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
@@
 
-```{r turker_morph_procrustes_other_pc23, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC2 and PC3 for turker / morphologist Procrustes space"}
+```{r turker_morph_procrustes_other_pc23, message=FALSE, warning=FALSE, fig.cap="PC2 and PC3 for turker / morphologist Procrustes space"}
 plot_tangent_space(coords, "PC2", "PC3")
@@
 
-```{r turker_morph_procrustes_other_pc34, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
+```{r turker_morph_procrustes_other_pc34, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
 plot_tangent_space(coords, "PC3", "PC4")
@@
 
-```{r turker_morph_procrustes_other_pc45, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PC4 and PC5 for turker / morphologist Procrustes space"}
+```{r turker_morph_procrustes_other_pc45, message=FALSE, warning=FALSE, fig.cap="PC4 and PC5 for turker / morphologist Procrustes space"}
 plot_tangent_space(coords, "PC4", "PC5")
@@
 
-```{r knitcitations_write, echo=FALSE, message=FALSE}
+```{r knitcitations_write, message=FALSE}
 write.bibtex(file="knitcitations.bib")

### 2015-04-29 manuscript/manuscript.Rmd
Code drop

@@
 
-We used the median shape of the combined experts' data as a baseline for all comparisons, as the "true" shape.
+We used the median shape of the combined experts' data as the "true" shape for all comparisons. To assess the accuracy of worker data, as compared
+
+To assess the differences between turker and expert data, we compared the median turker position to the median expert position of each landmark.
+We assumed that the expert median was indeed the true position of that landmark, and calculated the absolute Euclidian distance.
+Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy.
+We also examined the variance in turker landmarks.
+For each landmark, we rotated the cluster of points and took calculated the median absolute deviation between points along a single dimension.
+This
+
 To assess within-individual variation, we compared each worker's shape replicate to their own median shape.
 To assess within-group and between-group variation, we also compare those individual worker median shapes to the median shape of all other turkers and all other experts.
-We also measured the median absolute deviation (MAD) as a dispersion statistic for within-worker, within-group, and between-group comparisons.
-
-To test whether turkers and experts were statistically distinguishable,
+We also used the median absolute deviation (MAD) as a dispersion statistic for within-worker, within-group, and between-group comparisons.
+This is a statistic robust to outliers
 
@@
 
-```{r sum_table}
+```{r turker_v_expert, fig.cap="Test figure caption", anchor="figure"}
 no_st <- fish_landmarks %>% filter(role != "student") %>% mutate(role = factor(role))
 
-pcaed <- no_st %>% group_by(family, mark) %>% do(as_df(prcomp(cbind(.$x, .$y), scale=F, center=T)$x, role=as.character(.$role), ident=as.character(.$sha1mac))) %>% mutate(PC1 = as.numeric(PC1), PC2 = as.numeric(PC2))
+tmp <- no_st %>% filter(family == "Aca", mark %in% c("A1", "A2"))
 
-meds <- pcaed %>% group_by(family, role, mark) %>% summarise(med=median(PC1)) %>% ungroup() %>% spread(role, med) %>% transmute(family, mark, value = log(abs(turker - morphologist)), stat="median")
+accuracy <- no_st %>% group_by(family, mark) %>% summarise(median = dist_2d(x, y, role)) %>% group_by(family) %>% mutate(outlier=upper_outlier(median, extreme = 2))
 
-mads <- pcaed %>% group_by(family, role, mark) %>% summarise(mad=mad(PC1)) %>% ungroup() %>% spread(role, mad) %>% transmute(family, mark, value = log(turker / morphologist), stat="mad")
+precision <- no_st %>% group_by(role, family, mark) %>% summarise(var = mad_2d(x, y)) %>% spread(role, var) %>% group_by(family, mark) %>% transmute(mad = log(turker / morphologist))
 
-all_stats <- bind_rows(meds, mads) %>% spread(stat, value)
+full_family <- data_frame(
+  family=c("Aca", "Apo", "Bal", "Cha", "Gob", "Lab", "Pom", "Sco", "Tet"),
+  full_family=c("Acanthuridae", "Apogonidae", "Balistidae", "Chaetodontidae", "Gobiidae", "Labridae", "Pomacentridae", "Scorpaenidae", "Tetraodontidae")
+)
+
+all_stats <- left_join(accuracy, precision) %>% left_join(full_family) %>% mutate(outlier_label = ifelse(outlier, as.character(mark), NA))
+
+ggplot(all_stats, aes(median, mad, color=outlier)) + geom_point() + labs(x="Accuracy", y="Precision: log2 turker variance / expert variance ") + facet_wrap(~full_family)# + geom_dl(aes(label=outlier_label), size=1, list("top.bumptwice", dl.trans(y=y+0.1), cex=0.75)) + scale_color_manual(values=c("black", "red")) + ylim(-max(abs(all_stats$mad) + 0.25), max(abs(all_stats$mad) + 0.25)) + theme("legend.position" = "none")
 ```
@@
 
-```{r median_tbl, results="asis", anchor="table", fig.cap="Median table"}
+```{r median_tbl, eval=F, results="asis", anchor="table", fig.cap="Median table"}
 med_wide <- meds %>% spread(key=family, value=value)
@@
 
-```{r mad_tbl, results="asis", anchor="table", fig.cap="MAD table"}
+```{r mad_tbl, eval=F, results="asis", anchor="table", fig.cap="MAD table"}
 mad_wide <- mads %>% spread(key=family, value=value)
 mad_wide %>% knitr::kable(digits = 2, caption="MAD table")
-```
-
-```{r turker_v_expert, anchor="figure"}
-ggplot(all_stats, aes(median, mad, label=mark)) + geom_text() + labs(x="Accuracy", y="Precision")
 ```
@@
 
-plot_tangent_space(coords, "PC1", "PC2")
+plot_tangent_space(pca_df(coords), "PC1", "PC2")
 ```
@@
 
-res <- plotTangentSpace(full_coords, groups=groups, axis1=1, axis2=2, verbose=T, warpgrids=F)$pc.scores
+pcaed <- pca_df(full_coords) %>% separate(id, into=c("family", "name"), sep="\\|", extra="drop")
 
-res <- as.data.frame(res)
-res$id <- rownames(res)
-res2<- res %>% separate(id, into=c("family", "name"), sep="\\|", extra="drop") %>% tbl_df()
+pcaed %>% select(name, PC1) %>% write.table(file="lol.txt", quote=F, row.names=F, col.names=F, sep="\t")
 
-p <- ggplot(res2, aes(x=PC1, y=PC2, col=family)) + geom_point() + theme(legend.position="none")
-
-res2 %>% select(name, PC1) %>% write.table(file="lol.txt", quote=F, row.names=F, col.names=F, sep="\t")
+plot_tangent_space2(pcaed, labelmethod=list("chull.grid"))
 ```
@@
 ```{r timing_plot, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
-Landmarks were first received 6 minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
+Landmarks were first received `r timing_data[[1, 'time']] %>% dminutes()` after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
 # median time (in minutes) it takes an expert to measure these fishes
-timings <- (no_st %>% group_by(role) %>% summarise(duration=median(duration))) %>% mutate(total_time=duration * nrow(res2))
+timings <- (no_st %>% group_by(role) %>% summarise(duration=median(duration))) %>% mutate(total_time=duration * nrow(no_st))
 
@@
 
-Using a median-expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(res2))` to landmark all `r nrow(res2)` images.
-At 5x replication, this would take `r as.character(morpho_per_image * nrow(res2) * 5)`.
+Using a median-expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(no_st))` to landmark all `r nrow(no_st)` images.
+At 5x replication, this would take `r as.character(morpho_per_image * nrow(no_st) * 5)`.
 By comparison, turkers took a total of `r as.character(total_turker_time)` to complete all images at 5x replication.
@@
 ```{r turker_morph_procrustes_other_pc23, message=FALSE, warning=FALSE, fig.cap="PC2 and PC3 for turker / morphologist Procrustes space"}
-plot_tangent_space(coords, "PC2", "PC3")
+coords_df <- pca_df(coords)
+plot_tangent_space(coords_df, "PC2", "PC3")
 ```
@@
 ```{r turker_morph_procrustes_other_pc34, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
-plot_tangent_space(coords, "PC3", "PC4")
+plot_tangent_space(coords_df, "PC3", "PC4")
 ```
@@
 ```{r turker_morph_procrustes_other_pc45, message=FALSE, warning=FALSE, fig.cap="PC4 and PC5 for turker / morphologist Procrustes space"}
-plot_tangent_space(coords, "PC4", "PC5")
+plot_tangent_space(coords_df, "PC4", "PC5")
 ```

### 2015-04-29 R/functions.R

@@ -18,28 +18,26 @@ ip.lda <- function(object, newdata) {
   predict(object, newdata=newdata)$class
 }
 
-#' Formats a data frame for `geomorph::gpagen`
+#' Formats a data frame for \code{\link[geomorph]{gpagen}}
 #'
 #' @param df a data frame, with certain columns
 #' @return A 3D matrix
-#' @importFrom tidyr gather unite
-#' @importFrom magrittr %>%
+#' @import dplyr
 #' @export
 format_for_gpagen <- function(df) {
-  splut <- gather(df, variable, val, x:y) %>% unite(col=id, family, role, sha1mac)
+  splut <- tidyr::gather(df, variable, val, x:y) %>% tidyr::unite(col=id, family, role, sha1mac)
   xtabs(val ~ mark + variable + id, data=splut)
 }
 
 
 #' Hack the shit out of geomorph's slow procD.lm function
 #' @export
-#' @import geomorph
 my.procD.lm <- function (f1, iter = 999, RRPP = FALSE, int.first = FALSE, verbose = FALSE)
 {
   form.in <- formula(f1)
   Y <- eval(form.in[[2]], parent.frame())
   if (length(dim(Y)) == 3)
-    Y <- two.d.array(Y)
+    Y <- geomorph::two.d.array(Y)
   else Y <- as.matrix(Y)
   if (nrow(Y) != nrow(model.frame(form.in[-2])))
     stop("Different numbers of specimens in dependent and independent variables")

### 2015-04-29 R/plotting.R

@@ -1,22 +1,37 @@
-#' Plot specimens in tangent space
+#' Do PCA on Procrustes-aligned specimens
 #'
-#' This function plots a set of Procrustes-aligned specimens in tangent space along their principal axes. This is a custom replacement for `geomorph::plotTangentSpace` tailored for the fish reliability project.
-#'
-#' @param coords An array (p x k x n) containing landmark coordinates for a set of aligned specimens, possibly from `geomorph::gpagen`. Array labels must be in the format `group1_group2_individual`.
-#' @param axis1 The name of the first axis to plot
-#' @param axis2 The name of the second axis to plot
-#' @return A `ggplot` object
+#' @param coords An array (p x k x n) containing landmark coordinates for a set of aligned specimens, possibly from \link{geomorph::gpagen}.
+#' @param rename_rownames Rename and remove the rownames? If TRUE (the default), converts rownames to a column in the data frame called `id`. If a character string, instead calls that column by that string. If FALSE, preserves the rownames as-is.
+#' @return an object of class `tbl_df`, `tbl`, and `data.frame`, with columns PC1 to PC...
 #' @export
-#' @importFrom dplyr left_join group_by summarise_ do
-#' @import ggplot2
-plot_tangent_space <- function(coords, axis1="PC1", axis2="PC2") {
+#' @import dplyr
+pca_df <- function(coords, rename_rownames = TRUE) {
   x <- geomorph::two.d.array(coords)
   pc.res <- stats::prcomp(x)
   pcdata <- pc.res$x
 
   res <- as.data.frame(pcdata)
-  res$id <- rownames(res)
-  res2 <- res %>% tidyr::separate(id, into=c("family", "role","sha"), sep="_")
+  if (rename_rownames) {
+    if (rename_rownames == TRUE) rename_rownames <- "id"
+    res[[rename_rownames]] <- rownames(res)
+    rownames(res) <- NULL
+  }
+  tbl_df(res)
+}
+
+#' Plot specimens in tangent space
+#'
+#' This function plots a set of Procrustes-aligned specimens in tangent space along their principal axes. This is a custom replacement for \code{geomorph::plotTangentSpace} tailored for the fish reliability project.
+#'
[…]

### 2015-04-29 R/stats.R

@@ -1,0 +1,46 @@
+#' 2D median absolute deviation
+#'
+#' Compute the median absolute deviation for two dimensional data.
+#' @param x a numeric vector.
+#' @param y a numeric vector, of the same length as \code{x}.
+#' @param constant scale factor. See 'Details' in \code{\link[stats]{mad}}.
+#' @export
+#' @importFrom stats median
+mad_2d <- function(x, y, constant = 1.4826) {
+  if (length(x) != length(y))
+    stop("'x' and 'y' must have the same length")
+  cx <- median(x)
+  cy <- median(y)
+
+  constant * median(sqrt((x - cx)^2 + (y-cy)^2))
+}
+
+#' Median distance of two groups
+#'
+#' @param x a numeric vector
+#' @param y a numeric vector
+#' @param group a vector that specifies how \code{x} and \code{y} should be grouped
+#' @export
+#' @import dplyr
+#' @importFrom stats median dist
+dist_2d <- function(x, y, group) {
+  if (length(x) != length(y))
+    stop("'x' and 'y' must have the same length")
+  x <- as.numeric(x)
+  y <- as.numeric(y)
+  df <- data_frame(x, y, group) %>% group_by(group) %>% summarise(x=median(x), y=median(y)) %>% select(x,y)
+  dist(df)[1]
+}
+
+#' Detects an upper (more positive) outlier
+#'
+#' @param x a numeric vector
+#' @param y the cutoff point for determining an outlier
+#' @export
+#' @importFrom stats IQR quantile
+upper_outlier <- function (x, extreme = 2) {
+  lowerq <- quantile(x)[2]
+  upperq <- quantile(x)[4]
+  iqr <- IQR(x)
+  x > iqr * extreme + upperq
+}

### 2015-04-29 manuscript/manuscript.Rmd
Move JSON stuff to supplement

@@
 
-![A screenshot of the web app that turkers used to digitize images. A live demonstration is available at <https://jonchang.github.io/eol-mturk-landmark/>](interface.png)
-
 We developed an application that runs completely on the user's local web browser.
 This simplifies the infrastructure challenge of needing to serve many crowdsdourced workers simultaneously.
-The web application is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task.
-Below is an example JSON file that demonstrates the utility of our web app:
-
-```json
-{
-    "C": {
-        "kind": "point",
-        "help": "Click the center of the eye."
-    },
-    "D": {
-        "kind": "line",
-        "help": "Click and drag from the left edge of the eye
-                 to the right edge of the eye."
-    },
-    "O": {
-        "kind": "curve",
-        "help": "Click and drag over the outline of the eye,
-                 starting from the leftmost point of the eye."
-    }
-}
-```
-
-Each digitization task to be completed is given a short abbreviation to aid task identification ("C" for center, "D" for diameter, and "O" for outline), and the type of task, "point", "line" or "curve", for homologous landmarks, linear measurements, and sliding semilandmarks can be specified.
-There is also an optional short help snippet displayed inline, which serve as a brief reminder for each landmark and complements a larger and more detailed protocol document that workers are required to read before beginning work (Supplementary Material).
+The web application is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task (for more details, see Supplementary Material).
 
@@
 
-XXX: Best practices for turking
+Below is an example JSON file that demonstrates the utility of our web app:
+
+```json
+{
+    "C": {
+        "kind": "point",
+        "help": "Click the center of the eye."
+    },
+    "D": {
+        "kind": "line",
+        "help": "Click and drag from the left edge of the eye
+                 to the right edge of the eye."
+    },
+    "O": {
+        "kind": "curve",
+        "help": "Click and drag over the outline of the eye,
+                 starting from the leftmost point of the eye."
+    }
+}
+```
+
+Each digitization task to be completed is given a short abbreviation to aid task identification ("C" for center, "D" for diameter, and "O" for outline), and the type of task, "point", "line" or "curve", for homologous landmarks, linear measurements, and sliding semilandmarks can be specified.
+There is also an optional short help snippet displayed inline, which serve as a brief reminder for each landmark and complements a larger and more detailed protocol document that workers are required to read before beginning work.
+
+![A screenshot of the web app that turkers used to digitize images. A live demonstration is available at <https://jonchang.github.io/eol-mturk-landmark/>](interface.png)
 

### 2015-04-29 manuscript/manuscript.Rmd
Fix up figures

@@
     fig_caption: yes
+    fig_height: 5
+    fig_width: 7
     keep_tex: yes
@@
 library("BAMMtools")
+library("knitr")
+library("kfigr")
 library("knitcitations")
-library("kfigr")
 library("fish.reliability")
@@
 data(fish_families)
-knitr::opts_chunk$set(echo=FALSE, cache=1)
+knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center")
 ```
@@
 
-```{r turk_morph_procrustes_space, fig.cap="Hello", anchor="figure", message=FALSE}
+```{r turk_morph_procrustes_space, fig.cap="Turker morpho space", anchor="figure"}
 # Load data from the anonymized results file, remove students, drop bad marks
@@
 
-```{r bamm_plots, message=F, warning=F, anchor="figure"}
+```{r bamm_plots, message=F, warning=F, fig.show="hold", fig.cap="Family figures", anchor="figure"}
 df <- fish_families %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))
@@
 
+asdfsdf
+
 ```{r timing_plot, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
-Landmarks were first received `r timing_data[[1, 'time']] %>% dminutes()` after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark."}
+Landmarks were first received 8 minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark.", anchor="figure"}
 # median time (in minutes) it takes an expert to measure these fishes
@@
 
+<!-- reset figure numbering for supplement -->
+
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+
+```{r}
+knitr::opts_chunk$set(fig.cur=1)
+```
+
 Below is an example JSON file that demonstrates the utility of our web app:
@@
 
+<!-- spit out all figures before references -->
+\clearpage
+
 # References

### 2015-04-30 R/zzzdata.R
Add Makefile to transform raw data

@@ -16,7 +16,7 @@
 #'   \item{duration}{how long, in seconds, this image took to landmark}
 #'   \item{family}{the family that this specimen belonged to}
 #' }
-"fish_landmarks"
+"fish_reliability"
 
 
 

### 2015-04-30 data-raw/fish_families.R

@@ -1,0 +1,27 @@
+#!/usr/bin/env Rscript
+
+# anonymizes and cleans up results from the mturk results file
+
+library(readr)
+library(dplyr)
+library(stringr)
+library(rjson)
+library(tidyr)
+library(magrittr)
+library(lubridate)
+library(digest)
+library(devtools)
+
+secret <- read_lines("../../SECRET_KEY")
+
+results <- read_tsv("fishturk.results") %>% filter(Answer.marks != "") %>% mutate(tip=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 3], family=str_match(annotation, ".*/(.*)/([A-Za-z]+_[A-Za-z]+)")[, 2], Answer.marks=str_replace_all(Answer.marks, '""', '"'), start=parse_date_time(creationtime, "bdHMSY"), end=parse_date_time(assignmentsubmittime, "bdHMSY"), time_taken = end - start) %>% select(-start, -end)
+
+get_marks <- function(blob, ...) {
+  fromJSON(blob) %>% as.data.frame %>% t %>% as.data.frame %>% set_colnames(c("x", "y")) %>% mutate(mark=rownames(.)) %>% cbind(...)
+}
+
+fish_families <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=hmac(secret, .$workerId, algo="sha1"), family=.$family, tip=.$tip, time_taken=.$time_taken))
+
+write_csv(fish_families, path = "fish_families.csv")
+
+use_data(fish_families, overwrite = TRUE)

### 2015-04-30 data-raw/fish_reliability.R

@@ -1,0 +1,64 @@
+#!/usr/bin/env Rscript
+
+# anonymizes and cleans up results from a sqlite file
+
+library(stringr)
+library(reshape2)
+library(plyr)
+library(dplyr)
+library(rjson)
+library(tidyr)
+library(lubridate)
+library(digest)
+library(devtools)
+library(readr)
+
+secret <- read_lines("../../SECRET_KEY")
+morphos <- read_lines("../../MORPHOS")
+
+tb <- src_sqlite(path="reliability.sqlite")
+
+log <- tbl(tb, "log")
+plan <- tbl(tb, "plan")
+
+res <- plan %>% left_join(log, by="assignmentId") %>% collect()  %>% group_by(workerId) %>% filter(type=="completed", max_sequence-1 == max(sequence)) %>% distinct(workerId, sequence)
+
+alldata <- res %>% mutate(role=ifelse(workerId %in% morphos, "morphologist", ifelse(str_detect(workerId, "^A"), "turker", "student")), sequence=as.numeric(sequence))
+
+get_marks <- function(blob, name) {
+	jsoned <- fromJSON(blob)
+	marks <- fromJSON(jsoned$marks) %>% lapply(unlist) %>% melt() %>%
+	    rename(mark = L1) %>%	filter(str_detect(mark, "^[JEOPACD]")) %>%
+			group_by(mark) %>% transmute(variable=c("x", "y"), value) %>%
+			ungroup() %>% spread(variable, value) %>% cbind(name)
+	marks
+}
+
+get_duration <- function(blob, name) {
+	jsoned <- fromJSON(blob)
+	logger <- fromJSON(jsoned$logger) %>% lapply(unlist) %>% lapply(data.frame) %>% lapply(t) %>%
+		do.call(plyr::rbind.fill.matrix, .) %>% data.frame(stringsAsFactors=F) %>% tbl_df() %>%
+		unite(payload, -X1, -X2, sep=",") %>% rename(timestamp=X1, action=X2) %>%
+		mutate(timestamp = as.POSIXct(as.numeric(timestamp)/1000, origin="1970-01-01 00:00.00 UTC")) %>%
+		filter(str_detect(action, "^data")) %>%
+		summarise(duration=last(timestamp)-first(timestamp)) %>% cbind(name)
+	logger
+}
+
+get_urls <- function(blob, name) {
+	jsoned <- fromJSON(blob)
[…]

### 2015-04-30 manuscript/manuscript.Rmd
Update manuscript data refs

@@
 set.seed(111) # reproducible and stable sampling
-data(fish_landmarks)
+data(fish_reliability)
 data(fish_families)
@@
 ```{r turker_v_expert, fig.cap="Test figure caption", anchor="figure"}
-no_st <- fish_landmarks %>% filter(role != "student") %>% mutate(role = factor(role))
+no_st <- fish_reliability %>% filter(role != "student") %>% mutate(role = factor(role))
 
@@
 # Load data from the anonymized results file, remove students, drop bad marks
-df <- fish_landmarks %>%
+df <- fish_reliability %>%
 	filter(role != "student") %>% 
@@
 
-```{r bamm_plots, message=F, warning=F, fig.show="hold", fig.cap="Family figures", anchor="figure"}
+```{r sevenfamily_tangent_plot, message=F, warning=F, fig.show="hold", fig.cap="Family figures", anchor="figure"}
 df <- fish_families %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))
@@
 
-We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frédérich, Vanson Liu, Rosalia Aguilar, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the `r (fish_landmarks %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (fish_families %>% select(worker) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
+We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frédérich, Vanson Liu, Rosalia Aguilar, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the `r (fish_reliability %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (fish_families %>% select(worker) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
 We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for insightful discussions about fish morphology and digitization protocols.
-Katie Staab, Tonya Kane, and their `r fish_landmarks %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students served as beta testers for the methods in this study.
+Katie Staab, Tonya Kane, and their `r fish_reliability %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students served as beta testers for the methods in this study.
 This work was supported by an Encyclopedia of Life David M.

### 2015-04-30 manuscript/manuscript.Rmd
Update for new data workflow

@@
 library("scales")
-library("grid")
 library("geomorph")
@@
 -->
-
-XXX: Use r package captioner for captions
 
@@
 no_st <- fish_reliability %>% filter(role != "student") %>% mutate(role = factor(role))
-
-tmp <- no_st %>% filter(family == "Aca", mark %in% c("A1", "A2"))
 
@@
 # Also download Mendeley bibliography
-mendbib <- RCurl::getURL("https://www.dropbox.com/s/ehwm3c945k7rh0n/library.bib?dl=1", .opts=RCurl::curlOptions(followlocation=TRUE))
+mendbib <- RCurl::getURL("https://www.dropbox.com/s/ehwm3c945k7rh0n/library.bib?dl=1", .opts=RCurl::curlOptions(followlocation=TRUE), ssl.verifypeer = FALSE)
 cat(mendbib, file="library.bib")

### 2015-04-30 data-raw/fish_reliability.R

@@ -57,7 +57,7 @@ allurls <- alldata %>% distinct(workerId, list) %>% rowwise() %>% do(get_urls(.$
 
 hugedata <- ungroup(alldata) %>% transmute(workerId, sequence=as.integer(sequence), role) %>% inner_join(allmarks) %>% inner_join(alldurations) %>% inner_join(allurls)
 
-fish_reliability <- hugedata %>% select(workerId) %>% distinct(workerId) %>% rowwise() %>% do(data.frame(workerId=.$workerId, sha1mac=hmac(secret, .$workerId, algo="sha1"))) %>% left_join(hugedata) %>% mutate(fn=str_match(url, "([-a-zA-Z0-9_])+\\.jpg$")[, 1]) %>% select(-url, -workerId)
+fish_reliability <- hugedata %>% select(workerId) %>% distinct(workerId) %>% rowwise() %>% do(data.frame(workerId=.$workerId, sha1mac=hmac(secret, .$workerId, algo="sha1"))) %>% left_join(hugedata) %>% mutate(fn=str_match(url, "([-a-zA-Z0-9_])+\\.jpg$")[, 1]) %>% separate(fn, into="family", remove = T, extra = "drop") %>% select(-url, -workerId)
 
 write_csv(fish_reliability, path = "fish_reliability.csv")
 

### 2015-09-16 manuscript/manuscript.Rmd
Code drop -- preprint version

@@
 documentclass: inc/msarticle
+endfloat: yes
 fontsize: 11pt
@@
 lineno: yes
-linestretch: 1.05
+linestretch: 2
 csl: inc/methods-in-ecology-and-evolution.csl
@@
   code: ucla
-word_count: 6000?
+word_count: 7000
 ---
@@
 library("knitcitations")
+library("parallel")
 library("fish.reliability")
@@
 We find that crowdsourced workers perform similarly to experienced morphologists on the same digitization tasks.
-We also demonstrate the speed and accuracy of our method on 7 families of ray-finned fishes (Actinopterygii), and find that rates of body shape evolution are correlated with diversification rates.
+We also demonstrate the speed and accuracy of our method on seven families of ray-finned fishes (Actinopterygii).
 
 4.
-Crowdsourcing will enable the collection of morphological data across vast radiations of organisms, and will facilitate richer inference on the macroevolutionary processes that shape phenotypic diversity across the tree of life.
+Crowdsourcing will enable the collection of morphological data across vast radiations of organisms, and can facilitate richer inference on the macroevolutionary processes that shape phenotypic diversity across the tree of life.
+
+**Keywords**: crowdsourcing, morphometrics, phenotyping, morphology, comparative methods, macroevolution, Actinopterygii
 
@@
 
-Integrating phenotypic data, such as anatomy, behavior, physiology, and other traits, with phylogenies is powerful workflow for investigating the patterns of biological evolution.
+Integrating phenotypic data, such as anatomy, behavior, physiology, and other traits, with phylogenies is powerful strategy for investigating the patterns of biological evolution.
 Recent advances in next-generation sequencing [@Shendure2008; @Meyer2008] and sequence capture technologies [@Faircloth2012b; @Lemmon2012] have made phylogenetic inference of large radiations of organisms possible [@McCormack2012; @McCormack2013; @Faircloth2013; @Faircloth2014].
-These large phylogenies have been successfully used in conjunction with existing comprehensive phenotypic datasets to answer key questions about the mode and tempo of diversification and morphological evolution [@Rabosky2013].
 However, similar breakthroughs for generating new phenotypic datasets have been comparatively uncommon, likely due to the high expense and effort required [reviewed in @Burleigh2013].
@@
 Creating these large phenotypic datasets has generally required an extended dedicated effort of measuring and describing morphological or behavioral traits that are then coded into a comprehensive data matrix.
-One such example is the Phenoscaping project [<http://kb.phenoscape.org>; @Deans2015], and its complementary efforts in the Vertebrate Taxonomy Ontogeny [@Midford2013] and Hymenoptera Anatomy Ontology [@Yoder2010], which requires large amounts of researcher effort to compile.
-Other approaches include using machine learning [@Mckinney2013a] or natural language processing [@Cui2012a] to identify or infer phenotypes.
-These statistical techniques function ideally with either a large training dataset or a complex model [@Brill2003; @Halevy2009; @Hastie2009], both of which also require intensive researcher effort to build and validate.
+One such example is the Phenoscaping project [<http://kb.phenoscape.org>; @Deans2015], and related efforts in the Vertebrate Taxonomy Ontogeny [@Midford2013] and Hymenoptera Anatomy Ontology [@Yoder2010], which require large amounts of researcher effort to collate.
+Other approaches include using machine learning [@Dececchi2015], machine vision [@Corney2012; @Corney2012a], or natural language processing [@Cui2012a] to identify or infer phenotypes.
+These statistical techniques function ideally with either a large training dataset (e.g., a predefined ontogeny database) or a complex model [@Brill2003; @Halevy2009; @Hastie2009], both of which also require intensive researcher effort to build and validate.
 Finally, methods such as high-throughput infrared imaging, mass spectrometry, and chromatography have been successfully used in plant physiology [@Furbank2011]
@@
 These approaches all share a similar goal of collecting large comparative datasets, but also require large investments in researcher effort.
-This bottleneck has limited the taxonomic breadth and depth of research in comparative biology.
+This bottleneck in researcher availability has limited the scope of work in comparative biology.
 
-Here we present a method and toolkit to efficiently collect two-dimensional geometric morphometric phenotypic data at a "phylogenomic" scale. 
+Although it is now possible to build phylogenetic trees with thousands of tips, and phenotypic data sets have similarly been growing larger and larger, the traits that are typically studied at this scale tend to be simple: geographic occurrences [@Jetz2012], one or two continuous characters [@Harmon2010; @Rabosky2013], a single discrete character [@Goldberg2010; @Edwards2011; @Price2012], or some combination of these [@Zanne2014; @Pyron2014].
+A richer understanding of the forces that shape macroevolution requires the collection of more detailed phenotypic trait data at scale.
+
+Here we present a method and toolkit to efficiently collect two-dimensional geometric morphometric phenotypic data at a high-throughput "phenomic" scale. 
 We developed a novel web browser-based image landmarking application, and use Amazon Mechanical Turk (<https://www.mturk.com>) to distribute digitization tasks to remote workers (hereafter *turkers*) over the Internet, who are paid for their contributions.
 We evaluate the accuracy and precision of turkers by assigning identical image sets and digitization protocols to users who are experienced with fish morphology (hereafter *experts*), and compare the inter- and intra-observer differences between turkers and experts.
-As an example of how this new method may be used, we construct a phylogenetic analysis pipeline to download photographs and phylogenies of 7 actinopterygiian families from the web, collect Mechanical Turk shape results, and analyze the rate of diversification and body shape evolution using BAMM [@Rabosky2014a].
+To illustrate the efficiency of this approach, we construct a phylogenetic analysis pipeline to download photographs and phylogenies of seven actinopterygiian families from the web, collect Mechanical Turk shape results, analyze the rate of diversification and body shape evolution using BAMM [@Rabosky2014a], and compare the time required for this workflow to traditional approaches.
+We also discuss the role that crowdsourcing is best suited in large-scale morphological analyses, and suggest ways to integrate crowdsourced data as part of larger initiatives to digitize biodiversity.
 
@@
 We use MTurk as a platform to distribute our geometric morphometric tasks and financially compensate the worker accordingly.
-Scientific collection of data over MTurk and similar services has generally been limited to the fields of psychology and computer science, and there have been few attempts to crowdsource biological trait data [but see @Burleigh2013].
+Scientific collection of data over MTurk and similar services has generally been limited to the fields of psychology and computer science, and there have been few attempts to crowdsource biological trait data [@Burleigh2013].
 
@@
 
-We developed an application that runs completely on the user's local web browser.
-This simplifies the infrastructure challenge of needing to serve many crowdsdourced workers simultaneously.
-The web application is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task (for more details, see Supplementary Material).
+We developed an geometric morphometric digitization application that runs completely on the user's local web browser, using the HTML5 Canvas interface.
+This simplifies the infrastructure challenge of needing to serve many crowdsourced workers simultaneously, since workers will not need to download desktop software such as tpsDig (<http://life.bio.sunysb.edu/ee/rohlf/software.html>) before generating data.
+The web application is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task (Supplemental Figure S1).
+Point landmarks, semilandmark curves, and linear measurements are all supported.
+The software is available at <https://github.com/jonchang/eol-mturk-landmark>.
 
-Although digitizing and landmarking a single image [microtasks *sensu* @Good2013] is most effective for high-throughput work on MTurk, it is unsuitable for conducting controlled experiments.
-To solve this issue we created a server-side application backend that automatically distribute tasks according to a configurable set of images and experimental protocol.
-This application also mimics an official Amazon Mechanical Turk endpoint, to decouple the experimental backend from the image landmarking frontend.
-External non-MTurk workers can also participate in the same experiment, ensuring effective comparisons across these groups.
-The software is open source and uses PHP and SQLite, and can be installed on nearly all web hosting services (<https://github.com/jonchang/fake-mechanical-turk>).
+Although digitizing and landmarking a single image [microtasks *sensu* @Good2013] is effective for high-throughput work on MTurk, it is unsuitable for conducting controlled experiments.
+To solve this issue we also created a server-side application backend that automatically distributes tasks according to a configurable set of images and experimental protocol.
+This application mimics an official Amazon Mechanical Turk interface endpoint, to facilitate drop-in replacement for an existing MTurk workflow.
+External non-MTurk workers can also participate in the same experiment, ensuring consistent comparisons across separate groups.
+The software is available at <https://github.com/jonchang/fake-mechanical-turk>.
 
@@
 Collecting landmark-based geometric morphometric data at scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007].
-To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked both turkers and experts to landmark a set of 5 fish images, 5 times each.
+To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked turkers ($n = `r nrow(filter(fish_reliability, role == "turker") %>% distinct(sha1mac))`$) and experts ($n = `r nrow(filter(fish_reliability, role == "morphologist") %>% distinct(sha1mac))`$) to landmark a set of five fish images, five times each.
 All participants used the same protocol and same software to digitize the same set of fishes.
+The landmarks were carefully selected based on previously-published literature concerning fish shape [Supplemental Figure S2; @Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Claverie2014; @Thacker2014].
+We also ensured that the chosen landmarks included morphological features that were relatively straightforward to digitize (the position of the eye) and features that were likely to be more challenging to digitize (the position of the preopercle bone), in order to test for turker and expert differences over a spectrum of difficulties.
+We report the inter-observer reliability for turkers and experts by computing the ratio of the among-individual and the sum of the among-individual and measurement error variance components in a repeated measures nested MANOVA [@Palmer1986; @Zelditch2012].
 
-The landmarks were chosen according to previously-published literature concerning fish shape [@Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Thacker2014], and included landmarks both easy to digitize, such as the anterior margin of the eye, and challenging to digitize, such as the opercular series.
-We used 5 replicate measurements per worker per image to quantify measurement error within and among observers [@Palmer1986] using *geomorph* `r citep(citation("geomorph"))`.
+To assess the differences between turker and experts on a per-landmark basis, we first compared the median turker position to the median expert position of each landmark.
+We assumed that the expert median was the true position of that landmark, and calculated the absolute Euclidian distance.
+Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy.
+We then examined the variance in turker landmarks.
+For each landmark, we rotated the cloud of points to maximize variance in one dimension, and calculated the log-ratio of median absolute deviations (MAD) between turkers and experts.
+This rotation is a conservative approach for assessing the difference in variance between these two groups, because it maximizes any apparent differences in landmark position.
+A positive log-ratio indicated that experts had lower variance than turkers, while a negative log-ratio indicated that turkers had lower variance.
+For all subsequent analysis, we excluded landmarks where turkers performed especially poorly, where either the accuracy or precision components for a given landmark exceeded 1.5 times the interquartile range of that component.
 
-We used the median shape of the combined experts' data as the "true" shape for all comparisons. To assess the accuracy of worker data, as compared
+To determine whether turkers and experts were statistically distinguishable, we performed a non-parametric MANOVA using the randomized residual permutation procedure (RRPP) with 1,000 iterations [@Collyer2014].
+The RRPP method reduces the effect of the "curse of dimensionality" ($p >> n$, where the number of predictors greatly exceeds the number of observations), a common problem in geometric morphometrics, and has been shown to have increased statistical power compared to a method where the raw data are randomized instead [@Anderson2003].
+We test for a difference between mean turker and expert shapes against a null model of no difference between turker and expert changes, taking into account species-specific differences. 
+A difference between models was considered significant if the p-value was less than $\alpha = 0.05$.
 
-To assess the differences between turker and expert data, we compared the median turker position to the median expert position of each landmark.
-We assumed that the expert median was indeed the true position of that landmark, and calculated the absolute Euclidian distance.
-Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy.
-We also examined the variance in turker landmarks.
-For each landmark, we rotated the cluster of points and took calculated the median absolute deviation between points along a single dimension.
-This
+As a separate test, we use linear discriminant analysis [LDA, @Ripley1996], a statistical classification algorithm that finds features to differentiate between different classes of data, in this case turkers and experts.
+We assessed the accuracy of the LDA classification using 10-fold cross validation (CV), which splits our data into 10 equally-sized groups, using nine for training and one for validation [@Kohavi1995; @Hastie2009].
+An acceptable misclassification rate varies depends on application, but here we use a 25% misprediction rate as a standard for sufficient accuracy.
+This is a highly forgiving standard, since a 50% misprediction rate is no better than a coin flip, and a 25% misprediction rate would still erroneously classify one in four turkers as experts or vice versa.
+We also use quadratic discriminant analysis (QDA), which relaxes some of the assumptions of LDA, and similarly report the QDA misclassification rate.
 
-To assess within-individual variation, we compared each worker's shape replicate to their own median shape.
-To assess within-group and between-group variation, we also compare those individual worker median shapes to the median shape of all other turkers and all other experts.
-We also used the median absolute deviation (MAD) as a dispersion statistic for within-worker, within-group, and between-group comparisons.
-This is a statistic robust to outliers
-
-The curse of dimensionality is a common problem in machine learning, where $p$, the number of predictors, greatly exceeds $n$, the number of observations.
-Geometric morphometric datasets are typically extremely highly dimensional, with $2 * k - 1$ observations per 2D landmark.
-To reduce the dimensionality of our dataset, we orthogonalize our features used a generalized Procrustes superimposition as implement in the R package geomorph.
-We then use logistic regression on the PC axes, starting with PC1 and adding each subsequent axis in a stepwise fashion.
-We validate each model using $K$-fold cross-validation, which splits the data into $K$ equally sized groups, using $K-1$ groups to train the model and the last group to validate the model.
-This process is repeated for each subset of the groups.
-We set $K=10$ to validate the model, for its balanced statistical properties [@Kohavi1995; @Hastie2009].
-Using 10-fold CV, we calculated the confusion matrix for each model in turn, and plot the sensitivity vs specificify (also known as the receiver operator characteristic curve).
-
-To reduce the effect of the curse of dimensionality ($p >> n$, where the number of predictors  exceeds the number of observations), we first
-
-<!-- (XXX: Talk about setting the consensus morphologist shape as the "true" shape in terms of determining accuracy / precision of individual workers and the consensus turker shape) -->
-
-To determine whether turkers and experts were statistically distinguishable, we used the machine learning technique linear discriminant analysis [LDA, @Ripley1996] as implemented in *MASS* `r citep(citation("MASS"))`.
-A common error in assessing the efficacy of machine learning methods is training the model and then testing the model using the same data set for both.
-The researcher can make the model increasingly sophisticated to better fit the training dataset, but be unable to make accurate inferences to real-world data, due to overfitting.
-Conversely, withholding too much data for validation results in underfitting and the inability of the model to generalize its predictions to real data.
-To guard against these potentially misleading inferences, we used $K$-fold cross-validation (CV), which splits our data into $K$ equally-sized groups, using $K-1$ for training and $1$ for validation, and tests the model for all possible permutations of these groups.
-We used a value of $K=10$ to validate the LDA model, for its balanced approach to the bias-variance tradeoff [@Kohavi1995; @Hastie2009].
-Using 10-fold CV, we assessed the LDA model's misprediction rate.
-Acceptable misprediction rates vary depending upon the application, here we judge 20% misprediction rate to sufficently distinguish turkers and experts, while a 50% or higher rate is no better than a random guess.
-We also used quadratic discriminant analysis (QDA), to test the robustness of our results with a different model.
-All analyses were performed in R `r citep(citation())`.
-(XXX: How to interpret the result of this method.
-Use R package caret for easier comparisons among different machine learning techniques.
-Try out support vector machines.)
-
-We calculated the per-digitizer median shape for each image used, as well as the consensus turker and morphologist landmarks, and projected these shapes into Procrustes space.
-We calculated the 95% confidence ellipse to estimate the uncertainty around these shape estimates.
-(XXX: How to interpret the result of this method)
+We calculated the per-individual median shape for each species used, as well as the consensus turker and morphologist shapes, and projected these shapes into Procrustes space, to visualize the orthogonalized differences in median shape among and between the types of digitizers.
 
@@
 
-A common strategy in fish comparative studies is to examine evolutionary dynamics within a single family (XXX: cite more papers that do this) [@Frederich2013; @Thacker2014], potentially due to the extensive amount of time necessary to collect data.
+A common strategy in fish comparative studies is to examine evolutionary dynamics within a single family [@Ferry-Graham2001a; @Alfaro2005; @Alfaro2007; @Rocha2008; @Hernandez2009; @Dornburg2011; @Sorenson2013; @Santini2013; @Frederich2013; @Claverie2014; @Thacker2014], potentially due to the extensive amount of time necessary to collect data.
+To test whether our method can improve on the case where the data collection method is geometric morphometrics, we use the average time it took an expert to measure a single fish image and predict the time it would take for a single individual expert to measure all images at 5x replication, and compare it to the time it took turkers to collect these measurements at the same replication level.
+If the turkers in aggregate annotated images more quickly than a single expert would have, this suggests that the parallelization afforded by crowdsourcing is effective at reducing the total time required for data collection.
 
-To demonstrate the utility of obtaining comparative data using this method, we downloaded previously-published phylogenies for 7 fish families (Acanthurdiae, Apogonidae, Balistidae, Chaetodontidae, Labridae, Pomacentridae, and Tetraodontidae) from TreeBASE and Dryad (XXX: CITE TREES -- need to get info from Mike) and matched them to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using the EOL API [@Parr2014].
-If there were multiple images for a single species, we manually selected the best one.
-Crowdsourced workers collected body shape data following a standard protocol (Supplementary Material).
-We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] as implemented in the R package *geomorph* `r citep(citation("geomorph"))`.
+To demonstrate the utility of obtaining comparative data using this method, we use previously published phylogenies for seven fish families: Acanthuridae [@Sorenson2013], Balistoidae, Tetraodontidae [@Santini2013], Apogonidae, Chaetodontidae, Labridae [@Cowman2011; @Choat2012], and Pomacentridae [@Frederich2013].
+We matched `r nrow(distinct(fish_families, tip))` species to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [@Parr2014].
+Crowdsourced workers placed landmarks describing body shape variation following a standard protocol (Supplementary Material).
+The Cartesian position of these landmarks were used in a generalized Procrustes analyses [@Gower1975; @Rohlf1990], which centers, scales, and rotates landmark configurations to minimize the least-squares distance between shapes.
+We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] with the R package *geomorph* `r citep(citation("geomorph"))`, and used these principal components axes for subsequent analyses.
 
-To test whether our method can improve on the case where the data collection method is geometric morphometrics, we use the average time it took an expert to measure a single fish image and predict the time it would take for a single individual expert to measure our 7 fish families with 5 replicates, and compare it to the time it actually took turkers to collect these measurements at the same replication level.
-If the turkers in aggregate annotated images more quickly than a single expert would have, this suggests that the parallelization afforded by crowdsourcing is effective at reducing the time required for data collection.
-
-We used Bayesian Analysis of Macroevolutionary Mixtures [*BAMM*, @Rabosky2014a] to estimate the diversification rate as well as the rate of morphological evolution, retaining the PC axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998], for each of the 7 families.
-BAMM estimates the location of rate shifts in either diversification or character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method, where the number of rate regimes on a phylogeny is allowed to vary.
+We used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of speciation and body shape evolution for all seven families.
+For the characters describing body shape, we use the PC axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998].
+BAMM estimates the location of rate shifts in either diversification or character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of lineage diversification and trait evolution.
 We assessed convergence and mixing using Tracer [@rambaut2007tracer].
-We then used BAMMtools `r citep(citation("BAMMtools"))` to detect whether body shape affects change in the speciation or extinction rate.
-State-dependent diversification has often been invoked  to explain the high species richness enjoyed by certain lineages [@Mitter1988; @Barraclough1998]; this pipeline demonstrates a largely-automated way to detect these.
-`r citep(bibentry(bibtype="Article", title="A robust semi-parametric test for detecting trait-dependent diversification", author=c("Daniel L.
-Rabosky", "Huateng Huang"), year="in.review", journal="Systematic Biology"))`
+We also repeated each analysis and simulated under the prior (without data) to exclude rate heterogeneity that occurred solely due to stochastic processes.
+We use a Bayes Factor criterion of $BF > 5$ to enumerate the set of credible shifts [@Shi2015] and visualized them in R using BAMMtools `r citep(citation("BAMMtools"))`.
 
@@
 
-```{r turker_v_expert, fig.cap="Test figure caption", anchor="figure"}
+```{r geomorph_repeatability, warning=FALSE}
+only_good_points <- fish_reliability %>% filter(role != "student", !mark %in% c("A3", "A4", "D3", "D4", "O1", "O2", "O3","O4","O5", "P3", "P4"))
+reliability <- list()
+
+res <- only_good_points %>% gpa_with_vectors(family, role, sha1mac, sequence)
+vec <- res$vectors %>% group_by(family, sha1mac) %>% mutate(sequence = row_number(sequence))
+full_model <- procD.lm(res$coords ~ vec$family + vec$family:vec$sha1mac:vec$sequence, iter = 1)
+reliability$all <- ((full_model$MS[1] - full_model$MS[2])/5) / (full_model$MS[2] + ((full_model$MS[1] - full_model$MS[2])/5))
+
+res <- only_good_points %>% filter(role == "turker") %>% gpa_with_vectors(family, role, sha1mac, sequence)
+vec <- res$vectors %>% group_by(family, sha1mac) %>% mutate(sequence = row_number(sequence))
+full_model <- procD.lm(res$coords ~ vec$family + vec$family:vec$sha1mac:vec$sequence, iter = 1)
+reliability$turker <- ((full_model$MS[1] - full_model$MS[2])/5) / (full_model$MS[2] + ((full_model$MS[1] - full_model$MS[2])/5))
+
+res <- only_good_points %>% filter(role == "morphologist") %>% gpa_with_vectors(family, role, sha1mac, sequence)
+vec <- res$vectors %>% group_by(family, sha1mac) %>% mutate(sequence = row_number(sequence))
+full_model <- procD.lm(res$coords ~ vec$family + vec$family:vec$sha1mac:vec$sequence, iter = 1)
+reliability$expert <- ((full_model$MS[1] - full_model$MS[2])/5) / (full_model$MS[2] + ((full_model$MS[1] - full_model$MS[2])/5))
+
+res <- fish_reliability %>% filter(role == "morphologist") %>% gpa_with_vectors(family, role, sha1mac, sequence)
+vec <- res$vectors %>% group_by(family, sha1mac) %>% mutate(sequence = row_number(sequence))
+full_model <- procD.lm(res$coords ~ vec$family + vec$family:vec$sha1mac:vec$sequence, iter = 1)
+reliability$expert_all <- ((full_model$MS[1] - full_model$MS[2])/5) / (full_model$MS[2] + ((full_model$MS[1] - full_model$MS[2])/5))
+
+res <- fish_reliability %>% filter(role == "turker") %>% gpa_with_vectors(family, role, sha1mac, sequence)
+vec <- res$vectors %>% group_by(family, sha1mac) %>% mutate(sequence = row_number(sequence))
+full_model <- procD.lm(res$coords ~ vec$family + vec$family:vec$sha1mac:vec$sequence, iter = 1)
+reliability$turker_all <- ((full_model$MS[1] - full_model$MS[2])/5) / (full_model$MS[2] + ((full_model$MS[1] - full_model$MS[2])/5))
+rm(vec, res, full_model)
+```
+
+```{r turker_v_expert, fig.cap="Per-family breakdown of accuracy vs.
+precision for each landmark.
+Accuracy is represented as the difference between the median turker location for that landmark and the median expert location, with the expert location assumed to be the true location.
+Precision is represented as the log-ratio of median absolute deviations between turkers and experts.
+More positive numbers indicate better expert precision, whereas more negative numbers indicate better turker precision.
+Points highlighted in red are those determined to be outliers (1.5 $\times$ IQR).
+See Supplemental Information for a labeled version of this figure.", anchor="figure", message=FALSE, warning=FALSE}
 no_st <- fish_reliability %>% filter(role != "student") %>% mutate(role = factor(role))
@@
 
-all_stats <- left_join(accuracy, precision) %>% left_join(full_family) %>% mutate(outlier_label = ifelse(outlier, as.character(mark), NA))
+all_stats <- left_join(accuracy, precision) %>% mutate(outlier_label = ifelse(outlier, as.character(mark), NA))
 
-ggplot(all_stats, aes(median, mad, color=outlier)) + geom_point() + labs(x="Accuracy", y="Precision: log2 turker variance / expert variance ") + facet_wrap(~full_family)# + geom_dl(aes(label=outlier_label), size=1, list("top.bumptwice", dl.trans(y=y+0.1), cex=0.75)) + scale_color_manual(values=c("black", "red")) + ylim(-max(abs(all_stats$mad) + 0.25), max(abs(all_stats$mad) + 0.25)) + theme("legend.position" = "none")
+ggplot(all_stats, aes(median, mad, color=outlier)) + geom_point() + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance / expert variance ") + facet_wrap(~family) + scale_color_manual(values = c("black", "red")) + theme_bw() + theme(legend.position="none") + xlim(0, 230)
+ggsave("turker_precision.pdf", width=14, height=14, scale = 1/3)
+
+# ggplot(highlighted, aes(median, mad)) + geom_point(color = "white") + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
+expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
+# ggsave("turker_precision_black.pdf", width=14, height=14, scale = 1/3)
+# 
+# 
+# highlighted <- all_stats %>% mutate(should_color = family == "Tetraodontidae" & str_detect(mark, "O"))
+# 
+# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad + 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
+expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
+# 
+# ggsave("turker_precision_black_puffers.pdf", width=14, height=14, scale = 1/3)
+# 
+# highlighted <- all_stats %>% mutate(should_color = family == "Chaetodontidae" & str_detect(mark, "[AD]3"))
+# 
+# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad - 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
+expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
+# 
+# ggsave("turker_precision_black_chaets.pdf", width=14, height=14, scale = 1/3)
+# 
+# highlighted <- all_stats %>% mutate(should_color = family == "Balistidae" & str_detect(mark, "D[12]"))
+# 
+# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad - 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
+expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
+# 
+# ggsave("turker_precision_black_triggers.pdf", width=14, height=14, scale = 1/3)
+# 
+# 
+# highlighted <- all_stats %>% mutate(should_color = median > 10)
+# 
+# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad - 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
+expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
+
+# + geom_dl(aes(label=outlier_label), size=1, list("top.bumptwice", dl.trans(y=y+0.1), cex=0.75)) + scale_color_manual(values=c("black", "red")) + ylim(-max(abs(all_stats$mad) + 0.25), max(abs(all_stats$mad) + 0.25)) + theme("legend.position" = "none")
 ```
 
-```{r geomorph_lm, eval=FALSE}
-fmted <- format_for_gpagen(no_st)
-allsplit <- str_split_fixed(dimnames(fmted)$id, "_", 3)
-family_vec <- allsplit[, 1]
-role_vec <- allsplit[, 2]
-ind_vec <- allsplit[, 3]
-gpaed <- gpagen(fmted, ShowPlot=F)
+For nearly all landmarks, turkers only differ from the expert consensus by a few tens of pixels (Figure `r figr("turker_v_expert")`, Supplemental Figure S3).
+The most accurate and precise points are those that are related to the position of the eye (landmarks E1 and E2).
+The least accurate are those in the opercular series (O1-O5), particularly the ones related to the preopercle (O1-O3) likely because in certain groups (e.g., Tetraodontidae) the preopercle is difficult to visualize from external morphology alone.
+Experts were generally more precise than turkers, however there were some landmarks where the turkers converged on very similar locations.
+Based on these results we exclude in subsequent analyses the landmarks relating to the distal margins of all fins (A3, A4, P3, P4, D3, D4), the preopercle bones (O1-O3), the dorsal fin for triggerfishes (D1, D2), and the opercular opening for pufferfishes (O4-O5), due to low turker accuracy.
 
-role_by_ind <- procD.lm(gpaed$coords ~ family_vec * role_vec * ind_vec, iter=999, RRPP=T)
-ind_only <- procD.lm(gpaed$coords ~ family_vec:ind_vec, iter=999, RRPP=T)
+The inter-observer reliability of turkers and experts as measured by the ratio of among-individual and sum of the among-individual and measurement error ANOVA components was `r percent(reliability$turker)` and `r percent(reliability$expert)`, respectively.
+Although there is no current standard for acceptable levels of measurement reliability [@VonCramon-Taubadel2007], these percentages are not low enough to suggest pathologies in the measurement protocol.
 
-advanced.procD.lm(gpaed$coords ~ ind_vec,  ~ role_vec * ind_vec, ~ family_vec, iter=499)
+```{r geomorph_lm, warning=FALSE}
+res <- gpa_with_vectors(only_good_points, family, role, sha1mac)
+
+family <- res$vectors$family
+role <- res$vectors$role
+
+lm_result <- advanced.procD.lm(res$coords ~ family * role, ~ family)
+lm_p <- lm_result$P[2]
+lm_Z <- lm_result$Z[2]
+lm_F <- lm_result$F[2]
+rm(family, role)
 ```
 
-```{r median_tbl, eval=F, results="asis", anchor="table", fig.cap="Median table"}
-med_wide <- meds %>% spread(key=family, value=value)
-med_wide %>% knitr::kable(digits = 2, caption = "Median table")
-```
-
-```{r mad_tbl, eval=F, results="asis", anchor="table", fig.cap="MAD table"}
-mad_wide <- mads %>% spread(key=family, value=value)
-mad_wide %>% knitr::kable(digits = 2, caption="MAD table")
-```
-
-We compare the difference in accuracy and precision between turkers and experts in Figure `r figr("turker_v_expert")`.
-
-The most accurate and precise points are those that are related to the position of the eye (landmarks E1 and E2).
-The least accurate are those in the opercular series (O1-O5), particularly the ones related to the preopercle (O1-O3) as well as in species where the preopercle is difficult to visualize from external morphology alone (e.g., Tetraodontidae).
-We also plot the median difference and the
-
 ```{r cv_table, results="asis", anchor="table", message=FALSE}
-
-subsamp <- group_by(no_st, family, role) %>% sample_n(size=400) %>% ungroup()
+subsamp <- fish_reliability %>% filter(role !="student") %>% group_by(family, role) %>% sample_n(size=400) %>% ungroup() %>% mutate(role = factor(role))
 
@@
 
-CV_sum <- CV %>% summarise(family=family, error=estimator[["error"]]) %>% separate("family", "family", sep="_", extra="drop")
+CV_sum <- CV %>% summarise(family=family, error=estimator[["error"]])
 
@@
 
-CV_sum2 <- CV %>% summarise(family=family, qda_error=estimator[["error"]]) %>% separate("family", "family", sep="_", extra="drop")
-
-CV_sum <- left_join(CV_sum, CV_sum2) %>% left_join(full_family, by="family")
+CV_sum2 <- CV %>% summarise(family=family, qda_error=estimator[["error"]])
 
 # format for display
-CV_sum %>% transmute(`Family`=full_family, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.")
+CV_sum %>% left_join(CV_sum2) %>% transmute(`Family`=family, `LDA`=error, `QDA`=qda_error) %>% knitr::kable(digits=3, caption="Misprediction rate of linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) with 10-fold cross validation for each fish image.
+The discriminant model for each family was unable to meet the standard of one in four misclassifications, and in some cases, the more flexible QDA method performed worse than the LDA model.")
 ```
 
-Both linear and quadratic discriminant analysis with 10-fold cross validation (Table `r figr("cv_table")`) was unable to reliably distinguish between morphologists and crowdsourced workers using our acceptable misprediction threshold of 20%.
+The non-parametric MANOVA with RRPP failed to detect a significant difference between turker and expert shapes ($p = `r lm_p`$, $Z = `r lm_Z`$, $F=`r lm_F`$).
+Similarly, both linear and quadratic distriminant analysis with 10-fold cross validation (Table `r figr("cv_table")`) were unable to reliably distinguish between these two groups, for any given family.
+Although for some images the classifier showed slight improvement beyond a 50% coin flip, in all cases our model fell short based on a one in four (25%) acceptable misclassification rate.
+We conclude that, for any given sample of landmarks, it is challenging to statistically distinguish between expert-provided and turker-provided landmark configurations.
 
-```{r turk_morph_procrustes_space, fig.cap="Turker morpho space", anchor="figure"}
+```{r turk_morph_procrustes_space, fig.cap="Morphospace projection for each observer's mean shape.
+Blue points indicate experts, while red points indicate turkers.
+The mean shape for all turkers and experts for a given family is the point outlined in black for each family, and connected with a black line to help emphasize the difference between turker and expert mean shapes.
+The convex hull for each family is drawn to show the amount of among-observer shape variation.", anchor="figure"}
 # Load data from the anonymized results file, remove students, drop bad marks
@@
 	filter(role != "student") %>% 
-	separate(col=mark, into=c("cat", "cat_seq"), sep=1, remove=F, convert=T) %>%
-	filter(!(family %in% c("Bal") & mark %in% c("D1","D2")), !mark %in% c("O1","O2","O3","O4","O5", "A3", "D3", "D4"))
+	filter(!mark %in% c("O1","O2","O3", "A3", "A4", "D3", "D4", "P3", "P4"))
 
@@
 df <- df %>% group_by(family, mark, role, sha1mac) %>% summarise(x=median(x), y=median(y))
+#df <- df %>% group_by(family, mark, role, sha1mac, sequence)
 
-coords <- gpagen(format_for_gpagen(df), ShowPlot=F)$coords
+coords <- gpagen(format_for_gpagen(df, family, role, sha1mac, sequence), ShowPlot=F)$coords
 
-plot_tangent_space(pca_df(coords), "PC1", "PC2")
+#plot_tangent_space(pca_df(coords), "PC1", "PC2")
+#plot_tangent_space(pca_df(coords), "PC3", "PC4")
+
+# exclude D1 & D2 for triggers, O4 and O5 for puffers
+
+df <- df %>% filter(!(family == "Balistidae" & mark %in% c("D1", "D2")), !(family == "Tetraodontidae" & mark %in% c("O4", "O5")))
+coords <- gpagen(format_for_gpagen(df, family, role, sha1mac, sequence), ShowPlot=F)$coords
+
+turk_morph_pca <- pca_df(coords) %>% separate(id, c("family", "role", "ind", "seq"), sep = "_")
+
+write_csv(turk_morph_pca, "turk_morph_pca_for_supplement.csv")
+
+plot_tangent_space(turk_morph_pca, "PC1", "PC2")
 ```
 
-Post text
-
-```{r brokenstick}
+```{r brokenstick_reliability}
 # use broken stick method to get the principal components stopping point
@@
 
-We analyzed PC 1 through PC `r stopping_pc`, based on the broken-stick method of determining a PCA stopping point.
-Projecting consensus shapes into Procrustes space (Figure `r figr("turk_morph_procrustes_space")`) shows that within-group variation exceeds that of among group variation.
-The consensus turker shape and consensus expert shape are nearly identical for 5 of 7 families on PC1 and PC6 (Figure `r figr("turk_morph_procrustes_space")`, S1 and S2).
+We projected turker and expert shape configurations into morphospace (Figure `r figr("turk_morph_procrustes_space")`, Supplemental Figure S4) Although the overall space occupied by each family's shape configurations vary, in practice, the aggregated median turker and expert shapes are not qualitatively different.
+The only exception is the triggerfishes (Balistidae), likely due to turker confusion over the exact location of dorsal fin due to their reduced anterior dorsal fin.
 
@@
 
-XXX: Include BAMM speciation / morphological evolution analysis.
+```{r timing_plot, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
+Landmarks were first received eight minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark.", anchor="figure"}
+# median time (in minutes) it takes an expert to measure these fishes
+timings <- (no_st %>% group_by(role) %>% summarise(duration=median(duration))) %>% mutate(total_time=duration * nrow(no_st))
 
-```{r sevenfamily_tangent_plot, message=F, warning=F, fig.show="hold", fig.cap="Family figures", anchor="figure"}
+morpho_per_image <- timings[1, 2] %>% as.numeric() %>% dminutes() %>% round(1)
+
+# data collection time plot
+timing_data <- fish_families %>% ungroup() %>% group_by(tip) %>% summarise(time=min(time_taken)) %>% arrange(time) %>% mutate(frac=1:n()/n())
+
+ggplot(timing_data, aes(time, frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
+
+total_turker_time <- max(fish_families$time_taken) %>% as.numeric %>% dminutes %>% round(1)
+```
+
+Using a median expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(distinct(fish_families, tip)))` to landmark all `r nrow(distinct(fish_families, tip))` images.
+At 5x replication, this would take `r as.character(morpho_per_image * nrow(fish_families))`.
+By comparison, turkers took a total of `r as.character(total_turker_time)` to complete all images at 5x replication.
+
+```{r sevenfamily_tangent_plot, message=F, warning=F, fig.show="hold", fig.cap="Morphospace for seven families of ray-finned fishes.
+Each point indicates a separate species; families are separated by colors.
+The convex hull for each family is drawn to show area of morphospace occupied by each family.
+Figures for other PC axes are present in the Supplemental Material.", anchor="figure"}
+name_map <- data_frame(family = c("acanthurid", "apogonidae", "balistoidae", "chaetodontidae", "labridae", "pomacentridae", "tetraodontidae"),
+                       pretty_family = c("Acanthuridae", "Apogonidae", "Balistoidae", "Chaetodontidae", "Labridae", "Pomacentridae", "Tetraodontidae")
+)
+
 df <- fish_families %>% group_by(tip, mark, family) %>% summarise(x=median(x), y=median(y))
@@
 
+pcaed <- left_join(pcaed, name_map, by="family") %>% mutate(family = pretty_family)
+write_csv(pcaed, "seven_family_pca_for_supplement.csv")
+
 plot_tangent_space2(pcaed, labelmethod=list("chull.grid"))
@@
 
-asdfsdf
+```{r brokenstick_families}
+# use broken stick method to get the principal components stopping point
 
-```{r timing_plot, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
-Landmarks were first received 8 minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark.", anchor="figure"}
-# median time (in minutes) it takes an expert to measure these fishes
-timings <- (no_st %>% group_by(role) %>% summarise(duration=median(duration))) %>% mutate(total_time=duration * nrow(no_st))
+pc_res <- prcomp(two.d.array(full_coords))
+eigen_res <- pc_res$sdev^2
+bstick_res <- bstick(pc_res)
 
-morpho_per_image <- timings[1, 2] %>% as.numeric() %>% dminutes() %>% round(1)
-
-# data collection time plot
-timing_data <- fish_families %>% ungroup() %>% group_by(tip) %>% summarise(time=min(time_taken)) %>% arrange(time) %>% mutate(frac=1:n()/n())
-
-ggplot(timing_data, aes(time, frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
-
-total_turker_time <- max(fish_families$time_taken) %>% as.numeric %>% dminutes %>% round(1)
+stopping_pc <- which(eigen_res < bstick_res)[1] %>% as.numeric
 ```
 
-Using a median-expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(no_st))` to landmark all `r nrow(no_st)` images.
-At 5x replication, this would take `r as.character(morpho_per_image * nrow(no_st) * 5)`.
-By comparison, turkers took a total of `r as.character(total_turker_time)` to complete all images at 5x replication.
+```{r load_bamm_data, message=FALSE, echo=FALSE, warning=FALSE, results="hide"}
+base <- "data-raw/bamm/"
+base <- "../data-raw/bamm/"
+fams <- c("acanthurid", "apogonidae", "balistoidae", "chaetodontidae", "labridae", "pomacentridae", "tetraodontidae")
+pretty_names <- c("Acanthuridae", "Apogonidae", "Balistoidae", "Chaetodontidae", "Labridae", "Pomacentridae", "Tetraodontidae")
+names(fams) <- pretty_names
 
-The BAMMtools analysis identified areas of each phylogeny that were experiencing faster rates of body shape evolution. XXX: Talk about these in more detail.
+tre_trait <- lapply(paste0(base, "input_trees/", fams, ".tre"), read.tree)
+tre_speciation <- lapply(paste0(base, fams, ".tre"), read.tree)
+names(tre_trait) <- names(tre_speciation) <- fams
+
+cl <- makePSOCKcluster(detectCores(), useXDR=F)
+clusterExport(cl, list("base", "fams", "tre_speciation", "tre_trait"))
+evt_speciation <- parLapplyLB(cl, fams, function(fam) {
+  tre <- tre_speciation[[fam]]
+  BAMMtools::getEventData(tre, paste0(base, "div_runs/", fam, "_event_data.txt"), burnin=0.1)
+})
+evt_trait <- parLapplyLB(cl, fams, function(fam) {
+  tre <- tre_trait[[fam]]
+  BAMMtools::getEventData(tre, paste0(base, "trait_runs/", fam, "_PC1_event_data.txt"), burnin=0.1, type="trait")
+})
+names(evt_trait) <- names(evt_speciation) <- fams
+clusterExport(cl, list("evt_speciation", "evt_trait"))
+rtt_speciation <- parLapplyLB(cl, fams, function(fam) {
+  BAMMtools::getRateThroughTimeMatrix(evt_speciation[[fam]])
+})
+rtt_trait <- parLapplyLB(cl, fams, function(fam) {
+  BAMMtools::getRateThroughTimeMatrix(evt_trait[[fam]])
+})
+prior_speciation <- parLapplyLB(cl, fams, function(fam) {
+  BAMMtools::getBranchShiftPriors(tre_speciation[[fam]], paste0(base, "div_runs/", fam, "_prior_probs.txt"))
+})
+prior_trait <- parLapplyLB(cl, fams, function(fam) {
+  BAMMtools::getBranchShiftPriors(tre_trait[[fam]], paste0(base, "trait_runs/", fam, "_PC1_prior_probs.txt"))
+})
+
+names(pretty_names) <- names(prior_trait) <- names(prior_speciation) <- names(rtt_trait) <- names(rtt_speciation) <- fams
+stopCluster(cl)
+```
+
+```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 (a, b) and speciation (c, d) across seven families of fishes.
+Phylorate plots (a, c) color branch lengths by rates of shape evolution (a) and speciation (c), where warmer colors indicate faster rates of evolution.
+Significant rate shift events (pp > 0.95) are indicated on the phylorate plot as a red circle on the corresponding branch.
+Median log rates of shape evolution (b) and speciation (d) through time, where black lines indicate the background rate and red lines indicate the rate of evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a) or (c).", fig.height=9.5, out.height="0.9\\textheight", out.width="\\textwidth"}
+
+layout_mat <- matrix(c(
+  0,  1,  2,  3,  4,
+  5,  12, 13, 14, 15,
+  6,  16, 17, 18, 19,
+  7,  20, 21, 22, 23,
+  8,  24, 25, 26, 27,
+  9,  28, 29, 30, 31,
+  10, 32, 33, 34, 35,
+  11, 36, 37, 38, 39
+), ncol = 5, byrow = TRUE)
+
+htext <- function(str) {
+  plot.new()
+  text(0.5, 0.5, str, cex=1)
+}
+vtext <- function(str) {
+  plot.new()
+  text(0.5, 0.5, str, cex=1, srt=90)
+}
+
+customintervals <- list(
+  pomacentridae = c(0.000, 0.0025)
+)
+
+mantissa <- function(x) {
+  sapply(x, function(x) {
+    if (x == 0) 0
+    else {
+      log <- log10(abs(x))
+      10^(log - floor(log))
+    }
+  })
+}
+
+exponent <- function(x) {
+  sapply(x, function(x) {
+    if (x == 0) 0
+    else floor(log10(abs(x)))
+  })
+}
+
+plot_rtt <- function(rtt, shifts = list(), xlim=NULL, ylim=NULL) {
+  vartype <- if (rtt$type == "trait") "beta" else "lambda"
+  if (length(shifts) > 0) {
+    allshifts <- c(list(rtt), shifts)
+  } else {
+    allshifts <- list(rtt)
+  }
+  allshifts <- lapply(allshifts, function (x) {
+    x$avg <- apply(x[[vartype]], 2, median)
+    x$times <- rev(x$times)
+    x
+  })
+  xlim <- if (is.null(xlim)) range(sapply(allshifts, `[`, "times")) else xlim
+  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[`, "avg")) else ylim
+  old.par <- par(mar = c(2,3,2,1))
+  plot.new()
+  plot.window(rev(xlim), ylim, log = "y")
+  axis(at = axTicks(1), side = 1, labels = axTicks(1))
+  axis(side = 2, las = 2)
+  lines(allshifts[[1]]$times, allshifts[[1]]$avg)
+  if (length(shifts) > 0) {
+    lapply(2:length(allshifts), function (x) {
+      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red")
+    })
+  }
+  par(old.par)
+  return(invisible(NULL))
+}
+
+mar <- c(2,1,2,1)
+par(mar = c(0,0,0,0))
+layout(layout_mat, widths = c(1, 6, 6, 6, 6), heights = c(0.9, 9, 9, 9, 9, 9, 9, 9))
+htext("Shape Phylorates")
+text(0, 0.5, "a)", cex=1.2)
+htext("Shape Rate Through Time")
+text(0, 0.5, "b)", cex=1.2)
+htext("Speciation Phylorates")
+text(0, 0.5, "c)", cex=1.2)
+htext("Speciation Rate Through Time")
+text(0, 0.5, "d)", cex=1.2)
+lapply(pretty_names, vtext)
+par(mar = mar)
+lapply(fams, function(fam) {
+  interval <- if(is.null(customintervals[[fam]])) c(0, 0.001) else customintervals[[fam]] 
+  q <- plot.bammdata(evt_trait[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = interval)
+  best <- getBestShiftConfiguration(evt_trait[[fam]], prior = prior_trait[[fam]])
+  addBAMMshifts(best, cex = 2, par.reset=F)
+  axisPhylo()
+
+  shift_node <- getShiftNodesFromIndex(best, 1)
+  if (length(shift_node) > 0) {
+    background_rates <- getRateThroughTimeMatrix(evt_trait[[fam]], node = shift_node, nodetype = "exclude")
+    clade_rates <- lapply(shift_node, function (x) {
+      phy <- extract.clade(tre_trait[[fam]], x)
+      write.tree(phy, file = paste0(fam, "_trait_shift_", x, ".tre"))
+      getRateThroughTimeMatrix(evt_trait[[fam]], node = x)
+      })
+    plot_rtt(background_rates, clade_rates)
+  } else {
+    plot_rtt(rtt_trait[[fam]])
+  }
+
+  q <- plot.bammdata(evt_speciation[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = )
+  best <- getBestShiftConfiguration(evt_speciation[[fam]], prior = prior_speciation[[fam]])
+  addBAMMshifts(best, cex = 2, par.reset=F)
+  axisPhylo()
+
+  shift_node <- getShiftNodesFromIndex(best, 1)
+  if (length(shift_node) > 0) {
+    background_rates <- getRateThroughTimeMatrix(evt_speciation[[fam]], node = shift_node, nodetype = "exclude")
+    clade_rates <- lapply(shift_node, function (x) {
+      phy <- extract.clade(tre_speciation[[fam]], x)
+      write.tree(phy, file = paste0(fam, "_speciation_shift_", x, ".tre"))
+      getRateThroughTimeMatrix(evt_speciation[[fam]], node = x)
+      })
+    plot_rtt(background_rates, clade_rates)
+  } else {
+    plot_rtt(rtt_speciation[[fam]])
+  }
+})
+
+```
+
+Using the broken-stick method of determining a PCA stopping point, we analyzed PC 1 through PC `r stopping_pc`.
+We project per-species consensus shapes into Procrustes space (Figure `r figr("sevenfamily_tangent_plot")`, Supplemental Figure S5).
+The BAMMtools analysis uncovered substantial amounts of heterogeneity in the rate of body shape evolution and speciation in each family (Figure `r figr("bamm_plots")`).
+Significant shifts in the rate of shape evolution or speciation were detected in three families: Labridae, Apogonidae, and Pomacentridae.
+The significant shifts in speciation rate corroborate those found in @Cowman2011 through either MEDUSA [@Alfaro2009] or a relative cladogenesis statistic [@Nee1992].
+Two significant shifts in shape evolution rate occur in the wrasses (Labridae).
+The first rate shift occurs deep in the tree, corresponding to the lineage containing the labrine, scarine, and cheiline tribes.
+The other shift is nested within that group, in *Sparisoma*.
+One shift in speciation rate also occurs in the wrasses, encompassing the genera *Chlorurus* and *Scarus*.
+One shift in speciation rate occurs in the cardinalfishes (Apogonidae), encompassing members of the genera *Apogon*, *Archamia*, *Zoramia*, *Ostorhinchus*, *Cheilodpterus*, *Gossamia*, *Fowleria*, and *Phaeoptyx* [Apogonini + Apogonichthynini *sensu* @Mabuchi2014].
+One shift in the rate of shape evolution occurs in the damelfishes (Pomacentridae) in the genus *Amphiprion*.
 
@@
 
-XXX: Discuss what people currently do to estimate error in geometric morphometrics. Make the point that this method is really good at getting at an estimate of error.
+<!--
+Distal margin of the fin: would semilandmarks work better?
+-->
 
+We have shown that crowdsourcing through Amazon Mechanical Turk is a tractable approach for generating reliable trait data at an unprecedented scale.
+Using this framework, it is possible to distribute thousands of images to workers, collect the data, and send it to a comparative analysis pipeline.
+We have also demonstrated that it is possible to identify the set of geometric morphometric landmarks that can be reliably captured by nonspecialists.
 We found that for certain landmarks there was significant between and within group disagreement.
-Based on median average deviation, points belonging to the opercular series and those that concerned the distal margin of the dorsal and anal fins were particularly challenging for turkers to digitize, compared to the experts.
-Typically, geometric morphometric studies only use a single digitizer in efforts to reduce interdigitizer error.
-We argue that this common technique is flawed, as it papers over potential differences in how observers measure the same homologous landmark on the same organism.
-Instead, explicitly accounting for interobsever error is critical to determine the efficacy of each individual landmark and the replicability of the study as a whole.
-Examining interobserver error is a critical process that signals which landmarks can be relied on and which can be discarded, as we have done in this analysis.
-Despite the reduction in the amounts of landmark data collected, even a simplified protocol can produce meaningful biological results that are concordant with previous hypotheses in these groups.
+Based on median average deviation, points belonging to the opercular series and those that locating the distal margin of the dorsal and anal fins were particularly challenging, compared to the experts.
+Based on these results, nonspecialist turkers are unlikely to replace experts for all morphometric tasks.
+However, by digitizing less than 5% of our dataset with experts, we were able to identify groups of landmarks that exhibited extremely poor performance and excluded these.
+Furthermore, we were able to obtain biologically significant results from a dataset collected entirely by turkers.
+Through combining expert knowledge with the sheer scale of the Amazon Mechanical Turk workforce, it is possible to collect and assess large quantities of morphometric data, with an order of magnitude improvement in throughput over traditional approaches.
 
-Previous work in generating large phenotypic datasets typically relied on one or more of expertise, effort, and computational power.
-Collecting phenotypic data remains a large bottleneck in our ability to comprehensively test hypotheses about large radiations of organisms.
-We developed a method to gather phenotypic data using Amazon Mechanical Turk.
-Our results show that the quality of turkers' data are not significantly different from results collected from experts, and that these data can be gathered extremely quickly.
+## Reliability
 
-We attempted to distinguish between turker-generated landmarks and expert-generated landmarks using linear discriminant analysis.
-Even the image with the lowest misprediction rate, from the family `r filter(CV_sum, min(error) == error)$family`, still incorrectly predicted the digitizer's identity `r filter(CV_sum, min(error) == error)$error %>% as.numeric() %>% scales::percent()` of the time, suggesting that, given a set of landmarks, it is difficult to predict whether a turker or a morphologist created them.
-The quadratic discriminant analysis did not reliably improve prediction of worker identity over the LDA.
+One advantage of the crowdsourced method we develop here is that inter-observer error can be readily assessed.
+Traditional geometric morphometric studies often rely on a single observer for practical reasons (the pool of trained geometric morphometricians is limited), and to avoid individually-driven systematic biases in data collection.
+Although this common practice may reduce bias, it also precludes meaningful assessment of differences among observers.
+Our results show that inter-observer variance can be substantial for some landmarks even among expert digitizers.
+Therefore, explicitly accounting for inter-observer error is critical to determine the efficacy of each individual landmark and the replicability of the study as a whole.
+Inter-observer error signals which landmarks can be relied on and which merit further consideration, as we have done in this analysis.
+The quantification of inter-observer error is a strict requirement of our workflow, as it would otherwise be impossible to arrive at a single consensus shape across several turkers working independently.
+This requirement ensures that inter-observer error is not ignored or bypassed due to the difficulty of assessing it.
+
+In our analysis, we assessed the quality of a variety of landmarks between turkers and experts.
+Unsurprisingly, turkers performed exceptionally poorly for several landmarks requiring knowledge of fish anatomy.
+For example, the landmarks that describe the shape of the fish's caudal fin asked workers to mark the distal tip of the first principal fin ray.
+Even when turkers are armed with a definition and a comparison between procurrent and principal fin rays, the experts' experience and training allow them to substantially outperform turkers in identifying this point.
+Furthermore, experts generally had lower disagreement in their landmark placement when compared to turkers, even for landmarks that turkers found especially difficult.
+These differences between experts and MTurk workers have also been observed in image categorization tasks [@Deng2009; @VanHorn2015].
+However, it is possible that an improved training protocol could result in better collection of these difficult landmarks.
+Turkers have been found to perform well in extremely detailed video annotation tasks [@Vondrick2013], provided that researchers conduct pre-task training and post-task validation.
+Implementing these pre-task requirements would be a straightforward avenue to improve accuracy for future work.
+
+## The role of crowdsourced phenotypic data collection in modern comparative studies
 
@@
 For example, @Brusatte2014 used a 853 character discrete character matrix for 150 taxa to estimate the rate of morphological evolution in the transition from theropod dinosaurs to modern birds.
-These data were collected over the course of 20 years as part of the Therapod Working Group [@Norell2001; @Clark2002; @Hwang2002; @Hwang2004; @Makovicky2005; @Norell2006; @Turner2007; @Turner2012; @dryad_84t75].
+These data were collected over the course of 20 years as part of the Therapod Working Group [@dryad_84t75].
 @OLeary2013 combined the work of MorphoBank contributors [@OLeary2011] with literature review to generate 4,541 characters for 86 species.
-@Rabosky2013 examined 7,822 species of ray-finned fish but used a single quantitative measure (body size) to evaluate correlated speciation and morphological evolution.
-Newer approaches, typically involving statistical techniques such as machine vision or natural language processing, still require significant researcher expertise in either gathering large training datasets, or to develop new and more sophisticated machine learning models [@Brill2003; @Halevy2009].
+@Rabosky2013 examined 7,822 species of ray-finned fish and used a single quantitative measure (body size) collected from FishBase [@Froese2014], whose data are contributed from the scientific literature by experts.
+All of these studies share the same requirement for intensive researcher effort, but the data collected is generally either broad (many species) or deep (many characters).
+In this study, we collected a phenotypically rich dataset across great taxonomic breadth.
+This approach can easily be scaled to permit unprecedented, massive comparative analyses on new, phenotypically rich datasets.
+
+This method does not threaten to replace experienced morphologists.
+Though certain conspicuous landmarks can be rapidly collected by turkers, other types of analyses will require landmarks that can only be identified by experts and thus cannot use the high-throughput method presented here.
+Although this can likely be alleviated by implementing more sophisticated training regimes, the implicit anatomical knowledge that morphologists have must be made explicit in the form of a written protocol for turkers to follow.
+The cost of developing a clearer and simpler protocol that still captures the essence of the morphological characters of interest must be weighed against the benefit of higher-throughput from turker data collection, and for many such analyses this tradeoff is impractical.
+However, for such analyses where crowdsourcing is a viable alternative, our approach allows experts to move beyond data collection and into a role of developing training materials for nonspecialists and validating the data collected by crowdsourced workers.
+
+Approaches involving statistical techniques like machine vision and natural language processing have yet to make significant headway in automatically collecting morphological data.
+Although methods to automatically measure leaves exist [@Corney2012; @Corney2012a], these require 2D specimens to eliminate parallax error, as well as high-contrast mounting paper backgrounds for effective automatic outline detection.
+More sophisticated methods for lower-quality images or organisms with more 3D structure have yet to be developed.
+Natural language processing of the scientific literature could potentially be used for automatic extraction of morphological characters using DeepDive [@Peters2014; @Shin2015], but it may require impractically large corpus sizes [@Brill2003; @Halevy2009].
+Crowdsourcing can augment and enhance these statistical techniques.
+For example, the algorithm in @Corney2012 occasionally captures non-leaf objects and systematically underestimates leaf sizes.
+MTurk workers could improve this method by confirming the presence of a leaf in the image segment and measure the leaf size to ground truth the algorithm's results.
+
+A third alternative to using expert morphologists and crowdsourced workers to collect data is through citizen science.
+Citizen scientists are enthusiasts that volunteer to collect data or contribute annotations to a scientific endeavor.
+They can specialize in a particular field, such as birds, plants, or fungi.
+Compared to Amazon Mechanical Turk workers, citizen scientists are typically unpaid, but can produce higher quality work due to their expertise.
+For example, a study comparing citizen scientists and MTurk workers showed that for an image segmentation task MTurk workers had higher throughput and comparable accuracy to citizen scientists, but MTurk workers performed poorly when asked to identify birds to the species level [@VanHorn2015].
+
+## Suitability for other systems
 
@@
 An online methodology could conceivably improve on this analysis time, by iteratively refining its results as new data streamed in from Amazon's servers.
-(BAMM does not support adding new data mid-analysis).
-Other speedups could be found, particularly in identifying the best image for each species to use for analysis.
-This could easily be replaced with a protocol for image selection and uploaded to MTurk for categorization.
 
-The results of this pipeline also serve to test previous predictions of the rates of cladogenic speciation and mophological evolution [@Rabosky2013].
-Body shape and body size are usually correlated (XXX: find the Cite), and the first PC axis of body shape analyses is typically associated with features related to size.
-(XXX: make an extensive comparision between the results of the nat comm paper and the 7 families here)
+Although there are limitations in the type and accuracy of data that can be collected through MTurk crowdsourcing, even a simplified protocol can produce meaningful biological results that are concordant with previous hypotheses in these groups.
+We detected a significant shift in the rate of body shape evolution in Labridae, restricted to the wrasse tribes Labrini, Cheilini, and Scarini.
+The scarines and cheilines are mostly reef-associated [@Froese2014], which has been proposed as an environment that drives diversification rate changes in marine teleosts [@Alfaro2007; @Cowman2011; @Price2011].
+These results suggest that evolution of body form may also be influenced by environmental association [@Claverie2014].
+Although the example we present here was necessarily limited, extending this technique to generate new phenotypic datasets for existing large phylogenetic trees such as fishes [@Rabosky2013], birds [@Jetz2012], mammals [@Bininda-Emonds2007], and angiosperms [@Zanne2014] would be straightforward, especially for taxa where image data are already aggregated in a database such as FishBase [@Froese2014] or the Encyclopedia of Life [@Parr2014].
 
@@
 We use researcher expertise to identify a comparative hypothesis, and design a data collection protocol to specifically test this hypothesis.
-Users on Amazon Mechanical Turk supply a large source of worker effort that collects data per our protocol.
+Amazon Mechanical Turk supplies a large source of worker effort that collects data according to protocol.
 Finally, computational statistical techniques validate the accuracy of our data and identify outliers and other errors in data collection.
@@
 
-We stress that our work does not attempt to, and will likely never displace traditional morphological research and data collection.
-The role of these traditional techinques is critical for generating preliminary hypotheses.
-Our method instead performs high-throughput collection of phenotypic data to match the current speeds of molecular data collection.
-Collecting this data is not useful without having in mind a specific hypothesis to test.
-Our work fills the niche of gathering phenotypic data across large radiations, which is still a challenging open research question.
-Even seemingly obvious phenotypes, such as the woodiness of plant species, are incomplete and sampled in a biased manner, potentially misleading inference on a global scale [@FitzJohn2014].
-We suggest that our method can play a crucial role in increasing the amount of data that can be brought to bear against comparative hypotheses.
+Our work fills the niche of gathering phenotypic data across large radiations, which has been a challenging open research question [@Burleigh2013].
+Even seemingly obvious phenotypes, such as the woodiness of plant species, are incomplete and sampled in a biased manner [@FitzJohn2014], potentially misleading inference on a global scale.
+This method unlocks the potential of high-throughput data collection, and shifts the data bottleneck for morphological research onto acquiring suitable images for quantification, and developing higher-quality worker training regimens to enable collection of more sophisticated data.
+The burden is now on experienced taxonomists and morphologists to create protocols that are simple enough to be understood by MTurk workers, but comprehensive enough to test hypotheses of interest across the tree of life.
+Additionally, museums and other institutions must increase their efforts to make their biodiversity collections available digitally, including images suitable for morphological research.
+The problem of difficult-to-retrieve *dark data* is well-known [@Heidorn2008], but without either physical access to the collections or an image of the specimen, morphological data is impossible to acquire.
+
+Our results suggest that, where possible, crowdsourcing should be an integral part of any large-scale morphological analysis.
+Crowdsourcing should play a key role in unlocking the "dark data" present in biodiversity collections by providing a high-throughput way to extract the phenotypic data present in specimens.
+Furthermore, coordinating efforts from digitizing museum collections, natural language processing and machine vision software, citizen scientists, expert morphologists and taxonomists, and crowdsourced Mechanical Turk workers would result in an extremely powerful pipeline that could generate a "phenoscape" across the tree of life.
 
@@
 
-We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as Tina Marcroft, Bruno Frédérich, Vanson Liu, Rosalia Aguilar, Ryan Ellingson, Francesco Pickens, Colette LaRochelle, and the `r (fish_reliability %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (fish_families %>% select(worker) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
-We also thank Dan Rabosky, Brian Sidlauskas, Matthew McGee, Adam Summers, and Michael Burns for insightful discussions about fish morphology and digitization protocols.
-Katie Staab, Tonya Kane, and their `r fish_reliability %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students served as beta testers for the methods in this study.
+We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as T.
+Marcroft, B.
+Frederich, V.
+Liu, R.
+Aguilar, R.
+Ellingson, F.
+Pickens, C.
+LaRochelle, and the `r (fish_reliability %>% filter(role == "turker") %>% distinct(sha1mac) %>% nrow()) + (fish_families %>% select(worker) %>% distinct() %>% nrow()) ` Amazon Mechanical Turk workers that contributed their time and effort.
+We also thank D.
+Rabosky, B.
+Sidlauskas, M.
+McGee, A.
+Summers, and M.
+Burns for insightful discussions about fish morphology and digitization protocols.
+M.
+Venzon and T.
+Claverie provided unpublished figures that assisted this study.
+K.
+Staab and T.
+Kane allowed `r fish_reliability %>% filter(role == "student") %>% distinct(sha1mac) %>% nrow()` undergraduate students to beta test the methods.
 This work was supported by an Encyclopedia of Life David M.
 Rubenstein Fellowship (EOL-33066-13), a Stephen and Ruth Wainwright Fellowship, and a UCLA Research and Conference Award to JC.
-(XXX: Mike's grants?).
-Travel support to present this research was also provided by the Society for Study of Evolution to JC.
+Travel support to present this research was provided by the Society for Study of Evolution.
 
-## Author contributions
+# Data Accessibility
+
+All data are deposited online at the Encyclopedia of Life and Dryad.
+
+# Author contributions
 
@@
 
-## Supplementary material
-
-<!-- reset figure numbering for supplement -->
-
-\renewcommand\thefigure{S\arabic{figure}}
-\setcounter{figure}{0}
-
-```{r}
-knitr::opts_chunk$set(fig.cur=1)
-```
-
-Below is an example JSON file that demonstrates the utility of our web app:
-
-```json
-{
-    "C": {
-        "kind": "point",
-        "help": "Click the center of the eye."
-    },
-    "D": {
-        "kind": "line",
-        "help": "Click and drag from the left edge of the eye
-                 to the right edge of the eye."
-    },
-    "O": {
-        "kind": "curve",
-        "help": "Click and drag over the outline of the eye,
-                 starting from the leftmost point of the eye."
-    }
-}
-```
-
-Each digitization task to be completed is given a short abbreviation to aid task identification ("C" for center, "D" for diameter, and "O" for outline), and the type of task, "point", "line" or "curve", for homologous landmarks, linear measurements, and sliding semilandmarks can be specified.
-There is also an optional short help snippet displayed inline, which serve as a brief reminder for each landmark and complements a larger and more detailed protocol document that workers are required to read before beginning work.
-
-![A screenshot of the web app that turkers used to digitize images. A live demonstration is available at <https://jonchang.github.io/eol-mturk-landmark/>](interface.png)
-
-```{r turker_morph_procrustes_other_pc23, message=FALSE, warning=FALSE, fig.cap="PC2 and PC3 for turker / morphologist Procrustes space"}
-coords_df <- pca_df(coords)
-plot_tangent_space(coords_df, "PC2", "PC3")
-```
-
-```{r turker_morph_procrustes_other_pc34, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
-plot_tangent_space(coords_df, "PC3", "PC4")
-```
-
-```{r turker_morph_procrustes_other_pc45, message=FALSE, warning=FALSE, fig.cap="PC4 and PC5 for turker / morphologist Procrustes space"}
-plot_tangent_space(coords_df, "PC4", "PC5")
-```
-
-<!-- spit out all figures before references -->
-\clearpage
-
 # References
 
-```{r knitcitations_write, message=FALSE}
+```{r knitcitations_write, cache=FALSE, message=FALSE}
 write.bibtex(file="knitcitations.bib")

### 2015-09-16 R/fish.reliability.R

@@ -1,0 +1,19 @@
+#' fish.reliability: Functions and data for the fish reliability manuscript
+#'
+#' @section Plotting functions:
+#' \link{pca_df}, \link{plot_tangent_space}
+#'
+#' @section BAMM functions:
+#' \link{calculate_sampling}, \link{make_sampling_file}
+#'
+#' @section data:
+#' \link{fish_reliability}, \link{fish_families}
+#'
+#' @docType package
+#' @name fish.reliability
+#' @import dplyr
+#' @import grDevices
+#' @import directlabels
+#' @import dplyr
+#' @import ggplot2
+NULL

### 2015-09-16 R/functions.R

@@ -20,12 +20,21 @@ ip.lda <- function(object, newdata) {
 
 #' Formats a data frame for \code{\link[geomorph]{gpagen}}
 #'
-#' @param df a data frame, with certain columns
+#' @param df a data frame. must include columns "x" and "y"
+#' @param ... columns to include as labels
+#' @param .dots used to work around nonstandard evaluation
 #' @return A 3D matrix
-#' @import dplyr
 #' @export
-format_for_gpagen <- function(df) {
-  splut <- tidyr::gather(df, variable, val, x:y) %>% tidyr::unite(col=id, family, role, sha1mac)
+format_for_gpagen <- function(df, ...) {
+  format_for_gpagen_(df, .dots = lazyeval::lazy_dots(...))
+}
+
+
+#' @export
+#' @rdname format_for_gpagen
+format_for_gpagen_ <- function(df, ..., .dots) {
+  dots <- lazyeval::all_dots(.dots, ..., all_named = TRUE)
+  splut <- tidyr::gather(df, variable, val, x:y) %>% tidyr::unite_("id", names(dots))
   xtabs(val ~ mark + variable + id, data=splut)
 }
 
@@ -58,7 +67,9 @@ my.procD.lm <- function (f1, iter = 999, RRPP = FALSE, int.first = FALSE, verbos
   SS.obs <- anova.parts.obs$SS[1:k]
   P[, , 1] <- SS.obs
   method <- ifelse(RRPP, "RRPP", "resample")
-  result <- simplify2array(parallel::mclapply(1:iter, function(x) geomorph:::SS.random(Y, Xs, SS.obs, Yalt=method)$SS, mc.cores=parallel::detectCores()))
+  cl <- parallel::makeCluster(parallel::detectCores(), "PSOCK")
+  result <- simplify2array(parallel::parLapply(cl, 1:iter, function(x) geomorph:::SS.random(Y, Xs, SS.obs, Yalt=method)$SS))
+  parallel::stopCluster(cl)
   P[, , 1:iter+1] <- result
   P.val <- geomorph:::Pval.matrix(P)
   Z <- geomorph:::Effect.size.matrix(P)
@@ -76,3 +87,38 @@ my.procD.lm <- function (f1, iter = 999, RRPP = FALSE, int.first = FALSE, verbos
   }
   else anova.tab
 }
+
+#' Gets the GPA and also vectors of labels
+#' @export
[…]

### 2015-09-16 R/plotting.R

@@ -4,7 +4,6 @@
 #' @param rename_rownames Rename and remove the rownames? If TRUE (the default), converts rownames to a column in the data frame called `id`. If a character string, instead calls that column by that string. If FALSE, preserves the rownames as-is.
 #' @return an object of class `tbl_df`, `tbl`, and `data.frame`, with columns PC1 to PC...
 #' @export
-#' @import dplyr
 pca_df <- function(coords, rename_rownames = TRUE) {
   x <- geomorph::two.d.array(coords)
   pc.res <- stats::prcomp(x)
@@ -28,32 +27,102 @@ pca_df <- function(coords, rename_rownames = TRUE) {
 #' @param axis2 The name of the second axis to plot
 #' @return A \code{ggplot} object
 #' @export
-#' @import dplyr
-#' @import ggplot2
 plot_tangent_space <- function(coords_df, axis1="PC1", axis2="PC2") {
-  res2 <- coords_df %>% tidyr::separate(id, into=c("family", "role","sha"), sep="_")
-
   # consensus in Procrustes space for downstream highlighting
   dots <- list(
     x=lazyeval::interp(~median(axis1, na.rm=T), axis1=as.name(axis1)),
     y=lazyeval::interp(~median(axis2, na.rm=T), axis2=as.name(axis2)))
-  consensus <- res2 %>% group_by(family, role) %>% summarise_(.dots=dots)
+  consensus <- coords_df %>% group_by(family, role) %>% summarise_(.dots=dots)
 
-  chulls <- res2 %>% group_by(family) %>% do(.[chull(.[[axis1]], .[[axis2]]), ])
+  chulls <- coords_df %>% group_by(family) %>% do(.[chull(.[[axis1]], .[[axis2]]), ])
 
-  ggplot(res2, aes_string(axis1, axis2)) + geom_polygon(aes_string(axis1, axis2, fill="family", color="family"), data=chulls, alpha=0.5) + geom_point(aes(color=role)) + geom_path(aes(x, y, group=family), data=consensus) + geom_point(aes(x, y), data=consensus, size=3, color="black") + geom_point(aes(x, y, color=role), data=consensus, size=2)	+ geom_dl(aes(label=family), data=res2, list("smart.grid")) + theme_minimal() + theme(legend.position="none")
[…]

### 2015-09-16 R/stats.R

@@ -21,7 +21,6 @@ mad_2d <- function(x, y, constant = 1.4826) {
 #' @param y a numeric vector
 #' @param group a vector that specifies how \code{x} and \code{y} should be grouped
 #' @export
-#' @import dplyr
 #' @importFrom stats median dist
 dist_2d <- function(x, y, group) {
   if (length(x) != length(y))

### 2015-09-16 R/utils.R

@@ -1,0 +1,86 @@
+#' Do template substitution on a file
+#'
+#' Given a file with strings that look like \code{\{templ\}}, replace it
+#'
+#' @param template string with templates that look like \code{\{template\}}
+#' @param ... substitutions to perform
+#' @param .dots Used to work around non-standard evaluation. See
+#' \code{vignette("nse", "dplyr")} for details
+#' @return a character vector
+#' @export
+#' @examples
+#' substitute_template("{animal} says {noise}", animal="dog", noise="bark")
+substitute_template <- function(template, ...) {
+  substitute_template_(template, .dots = lazyeval::lazy_dots(...))
+}
+
+#' @export
+#' @rdname substitute_template
+substitute_template_ <- function(template, ..., .dots) {
+  dots <- lazyeval::all_dots(.dots, ..., all_named = TRUE)
+  for (name in names(dots)) {
+    value <- lazyeval::lazy_eval(dots[[name]])
+    template <- str_replace_all(template, fixed(paste0("{", name, "}")), value)
+  }
+  template
+}
+
+
+#' Calculates sampling fractions for BAMM
+#'
+#' @param phy an object of class \code{"phylo"}
+#' @param grouping the grouping level at which to calculate sampling fraction
+#' @param richness a data frame, with columns of groups (e.g., family, genus)
+#' and a final column \code{richness} specifying how many species are in that group
+#' @param tips that should be retained in the tree, but ignored for the
+#' purposes of calculating taxon sampling.
+#' @return a list with two components:
+#' \describe{
+#'   \item{global}{a numeric vector of length 1; the global sampling fraction
+#'   in \code{phy}}
+#'   \item{clade}{a data frame with three columns: tip labels, the grouping,
+#'   and the sampling fraction}
+#' }
+#' @seealso \code{\link{make_sampling_file}} to write a sampling file in BAMM's format
+#' @export
+calculate_sampling <- function(phy, grouping, richness, ignore.tips=c()) {
+  calculate_sampling_(phy, lazyeval::lazy(grouping), richness, ignore.tips=c())
+}
+
+#' @rdname calculate_sampling
+#' @export
+calculate_sampling_ <- function(phy, grouping, richness, ignore.tips=c()) {
+  quot <- as.character(grouping$expr)
+  tips <- data_frame(tip = phy$tip.label) %>%
[…]

### 2015-09-16 data-raw/fish_families.R

@@ -24,4 +24,4 @@ fish_families <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=hmac(s
 
 write_csv(fish_families, path = "fish_families.csv")
 
-use_data(fish_families, overwrite = TRUE)
+use_data(fish_families, overwrite = TRUE, compress = "xz")

### 2015-09-16 data-raw/fish_reliability.R

@@ -61,4 +61,4 @@ fish_reliability <- hugedata %>% select(workerId) %>% distinct(workerId) %>% row
 
 write_csv(fish_reliability, path = "fish_reliability.csv")
 
-use_data(fish_reliability, overwrite = TRUE)
+use_data(fish_reliability, overwrite = TRUE, compress="xz")

### 2015-09-17 manuscript/manuscript.Rmd
Travis: fix format_for_gpagen error
Travis: fix sequence #2

@@
 
-coords <- gpagen(format_for_gpagen(df, family, role, sha1mac, sequence), ShowPlot=F)$coords
+coords <- gpagen(format_for_gpagen(df, family, role, sha1mac), ShowPlot=F)$coords
 
@@
 df <- df %>% filter(!(family == "Balistidae" & mark %in% c("D1", "D2")), !(family == "Tetraodontidae" & mark %in% c("O4", "O5")))
-coords <- gpagen(format_for_gpagen(df, family, role, sha1mac, sequence), ShowPlot=F)$coords
+coords <- gpagen(format_for_gpagen(df, family, role, sha1mac), ShowPlot=F)$coords
 
-turk_morph_pca <- pca_df(coords) %>% separate(id, c("family", "role", "ind", "seq"), sep = "_")
+turk_morph_pca <- pca_df(coords) %>% separate(id, c("family", "role", "ind"), sep = "_")
 

### 2015-09-17 manuscript/supplement.Rmd
Add supplement Rmd

---
title: Supplementary material for "Crowdsourced geometric morphometrics enable rapid large-scale collection and
  analysis of phenotypic data"
author:
- affiliation: ucla
  email: jonathan.chang\@ucla.edu
  footnote: Author for correspondence jonathan.chang\@ucla.edu
  name: Jonathan Chang
- affiliation: ucla
  name: Michael E. Alfaro
bibliography:
- library.bib
- knitcitations.bib
output:
  pdf_document:
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    keep_tex: yes
    latex_engine: xelatex
    template: inc/mstemplate.tex
  html_document:
    toc: yes
  word_document: default
documentclass: inc/msarticle
endfloat: yes
fontsize: 11pt
geometry: margin=1.2in
layout: preprint
lineno: no
linestretch: 2
csl: inc/methods-in-ecology-and-evolution.csl
address:
- address: |
    Department of Ecology and Evolutionary Biology, University of California, Los Angeles, CA, USA
  code: ucla
---

```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
library("MASS")
library("readr")
library("dplyr")
library("ggplot2")
library("stringr")
library("magrittr")
library("tidyr")
library("reshape2")
library("scales")
library("geomorph")
library("rjson")
library("ipred")
library("vegan")
library("lubridate")
library("directlabels")
library("BAMMtools")
library("knitr")
library("kfigr")
library("knitcitations")
library("parallel")
library("fish.reliability")
options("citation_format"="pandoc")
cleanbib()
set.seed(111) # reproducible and stable sampling
data(fish_reliability)
data(fish_families)
knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center")
```

# Supplementary material

<!-- reset figure numbering for supplement -->

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}

```{r}
knitr::opts_chunk$set(fig.cur=1)
```

Below is an example JSON file that demonstrates the utility of our web app:

```json
{
    "C": {
        "kind": "point",
        "help": "Click the center of the eye."
    },
    "D": {
        "kind": "line",
        "help": "Click and drag from the left edge of the eye
                 to the right edge of the eye."
    },
    "O": {
        "kind": "curve",
        "help": "Click and drag over the outline of the eye,
                 starting from the leftmost point of the eye."
    }
}
```

Each digitization task to be completed is given a short abbreviation to aid task identification ("C" for center, "D" for diameter, and "O" for outline), and the type of task, "point", "line" or "curve", for homologous landmarks, linear measurements, and sliding semilandmarks can be specified. There is also an optional short help snippet displayed inline, which serve as a brief reminder for each landmark and complements a larger and more detailed protocol document that workers are required to read before beginning work.

![A screenshot of the web app that turkers used to digitize images. A live demonstration is available at <https://jonchang.github.io/eol-mturk-landmark/>](interface.png)

![Landmarks (J1) rostral tip of premaxilla (J2) ventral tip of premaxilla (J3) rostral tip of dentary (E1) anterior margin of midline through eye (E2) posterior margin of midline through eye (O1) dorsal end of preopercle (O2) ventral elbow of preopercle (O3) anterior end of preopercle (O4) dorsal end of opercle (O5) posterior end of opercle (D1) anterior insertion of dorsal fin (D2) distal tip of the anterior dorsal fin ray (D3) distal tip of the posterior dorsal fin ray (D4) posterior insertion of dorsal fin (P1) dorsal insertion of pectoral fin (P2) distal tip of the dorsal pectoral fin ray (P3) distal tip of the ventral pectoral fin ray (P4) ventral insertion of anal fin (A1) anterior insertion of anal fin (A2) distal tip of the anterior anal fin ray (A3) distal tip of the posterior anal fin ray (A4) posterior insertion of anal fin (C1) dorsal insertion of the caudal fin (C2) distal tip of the dorsal caudal fin ray (C3) distal tip of the the ventral caudal fin ray (C4) ventral insertion of the caudal fin (C5) midpoint of the caudal margin of the caudal peduncle](landmark_figure/landmarks.pdf)

```{r labeled_turker_v_morph, fig.cap="Version of Figure 1 where points are annotated with the landmark label.", warning=F, message=F}
no_st <- fish_reliability %>% filter(role != "student") %>% mutate(role = factor(role))

accuracy <- no_st %>% group_by(family, mark) %>% summarise(median = dist_2d(x, y, role)) %>% group_by(family) %>% mutate(outlier=upper_outlier(median, extreme = 2))

precision <- no_st %>% group_by(role, family, mark) %>% summarise(var = mad_2d(x, y)) %>% spread(role, var) %>% group_by(family, mark) %>% transmute(mad = log(turker / morphologist))

full_family <- data_frame(
  family=c("Aca", "Apo", "Bal", "Cha", "Gob", "Lab", "Pom", "Sco", "Tet"),
  full_family=c("Acanthuridae", "Apogonidae", "Balistidae", "Chaetodontidae", "Gobiidae", "Labridae", "Pomacentridae", "Scorpaenidae", "Tetraodontidae")
)

all_stats <- left_join(accuracy, precision) %>% mutate(outlier_label = ifelse(outlier, as.character(mark), NA))

ggplot(all_stats, aes(median, mad, color=outlier, label=mark)) + geom_text(size=2.5) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance / expert variance ") + facet_wrap(~family) + scale_color_manual(values = c("black", "red")) + theme_bw() + theme(legend.position="none") + xlim(0, 230)
```

```{r turker_morph_procrustes_other_pc34, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
turk_morph_pca <- read_csv("turk_morph_pca_for_supplement.csv")

plot_tangent_space(turk_morph_pca, "PC3", "PC4")
```

```{r turker_morph_procrustes_other_pc56, message=FALSE, warning=FALSE, fig.cap="PC5 and PC6 for turker / morphologist Procrustes space", eval=FALSE}
plot_tangent_space(turk_morph_pca, "PC5", "PC6")
```

```{r sevenfamily_extra_pc, message=FALSE, warning=FALSE, fig.cap = "PC3-6 for seven families of ray finned fishes."}
pcaed <- read_csv("seven_family_pca_for_supplement.csv")
plot_tangent_space2(pcaed, "PC3", "PC4")
plot_tangent_space2(pcaed, "PC5", "PC6")
```

\clearpage

### 2015-09-22 manuscript/manuscript.Rmd
Code drop

@@
 ```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
-library("MASS")
+?library("MASS")
 library("readr")
@@
 knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center")
+options(mc.cores = parallel::detectCores())
 ```
@@
 Although it is now possible to build phylogenetic trees with thousands of tips, and phenotypic data sets have similarly been growing larger and larger, the traits that are typically studied at this scale tend to be simple: geographic occurrences [@Jetz2012], one or two continuous characters [@Harmon2010; @Rabosky2013], a single discrete character [@Goldberg2010; @Edwards2011; @Price2012], or some combination of these [@Zanne2014; @Pyron2014].
-A richer understanding of the forces that shape macroevolution requires the collection of more detailed phenotypic trait data at scale.
+A richer understanding of the forces that shape macroevolution requires the collection of more detailed phenotypic trait data at a broad scale.
 
@@
 
-Collecting landmark-based geometric morphometric data at scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007].
+Collecting landmark-based geometric morphometric data at a broad scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007].
 To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked turkers ($n = `r nrow(filter(fish_reliability, role == "turker") %>% distinct(sha1mac))`$) and experts ($n = `r nrow(filter(fish_reliability, role == "morphologist") %>% distinct(sha1mac))`$) to landmark a set of five fish images, five times each.
@@
 The landmarks were carefully selected based on previously-published literature concerning fish shape [Supplemental Figure S2; @Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Claverie2014; @Thacker2014].
-We also ensured that the chosen landmarks included morphological features that were relatively straightforward to digitize (the position of the eye) and features that were likely to be more challenging to digitize (the position of the preopercle bone), in order to test for turker and expert differences over a spectrum of difficulties.
+We also ensured that the chosen landmarks included morphological features that were relatively straightforward to digitize (e.g., the position of the eye) and features that were likely to be more challenging to digitize (e.g., the most anterior and most dorsal points of the preopercle), in order to test for turker and expert differences over a spectrum of difficulties.
 We report the inter-observer reliability for turkers and experts by computing the ratio of the among-individual and the sum of the among-individual and measurement error variance components in a repeated measures nested MANOVA [@Palmer1986; @Zelditch2012].
 
-To assess the differences between turker and experts on a per-landmark basis, we first compared the median turker position to the median expert position of each landmark.
-We assumed that the expert median was the true position of that landmark, and calculated the absolute Euclidian distance.
+To assess the differences between turker and experts on a per-landmark basis, we first compared for each landmark the median position of all turkers to the median position of all experts.
+We assumed that the expert median was the true position of that landmark, and calculated the absolute Euclidian distance in pixels.
 Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy.
@@
 To demonstrate the utility of obtaining comparative data using this method, we use previously published phylogenies for seven fish families: Acanthuridae [@Sorenson2013], Balistoidae, Tetraodontidae [@Santini2013], Apogonidae, Chaetodontidae, Labridae [@Cowman2011; @Choat2012], and Pomacentridae [@Frederich2013].
-We matched `r nrow(distinct(fish_families, tip))` species to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [@Parr2014].
+We matched `r nrow(distinct(fish_families, tip))` species to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [Supplemental Table S1; @Parr2014].
 Crowdsourced workers placed landmarks describing body shape variation following a standard protocol (Supplementary Material).
 The Cartesian position of these landmarks were used in a generalized Procrustes analyses [@Gower1975; @Rohlf1990], which centers, scales, and rotates landmark configurations to minimize the least-squares distance between shapes.
-We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] with the R package *geomorph* `r citep(citation("geomorph"))`, and used these principal components axes for subsequent analyses.
+We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] with the R package *geomorph* `r citep(citation("geomorph"))`, and retain the principal component axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998] for all subsequent analyses.
 
-We used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of speciation and body shape evolution for all seven families.
-For the characters describing body shape, we use the PC axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998].
-BAMM estimates the location of rate shifts in either diversification or character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of lineage diversification and trait evolution.
+We used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of body shape evolution for all seven families.
+BAMM estimates the location of rate shifts in character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of lineage diversification and trait evolution.
 We assessed convergence and mixing using Tracer [@rambaut2007tracer].
@@
 The inter-observer reliability of turkers and experts as measured by the ratio of among-individual and sum of the among-individual and measurement error ANOVA components was `r percent(reliability$turker)` and `r percent(reliability$expert)`, respectively.
-Although there is no current standard for acceptable levels of measurement reliability [@VonCramon-Taubadel2007], these percentages are not low enough to suggest pathologies in the measurement protocol.
+Although there is no current standard for acceptable levels of measurement reliability [@VonCramon-Taubadel2007], these percentages are not low enough to suggest weaknesses in the measurement protocol.
 
@@
 
-We projected turker and expert shape configurations into morphospace (Figure `r figr("turk_morph_procrustes_space")`, Supplemental Figure S4) Although the overall space occupied by each family's shape configurations vary, in practice, the aggregated median turker and expert shapes are not qualitatively different.
+We projected turker and expert shape configurations into morphospace (Figure `r figr("turk_morph_procrustes_space")`, Supplemental Figure S4) Although the overall space occupied by each family's shape configurations vary, the aggregated median turker and expert shapes are not qualitatively different.
 The only exception is the triggerfishes (Balistidae), likely due to turker confusion over the exact location of dorsal fin due to their reduced anterior dorsal fin.
@@
 
-Using a median expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(distinct(fish_families, tip)))` to landmark all `r nrow(distinct(fish_families, tip))` images.
-At 5x replication, this would take `r as.character(morpho_per_image * nrow(fish_families))`.
-By comparison, turkers took a total of `r as.character(total_turker_time)` to complete all images at 5x replication.
+Turkers in aggregate required `r as.character(total_turker_time)` to complete all `r nrow(distinct(fish_families, tip))` images at 5x replication `r figr("timing_plot")`.
+We estimate that a single expert would need `r as.character(morpho_per_image * nrow(distinct(fish_families, tip)))` to complete all images at 1x replication, extrapolated from a median expert time per image of `r as.character(morpho_per_image)`.
+At 5x replication, our projected expert would instead need `r as.character(morpho_per_image * nrow(fish_families))`.
 
@@
 
+```{r}
+
+par_get_rtt_matrix <- function(ephy, start.time=NULL, end.time=NULL, nslices=100, node=NULL, nodetype = 'include', mc.cores = getOption("mc.cores", 2L)) {
+
+	if (!'bammdata' %in% class(ephy)) {
+		stop("Object ephy must be of class 'bammdata'\n");
+	}
+  if (ephy$type == 'diversification') {
+	  stop("'diversification' not supported")
+	}
+
+	if (is.null(node)) {
+		nodeset <- ephy$edge[,2];
+	} else if (!is.null(node) & nodetype == 'include') {
+		#nodeset <- getDesc(ephy, node)$desc_set;
+		nodeset <- unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set))
+	} else if (!is.null(node) & nodetype == 'exclude') {
+		nodeset <- setdiff( ephy$edge[,2], unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set)));
+	} else {
+		stop('error in getRateThroughTimeMatrix\n');
+	}
+
+	bt <- branching.times(as.phylo.bammdata(ephy));
+	maxpossible <- max(bt[as.character(intersect(nodeset, ephy$edge[,1]))]);
+
+	if (is.null(start.time)) {
+		start.time <- max(bt) - maxpossible;
+	}
+	if (is.null(end.time)) {
+		end.time <- max(bt);
+	}
+
+	decimals = function(x){
+	if(x%%1 != 0)	{
+		return(nchar(strsplit(as.character(x),".",fixed=TRUE)[[1]][[2]]));
+	}		else	{
+		return(10);
+	}
+}
+
+	tvec <- seq(start.time, end.time, length.out= nslices);
+	tol = 1*10^-decimals(ephy$eventBranchSegs[[1]][1,2]);
+	mm <- matrix(NA, nrow=length(ephy$eventBranchSegs), ncol=length(tvec));
+
+	# this is much faster than 'safeCompare'.
+	goodTime <- function (vec, val, tol) {
+		(vec[,2] - val <= tol) & (val - vec[,3] <= tol);
+	}
+
+	res <- mclapply(mc.cores = mc.cores, 1:length(ephy$eventBranchSegs), function(i) {
+	  	es <- ephy$eventBranchSegs[[i]];
+		events <- ephy$eventData[[i]];
+
+		isGoodNode <- rep(TRUE, nrow(es));
+
+		sapply(1:length(tvec), function(k) {
+			isGoodTime <- goodTime(es, tvec[k], tol=tol);
+			if (!(is.null(node))) { # only enter this if not the root. otherwise, only have to set once per i.
+				isGoodNode <- es[,1] %in% nodeset;	
+			}
+
+			estemp <- es[isGoodTime & isGoodNode, ];
+			tvv <- tvec[k] - events$time[estemp[,4]];
+			rates <- exponentialRate(tvv, events$lam1[estemp[,4]], events$lam2[estemp[,4]]);
+			mean(rates);
+		})
+	})
+
+	mm <- do.call(rbind, res)
+
+	obj <- list();
+	obj$beta <- mm;
+	obj$times <- tvec;
+
+	class(obj) <- 'bamm-ratematrix';
+	obj$type = 'trait';	
+	return(obj);
+}
+
+```
+
 ```{r load_bamm_data, message=FALSE, echo=FALSE, warning=FALSE, results="hide"}
-base <- "data-raw/bamm/"
+#base <- "data-raw/bamm/"
 base <- "../data-raw/bamm/"
@@
 tre_trait <- lapply(paste0(base, "input_trees/", fams, ".tre"), read.tree)
-tre_speciation <- lapply(paste0(base, fams, ".tre"), read.tree)
-names(tre_trait) <- names(tre_speciation) <- fams
+names(tre_trait) <- fams
 
 cl <- makePSOCKcluster(detectCores(), useXDR=F)
-clusterExport(cl, list("base", "fams", "tre_speciation", "tre_trait"))
-evt_speciation <- parLapplyLB(cl, fams, function(fam) {
-  tre <- tre_speciation[[fam]]
-  BAMMtools::getEventData(tre, paste0(base, "div_runs/", fam, "_event_data.txt"), burnin=0.1)
-})
+clusterExport(cl, list("base", "fams", "tre_trait"))
 evt_trait <- parLapplyLB(cl, fams, function(fam) {
@@
 })
-names(evt_trait) <- names(evt_speciation) <- fams
-clusterExport(cl, list("evt_speciation", "evt_trait"))
-rtt_speciation <- parLapplyLB(cl, fams, function(fam) {
-  BAMMtools::getRateThroughTimeMatrix(evt_speciation[[fam]])
-})
+names(evt_trait) <- fams
+clusterExport(cl, list("evt_trait"))
 rtt_trait <- parLapplyLB(cl, fams, function(fam) {
   BAMMtools::getRateThroughTimeMatrix(evt_trait[[fam]])
-})
-prior_speciation <- parLapplyLB(cl, fams, function(fam) {
-  BAMMtools::getBranchShiftPriors(tre_speciation[[fam]], paste0(base, "div_runs/", fam, "_prior_probs.txt"))
 })
@@
 
-names(pretty_names) <- names(prior_trait) <- names(prior_speciation) <- names(rtt_trait) <- names(rtt_speciation) <- fams
+names(pretty_names) <- names(prior_trait) <- names(rtt_trait) <- fams
 stopCluster(cl)
@@
 
-```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 (a, b) and speciation (c, d) across seven families of fishes.
-Phylorate plots (a, c) color branch lengths by rates of shape evolution (a) and speciation (c), where warmer colors indicate faster rates of evolution.
+```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 across seven families of fishes.
+(a) Phylorate plots color branch lengths by rates of shape evolution, where warmer colors indicate faster rates of evolution.
 Significant rate shift events (pp > 0.95) are indicated on the phylorate plot as a red circle on the corresponding branch.
-Median log rates of shape evolution (b) and speciation (d) through time, where black lines indicate the background rate and red lines indicate the rate of evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a) or (c).", fig.height=9.5, out.height="0.9\\textheight", out.width="\\textwidth"}
-
-layout_mat <- matrix(c(
-  0,  1,  2,  3,  4,
-  5,  12, 13, 14, 15,
-  6,  16, 17, 18, 19,
-  7,  20, 21, 22, 23,
-  8,  24, 25, 26, 27,
-  9,  28, 29, 30, 31,
-  10, 32, 33, 34, 35,
-  11, 36, 37, 38, 39
-), ncol = 5, byrow = TRUE)
+(b) Median log rates of shape evolution through time, where black lines indicate the background rate and red lines indicate the rate of phenotypic evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a).", fig.width = 6, fig.height = 3, out.height="3in", out.width="6in"}
 
@@
 
-plot_rtt <- function(rtt, shifts = list(), xlim=NULL, ylim=NULL) {
+plot_rtt <- function(rtt, shifts = list(), xlim=NULL, ylim=NULL, labels = TRUE) {
   vartype <- if (rtt$type == "trait") "beta" else "lambda"
@@
   })
-  xlim <- if (is.null(xlim)) range(sapply(allshifts, `[`, "times")) else xlim
-  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[`, "avg")) else ylim
-  old.par <- par(mar = c(2,3,2,1))
+  xlim <- if (is.null(xlim)) c(0, max(sapply(allshifts, `[[`, "times"))) else xlim
+  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[[`, "avg")) else ylim
   plot.new()
@@
   axis(at = axTicks(1), side = 1, labels = axTicks(1))
-  axis(side = 2, las = 2)
+  axis(at = c(1e-4, 1e-3, 1e-2, 1e-1), side = 2, las = 2, labels = labels)
   lines(allshifts[[1]]$times, allshifts[[1]]$avg)
@@
     lapply(2:length(allshifts), function (x) {
-      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red")
+      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red", lwd=1.5)
     })
   }
-  par(old.par)
   return(invisible(NULL))
@@
 
-mar <- c(2,1,2,1)
+mar <- c(1.5,1,0,0)+0.1
 par(mar = c(0,0,0,0))
-layout(layout_mat, widths = c(1, 6, 6, 6, 6), heights = c(0.9, 9, 9, 9, 9, 9, 9, 9))
-htext("Shape Phylorates")
-text(0, 0.5, "a)", cex=1.2)
-htext("Shape Rate Through Time")
-text(0, 0.5, "b)", cex=1.2)
-htext("Speciation Phylorates")
-text(0, 0.5, "c)", cex=1.2)
-htext("Speciation Rate Through Time")
-text(0, 0.5, "d)", cex=1.2)
-lapply(pretty_names, vtext)
+
+layout_mat <- matrix(c(
+  8, 10, 12, 14, 16, 18, 20, 22, # shape phylorates
+  9, 11, 13, 15, 17, 19, 21, 23, # shape rate through time
+  0, 1, 2, 3, 4, 5, 6, 7
+), nrow = 3, byrow = TRUE)
+
+layout(layout_mat, widths = c(1.3, 12, 9, 9, 9, 9, 9, 9), heights = c(5, 6, 1))
+lapply(pretty_names, htext) # family names
+vtext("(a) Shape Phylorates")
+vtext("(b) Shape Rate Through Time")
 par(mar = mar)
+
+ctr <- 1
 lapply(fams, function(fam) {
+  par(mar=mar)
+  if (ctr == 1) {
+    par(mar=c(1.5,4,0,0)+0.1)
+  }
   interval <- if(is.null(customintervals[[fam]])) c(0, 0.001) else customintervals[[fam]] 
@@
   addBAMMshifts(best, cex = 2, par.reset=F)
-  axisPhylo()
-
+  #axis_phylo(label_ticks = FALSE)
+  axisPhylo(col.axis = "transparent")
   shift_node <- getShiftNodesFromIndex(best, 1)
   if (length(shift_node) > 0) {
-    background_rates <- getRateThroughTimeMatrix(evt_trait[[fam]], node = shift_node, nodetype = "exclude")
+    background_rates <- par_get_rtt_matrix(evt_trait[[fam]], node = shift_node, nodetype = "exclude", start.time=0)
     clade_rates <- lapply(shift_node, function (x) {
@@
       write.tree(phy, file = paste0(fam, "_trait_shift_", x, ".tre"))
-      getRateThroughTimeMatrix(evt_trait[[fam]], node = x)
+      par_get_rtt_matrix(evt_trait[[fam]], node = x)
       })
-    plot_rtt(background_rates, clade_rates)
+    plot_rtt(background_rates, clade_rates, ylim=c(4e-5, 0.1), labels = ctr == 1)
   } else {
-    plot_rtt(rtt_trait[[fam]])
+    plot_rtt(rtt_trait[[fam]], ylim=c(4e-5, 0.1), labels = ctr == 1)
   }
-
-  q <- plot.bammdata(evt_speciation[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = )
-  best <- getBestShiftConfiguration(evt_speciation[[fam]], prior = prior_speciation[[fam]])
-  addBAMMshifts(best, cex = 2, par.reset=F)
-  axisPhylo()
-
-  shift_node <- getShiftNodesFromIndex(best, 1)
-  if (length(shift_node) > 0) {
-    background_rates <- getRateThroughTimeMatrix(evt_speciation[[fam]], node = shift_node, nodetype = "exclude")
-    clade_rates <- lapply(shift_node, function (x) {
-      phy <- extract.clade(tre_speciation[[fam]], x)
-      write.tree(phy, file = paste0(fam, "_speciation_shift_", x, ".tre"))
-      getRateThroughTimeMatrix(evt_speciation[[fam]], node = x)
-      })
-    plot_rtt(background_rates, clade_rates)
-  } else {
-    plot_rtt(rtt_speciation[[fam]])
-  }
+  ctr <<- ctr + 1
 })
@@
 We project per-species consensus shapes into Procrustes space (Figure `r figr("sevenfamily_tangent_plot")`, Supplemental Figure S5).
-The BAMMtools analysis uncovered substantial amounts of heterogeneity in the rate of body shape evolution and speciation in each family (Figure `r figr("bamm_plots")`).
-Significant shifts in the rate of shape evolution or speciation were detected in three families: Labridae, Apogonidae, and Pomacentridae.
-The significant shifts in speciation rate corroborate those found in @Cowman2011 through either MEDUSA [@Alfaro2009] or a relative cladogenesis statistic [@Nee1992].
+The BAMMtools analysis uncovered heterogeneity in the rate of body shape evolution in each family (Figure `r figr("bamm_plots")`).
+Significant shifts in the rate of shape evolution were detected within two families: Labridae and Pomacentridae.
 Two significant shifts in shape evolution rate occur in the wrasses (Labridae).
@@
 The other shift is nested within that group, in *Sparisoma*.
-One shift in speciation rate also occurs in the wrasses, encompassing the genera *Chlorurus* and *Scarus*.
-One shift in speciation rate occurs in the cardinalfishes (Apogonidae), encompassing members of the genera *Apogon*, *Archamia*, *Zoramia*, *Ostorhinchus*, *Cheilodpterus*, *Gossamia*, *Fowleria*, and *Phaeoptyx* [Apogonini + Apogonichthynini *sensu* @Mabuchi2014].
 One shift in the rate of shape evolution occurs in the damelfishes (Pomacentridae) in the genus *Amphiprion*.

### 2015-09-22 manuscript/supplement.Rmd
Code drop

@@
 endfloat: yes
-fontsize: 11pt
-geometry: margin=1.2in
+fontsize: 10pt
+geometry: margin=1in
 layout: preprint
 lineno: no
-linestretch: 2
+linestretch: 1
 csl: inc/methods-in-ecology-and-evolution.csl
@@
 \setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
 
@@
 ![A screenshot of the web app that turkers used to digitize images. A live demonstration is available at <https://jonchang.github.io/eol-mturk-landmark/>](interface.png)
+
+## Testing turker vs expert consistency
+
+```{r geomorph_consistency, warning=FALSE}
+res <- gpa_with_vectors(only_good_points, family, role, sha1mac, sequence)
+
+family <- res$vectors$family
+role <- res$vectors$role
+repno <- as.integer(res$vectors$sequence) %% 5
+sha1mac <- as.factor(res$vectors$sha1mac)
+
+devtools::install_version("geomorph", version = "2.1.6", local=F, quick=T, dependencies=F, upgrade_dependencies=F, keep_source=T)
+
+devtools::install("~/geomorph/", local=F, quick=T, dependencies=F, upgrade_dependencies=F, keep_source=T)
+
+lm_result <- advanced.procD.lm(res$coords ~ family*role*sha1mac, ~family*sha1mac, iter=9)
+?lm_p <- lm_result$P[2]
+lm_Z <- lm_result$Z[2]
+lm_F <- lm_result$F[2]
+rm(family, role, repno, sha1mac)
+```
 
@@
 \clearpage
+
+```{r fish_families_table, message=FALSE, warning=FALSE, as.is=T}
+fish_families_urls <- read_csv("fish_families_urls.csv")
+knitr::kable(select(fish_families_urls, -eol_guid), caption='Families, species names, and links to the images hosted on Encyclopedia of Life for the section "Example: a phenomic pipeline for comparative phylogenetic analysis".')
+```

### 2015-09-22 R/functions.R

@@ -39,54 +39,6 @@ format_for_gpagen_ <- function(df, ..., .dots) {
 }
 
 
-#' Hack the shit out of geomorph's slow procD.lm function
-#' @export
-my.procD.lm <- function (f1, iter = 999, RRPP = FALSE, int.first = FALSE, verbose = FALSE)
-{
-  form.in <- formula(f1)
-  Y <- eval(form.in[[2]], parent.frame())
-  if (length(dim(Y)) == 3)
-    Y <- geomorph::two.d.array(Y)
-  else Y <- as.matrix(Y)
-  if (nrow(Y) != nrow(model.frame(form.in[-2])))
-    stop("Different numbers of specimens in dependent and independent variables")
-  form.new <- as.formula(paste(c("Y", form.in[[3]]), collapse = "~"))
-  if (int.first == TRUE)
-    ko = TRUE
-  else ko = FALSE
-  Terms <- terms(form.new, keep.order = ko)
-  mod.mf <- model.frame(Terms)
-  if (any(is.na(Y)) == T)
-    stop("Response data matrix (shape) contains missing values. Estimate these first (see 'estimate.missing').")
-  anova.parts.obs <- geomorph:::anova.parts(f1 = form.new, Yalt = "observed",
-                                 keep.order = ko)
-  anova.tab <- anova.parts.obs$table
-  Xs <- geomorph:::mod.mats(form.new, mod.mf, keep.order = ko)
-  k <- length(Xs$Xs) - 1
-  P <- array(0, c(k, 1, iter + 1))
-  SS.obs <- anova.parts.obs$SS[1:k]
-  P[, , 1] <- SS.obs
-  method <- ifelse(RRPP, "RRPP", "resample")
-  cl <- parallel::makeCluster(parallel::detectCores(), "PSOCK")
-  result <- simplify2array(parallel::parLapply(cl, 1:iter, function(x) geomorph:::SS.random(Y, Xs, SS.obs, Yalt=method)$SS))
-  parallel::stopCluster(cl)
-  P[, , 1:iter+1] <- result
-  P.val <- geomorph:::Pval.matrix(P)
-  Z <- geomorph:::Effect.size.matrix(P)
-  anova.tab <- data.frame(anova.tab, Z = c(Z, NA, NA), P.value = c(P.val,
-                                                                   NA, NA))
-  if (RRPP == TRUE) {
-    anova.title = "\nRandomized Residual Permutation Procedure used\n"
-  }
[…]

### 2015-09-22 manuscript/manuscript.Rmd
Fixup errors

@@
 ```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
-?library("MASS")
+library("MASS")
 library("readr")

### 2015-09-22 manuscript/supplement.Rmd
Fixup errors

@@
 library("fish.reliability")
-options("citation_format"="pandoc")
+options("citation_format"="pandoc", width=88)
 cleanbib()
@@
 
+## Landmarks used
+
+The landmarks used are shown in Supplemental Figure S2. The image is redrawn from @Chakrabarty2005, which was itself redrawn from @Nelson1994.
+
+![Landmarks (J1) rostral tip of premaxilla (J2) ventral tip of premaxilla (J3) rostral tip of dentary (E1) anterior margin of midline through eye (E2) posterior margin of midline through eye (O1) dorsal end of preopercle (O2) ventral elbow of preopercle (O3) anterior end of preopercle (O4) dorsal end of opercle (O5) posterior end of opercle (D1) anterior insertion of dorsal fin (D2) distal tip of the anterior dorsal fin ray (D3) distal tip of the posterior dorsal fin ray (D4) posterior insertion of dorsal fin (P1) dorsal insertion of pectoral fin (P2) distal tip of the dorsal pectoral fin ray (P3) distal tip of the ventral pectoral fin ray (P4) ventral insertion of anal fin (A1) anterior insertion of anal fin (A2) distal tip of the anterior anal fin ray (A3) distal tip of the posterior anal fin ray (A4) posterior insertion of anal fin (C1) dorsal insertion of the caudal fin (C2) distal tip of the dorsal caudal fin ray (C3) distal tip of the the ventral caudal fin ray (C4) ventral insertion of the caudal fin (C5) midpoint of the caudal margin of the caudal peduncle.](landmark_figure/landmarks.pdf)
+
 ## Testing turker vs expert consistency
@@
 ```{r geomorph_consistency, warning=FALSE}
+only_good_points <- fish_reliability %>% filter(role != "student", !mark %in% c("A3", "A4", "D3", "D4", "O1", "O2", "O3","O4","O5", "P3", "P4"))
 res <- gpa_with_vectors(only_good_points, family, role, sha1mac, sequence)
@@
 
-devtools::install_version("geomorph", version = "2.1.6", local=F, quick=T, dependencies=F, upgrade_dependencies=F, keep_source=T)
-
-devtools::install("~/geomorph/", local=F, quick=T, dependencies=F, upgrade_dependencies=F, keep_source=T)
-
-lm_result <- advanced.procD.lm(res$coords ~ family*role*sha1mac, ~family*sha1mac, iter=9)
-?lm_p <- lm_result$P[2]
+lm_result <- advanced.procD.lm(res$coords ~ family*role*sha1mac, ~family*sha1mac, .parallel.cores = 24, .parallel.cluster = parallel::makeForkCluster(24))
+lm_p <- lm_result$P[2]
 lm_Z <- lm_result$Z[2]
@@
 
-![Landmarks (J1) rostral tip of premaxilla (J2) ventral tip of premaxilla (J3) rostral tip of dentary (E1) anterior margin of midline through eye (E2) posterior margin of midline through eye (O1) dorsal end of preopercle (O2) ventral elbow of preopercle (O3) anterior end of preopercle (O4) dorsal end of opercle (O5) posterior end of opercle (D1) anterior insertion of dorsal fin (D2) distal tip of the anterior dorsal fin ray (D3) distal tip of the posterior dorsal fin ray (D4) posterior insertion of dorsal fin (P1) dorsal insertion of pectoral fin (P2) distal tip of the dorsal pectoral fin ray (P3) distal tip of the ventral pectoral fin ray (P4) ventral insertion of anal fin (A1) anterior insertion of anal fin (A2) distal tip of the anterior anal fin ray (A3) distal tip of the posterior anal fin ray (A4) posterior insertion of anal fin (C1) dorsal insertion of the caudal fin (C2) distal tip of the dorsal caudal fin ray (C3) distal tip of the the ventral caudal fin ray (C4) ventral insertion of the caudal fin (C5) midpoint of the caudal margin of the caudal peduncle](landmark_figure/landmarks.pdf)
+We also used a non-parametric MANOVA with RRPP to determine whether turkers were less consistent among individual replicates as compared to experts.
+The non-parametric MANOVA with RRPP failed to detect a significant difference between turker and expert shapes ($p = `r lm_p`$, $Z = `r lm_Z`$, $F=`r lm_F`$).
+
+## R information
+
+Information on the versions of R packages used to analyze this data.
+
+```{r session_info}
+devtools::session_info()
+```
 
@@
 ```
+
+## References

### 2015-09-22 manuscript/manuscript.Rmd
Re-roxygenize

@@
 data(fish_families)
-knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center")
+knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center", warning = FALSE)
 options(mc.cores = parallel::detectCores())

### 2015-09-22 R/zzzdata.R
Fix worker IDs getting lost

@@ -33,6 +33,7 @@
 #'   \item{y}{the y-coordinate of this landmark}
 #'   \item{family}{the family that this specimen belonged to}
 #'   \item{tip}{the species digitized}
-#'   \item{time_taken}{the amount of time taken by this worker to digitize this image}
+#'   \item{time_taken}{the amount of time taken for a worker to submit this task}
+#'   \item{worker_time_spent}{the amount of time a worker actually took to digitize the image}
 #' }
 "fish_families"

### 2015-09-22 data-raw/fish_families.R

@@ -20,7 +20,20 @@ get_marks <- function(blob, ...) {
   fromJSON(blob) %>% as.data.frame %>% t %>% as.data.frame %>% set_colnames(c("x", "y")) %>% mutate(mark=rownames(.)) %>% cbind(...)
 }
 
-fish_families <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=hmac(secret, .$workerId, algo="sha1"), family=.$family, tip=.$tip, time_taken=.$time_taken))
+get_duration <- function(blob, ...) {
+  fromJSON(blob) %>% lapply(unlist) %>% lapply(data.frame) %>% lapply(t) %>%
+    do.call(plyr::rbind.fill.matrix, .) %>% data.frame(stringsAsFactors=F) %>% tbl_df() %>%
+    unite(payload, -X1, -X2, sep=",") %>% rename(timestamp=X1, action=X2) %>%
+    mutate(timestamp = as.POSIXct(as.numeric(timestamp)/1000, origin="1970-01-01 00:00.00 UTC")) %>%
+    filter(str_detect(action, "^data")) %>%
+    summarise(worker_time_spent=last(timestamp)-first(timestamp)) %>% cbind(...)
+}
+
+durations <- rowwise(results) %>% do(get_duration(.$Answer.logger, worker=hmac(secret, .$workerid, algo="sha1"), family=.$family, tip=.$tip))
+
+marks <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=hmac(secret, .$workerid, algo="sha1"), family=.$family, tip=.$tip, time_taken=.$time_taken))
+
+fish_families <- left_join(marks, durations)
 
 write_csv(fish_families, path = "fish_families.csv")
 

### 2015-09-22 data-raw/fish_families.R
Ungroup fish_families data

@@ -33,7 +33,7 @@ durations <- rowwise(results) %>% do(get_duration(.$Answer.logger, worker=hmac(s
 
 marks <- rowwise(results) %>% do(get_marks(.$Answer.marks, worker=hmac(secret, .$workerid, algo="sha1"), family=.$family, tip=.$tip, time_taken=.$time_taken))
 
-fish_families <- left_join(marks, durations)
+fish_families <- left_join(marks, durations) %>% ungroup
 
 write_csv(fish_families, path = "fish_families.csv")
 

### 2016-08-05 manuscript/manuscript.Rmd
Code drop

@@
 
-**Keywords**: crowdsourcing, morphometrics, phenotyping, morphology, comparative methods, macroevolution, Actinopterygii
+**Keywords**: comparative methods, macroevolution, Actinopterygii, Mechanical Turk, large scale annotation
 
@@
 Integrating phenotypic data, such as anatomy, behavior, physiology, and other traits, with phylogenies is powerful strategy for investigating the patterns of biological evolution.
-Recent advances in next-generation sequencing [@Shendure2008; @Meyer2008] and sequence capture technologies [@Faircloth2012b; @Lemmon2012] have made phylogenetic inference of large radiations of organisms possible [@McCormack2012; @McCormack2013; @Faircloth2013; @Faircloth2014].
+Recent advances in next-generation sequencing [@Shendure2008; @Meyer2008] and sequence capture technologies [@Faircloth2012b; @Lemmon2012] have made phylogenetic inference of large radiations of organisms possible [@McCormack2012; @McCormack2013; @Faircloth2013; @Faircloth2015].
 However, similar breakthroughs for generating new phenotypic datasets have been comparatively uncommon, likely due to the high expense and effort required [reviewed in @Burleigh2013].
@@
 
-Although it is now possible to build phylogenetic trees with thousands of tips, and phenotypic data sets have similarly been growing larger and larger, the traits that are typically studied at this scale tend to be simple: geographic occurrences [@Jetz2012], one or two continuous characters [@Harmon2010; @Rabosky2013], a single discrete character [@Goldberg2010; @Edwards2011; @Price2012], or some combination of these [@Zanne2014; @Pyron2014].
-A richer understanding of the forces that shape macroevolution requires the collection of more detailed phenotypic trait data at a broad scale.
+Although it is now possible to build phylogenetic trees with thousands of tips, and phenotypic data sets have similarly been growing larger and larger, studies at this scale tend to be limited to a few broad types of traits, including geographic occurrences [@Jetz2012], one or two continuous characters [@Harmon2010; @Rabosky2013], a single discrete character [@Goldberg2010; @Edwards2011; @Price2012], or some combination of these [@Zanne2014; @Pyron2014].
+Most morphological evolutionary studies are constrained by a fundamental tradeoff in effort.
+Although the collection of detailed phenotypic measurements is often required to fully analyze complex form-function or ecology-phenotype relationships [@Schluter2000; @Alfaro2004; @Alfaro2005; @Wainwright2005; @Collar2006; @Price2010; @Frederich2013], rich methods of data collection such as computed tomography (CT) scanning are time intensive and do not permit easy scaling to hundreds or thousands of species.
+Analysis of more complex traits at this scale has the potential to greatly enrich our understanding of macroevolutionary processes, by permitting more refined hypothesis testing.
 
@@
 We evaluate the accuracy and precision of turkers by assigning identical image sets and digitization protocols to users who are experienced with fish morphology (hereafter *experts*), and compare the inter- and intra-observer differences between turkers and experts.
-To illustrate the efficiency of this approach, we construct a phylogenetic analysis pipeline to download photographs and phylogenies of seven actinopterygiian families from the web, collect Mechanical Turk shape results, analyze the rate of diversification and body shape evolution using BAMM [@Rabosky2014a], and compare the time required for this workflow to traditional approaches.
+To illustrate the efficiency of this approach, we construct a phylogenetic analysis pipeline to download photographs and phylogenies of seven actinopterygiian families from the web, collect Mechanical Turk shape results, analyze the body shape evolution using BAMM [@Rabosky2014a], and compare the time required for this workflow to traditional approaches.
+Though we focus on collecting two-dimensional geometric morphometric data, we address the challenges that will be common to all studies that crowdsource phenotypic data.
 We also discuss the role that crowdsourcing is best suited in large-scale morphological analyses, and suggest ways to integrate crowdsourced data as part of larger initiatives to digitize biodiversity.
@@
 Amazon Mechanical Turk ("MTurk") is a web-based service where Requesters can request work, known as Human Intelligence Tasks ("HITs") to be performed by Workers.
-Workers work from home and submit the tasks over the Internet, where Requesters review it, and, if they are satisfied with the results, accept the work and pay the Worker.
+Workers work from home and submit the tasks over the Internet, where Requesters review it, and, if they are satisfied with the results, accept the work and pay the Worker [for a detailed overview, see @Mason2012a].
 We use MTurk as a platform to distribute our geometric morphometric tasks and financially compensate the worker accordingly.
@@
 This simplifies the infrastructure challenge of needing to serve many crowdsourced workers simultaneously, since workers will not need to download desktop software such as tpsDig (<http://life.bio.sunysb.edu/ee/rohlf/software.html>) before generating data.
-The web application is configured with a simple JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task (Supplemental Figure S1).
+The web application is configured with a JavaScript Object Notation (JSON) file that describes the landmarks necessary to complete an image digitization task (Supplemental Figure S1).
 Point landmarks, semilandmark curves, and linear measurements are all supported.
@@
 To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked turkers ($n = `r nrow(filter(fish_reliability, role == "turker") %>% distinct(sha1mac))`$) and experts ($n = `r nrow(filter(fish_reliability, role == "morphologist") %>% distinct(sha1mac))`$) to landmark a set of five fish images, five times each.
-All participants used the same protocol and same software to digitize the same set of fishes.
+Turkers were compensated \$25 for the entire task.
+All participants used the same protocol (Supplemental Material) and same software to digitize the same set of fishes (Supplemental Table S1, S2).
 The landmarks were carefully selected based on previously-published literature concerning fish shape [Supplemental Figure S2; @Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Claverie2014; @Thacker2014].
@@
 We report the inter-observer reliability for turkers and experts by computing the ratio of the among-individual and the sum of the among-individual and measurement error variance components in a repeated measures nested MANOVA [@Palmer1986; @Zelditch2012].
+To test whether workers were consistently measuring the same shape, we examined the per-worker consistency, as estimated by the morphological disparity [Procrustes variance; @Zelditch2012] of each worker's measured shapes.
+We then summarized the consistency within groups, and compared the median consistency of turkers and experts.
+To determine whether turkers improved with experience, we excluded the three images that turkers worked on, and calculated thse distance between their mean shape and the mean shape of experts.
+We then repeated this, but without excluding the first three images that turkers digitized.
+To determine whether turkers worked faster with experience, we compared the time it took turkers to complete their first image compared to their fifth image.
 
@@
 Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy.
+Because the specimens digitized in this study varied in size, we also report turker accuracy as both distance in millimeters and as a fraction of the specimen's total length (TL).
 We then examined the variance in turker landmarks.
@@
 
-To determine whether turkers and experts were statistically distinguishable, we performed a non-parametric MANOVA using the randomized residual permutation procedure (RRPP) with 1,000 iterations [@Collyer2014].
+To determine whether turkers and experts were statistically distinguishable, we performed a non-parametric MANOVA using the randomized residual permutation procedure (RRPP) with 1,000 iterations [@Collyer2015].
 The RRPP method reduces the effect of the "curse of dimensionality" ($p >> n$, where the number of predictors greatly exceeds the number of observations), a common problem in geometric morphometrics, and has been shown to have increased statistical power compared to a method where the raw data are randomized instead [@Anderson2003].
@@
 A common strategy in fish comparative studies is to examine evolutionary dynamics within a single family [@Ferry-Graham2001a; @Alfaro2005; @Alfaro2007; @Rocha2008; @Hernandez2009; @Dornburg2011; @Sorenson2013; @Santini2013; @Frederich2013; @Claverie2014; @Thacker2014], potentially due to the extensive amount of time necessary to collect data.
-To test whether our method can improve on the case where the data collection method is geometric morphometrics, we use the average time it took an expert to measure a single fish image and predict the time it would take for a single individual expert to measure all images at 5x replication, and compare it to the time it took turkers to collect these measurements at the same replication level.
+To demonstrate the utility of obtaining comparative data using our method, we use previously published phylogenies for seven fish families: Acanthuridae [@Sorenson2013], Balistoidae, Tetraodontidae [@Santini2013], Apogonidae, Chaetodontidae, Labridae [@Cowman2011; @Choat2012], and Pomacentridae [@Frederich2013].
+We match species in these phylogenies to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [Supplemental Table S5; @Parr2014].
+Crowdsourced workers placed landmarks describing body shape variation following a standard protocol (Supplemental Material) and were compensated \$0.15 per completed image.
+
+To test whether our method could be faster than a single expert digitizing a dataset, we extrapolated the time it would take for a single expert to measure all images at 1x replication, based on the average time an expert took to digitize a single image.
+We compared this predicted measurement time to the total time required for turkers to complete all digitization tasks at 5x replication, from initial upload to final submission.
 If the turkers in aggregate annotated images more quickly than a single expert would have, this suggests that the parallelization afforded by crowdsourcing is effective at reducing the total time required for data collection.
 
-To demonstrate the utility of obtaining comparative data using this method, we use previously published phylogenies for seven fish families: Acanthuridae [@Sorenson2013], Balistoidae, Tetraodontidae [@Santini2013], Apogonidae, Chaetodontidae, Labridae [@Cowman2011; @Choat2012], and Pomacentridae [@Frederich2013].
-We matched `r nrow(distinct(fish_families, tip))` species to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [Supplemental Table S1; @Parr2014].
-Crowdsourced workers placed landmarks describing body shape variation following a standard protocol (Supplementary Material).
-The Cartesian position of these landmarks were used in a generalized Procrustes analyses [@Gower1975; @Rohlf1990], which centers, scales, and rotates landmark configurations to minimize the least-squares distance between shapes.
+The Cartesian position of turker-collected landmarks were used in a generalized Procrustes analyses [@Gower1975; @Rohlf1990], which centers, scales, and rotates landmark configurations to minimize the least-squares distance between shapes.
 We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] with the R package *geomorph* `r citep(citation("geomorph"))`, and retain the principal component axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998] for all subsequent analyses.
 
-We used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of body shape evolution for all seven families.
-BAMM estimates the location of rate shifts in character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of lineage diversification and trait evolution.
-We assessed convergence and mixing using Tracer [@rambaut2007tracer].
+To illustrate the potential of how crowdsourcing could be integrated into an pipeline that could allow rapid collection and analysis of phenotypic data, we used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of body shape evolution for all seven families.
+BAMM estimates the location of rate shifts in character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of trait evolution.
+Any missing trait data is treated as a latent variable in the analysis.
+We assessed convergence and mixing using Tracer [@Rambaut2014].
 We also repeated each analysis and simulated under the prior (without data) to exclude rate heterogeneity that occurred solely due to stochastic processes.
-We use a Bayes Factor criterion of $BF > 5$ to enumerate the set of credible shifts [@Shi2015] and visualized them in R using BAMMtools `r citep(citation("BAMMtools"))`.
+We use a Bayes Factor criterion of $BF > 5$ to enumerate the set of credible shifts [@Shi2015] and visualized them using BAMMtools [@Rabosky2014c].
 
@@
 Points highlighted in red are those determined to be outliers (1.5 $\times$ IQR).
-See Supplemental Information for a labeled version of this figure.", anchor="figure", message=FALSE, warning=FALSE}
+A labeled version of this figure is available as Supplemental Figure S3.
+Photo credit J.E.
+Randall (used with permission under a CC-BY-NC 3.0 license).", anchor="figure", message=FALSE, warning=FALSE, fig.width = 5, fig.height = 5, fig.keep="last"}
 no_st <- fish_reliability %>% filter(role != "student") %>% mutate(role = factor(role))
@@
 
-ggplot(all_stats, aes(median, mad, color=outlier)) + geom_point() + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance / expert variance ") + facet_wrap(~family) + scale_color_manual(values = c("black", "red")) + theme_bw() + theme(legend.position="none") + xlim(0, 230)
-ggsave("turker_precision.pdf", width=14, height=14, scale = 1/3)
+## quantile(all_stats$median, seq(0,1,0.1))[9]
 
-# ggplot(highlighted, aes(median, mad)) + geom_point(color = "white") + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
-expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
-# ggsave("turker_precision_black.pdf", width=14, height=14, scale = 1/3)
-# 
-# 
-# highlighted <- all_stats %>% mutate(should_color = family == "Tetraodontidae" & str_detect(mark, "O"))
-# 
-# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad + 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
-expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
-# 
-# ggsave("turker_precision_black_puffers.pdf", width=14, height=14, scale = 1/3)
-# 
-# highlighted <- all_stats %>% mutate(should_color = family == "Chaetodontidae" & str_detect(mark, "[AD]3"))
-# 
-# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad - 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
-expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
-# 
-# ggsave("turker_precision_black_chaets.pdf", width=14, height=14, scale = 1/3)
-# 
-# highlighted <- all_stats %>% mutate(should_color = family == "Balistidae" & str_detect(mark, "D[12]"))
-# 
-# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad - 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
-expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
-# 
-# ggsave("turker_precision_black_triggers.pdf", width=14, height=14, scale = 1/3)
-# 
-# 
-# highlighted <- all_stats %>% mutate(should_color = median > 10)
-# 
-# ggplot(highlighted, aes(median, mad, color=should_color, label=mark)) + geom_point() + geom_text(data = filter(highlighted, should_color) %>% mutate(median = median + 4, mad = mad - 0.2)) + scale_color_manual(values = c("gray80", "red")) + labs(x="Accuracy: median turker - median expert", y="Precision: log turker variance ?
-expert variance ") + theme_black() + theme(legend.position = "none") + xlim(0, 130)
+plot_png = function(path, x, y, size)
+{
+  require('png')
+  logo <- readPNG(path, native=T) # read the file
+  dims<-dim(logo)[1:2] #number of x-y pixels for the logo (aspect ratio)
+  AR<-dims[1]/dims[2]
+  oldpar <- par(usr=c(0, 1, 0, 1))
+  rasterImage(logo, x-(size/2), y-(AR*size/2), x+(size/2), y+(AR*size/2), interpolate=TRUE)
+  par(oldpar)
+}
 
-# + geom_dl(aes(label=outlier_label), size=1, list("top.bumptwice", dl.trans(y=y+0.1), cex=0.75)) + scale_color_manual(values=c("black", "red")) + ylim(-max(abs(all_stats$mad) + 0.25), max(abs(all_stats$mad) + 0.25)) + theme("legend.position" = "none")
+plot.new()
+par(mfrow = c(3, 3))
+par(cex = 0.6)
+par(mar = c(0, 0, 0, 0), oma = c(4, 4, 0.5, 0.5))
+ctr <- 1
+xlim <- c(0, 240)
+ylim <- c(-2.4, 5.5)
+for (fam in unique(all_stats$family)) {
+  wanted <- all_stats %>% filter(family == fam)
+  plot(1, xlim = xlim, ylim = ylim, type = "n", axes = FALSE)
+  plot_png(sprintf("fish_cutouts/%s.png", fam), 0.75, 0.7, 0.5)
+  points(mad ~ median, data = wanted, pch = 16, col = ifelse(wanted$outlier, "red", "black"))
+  mtext(fam, side = 3, line = -1.5, cex = 0.6)
+  if (ctr %in% c(1, 4, 7)) {
+    axis(2)
+  }
+  if (ctr %in% 7:9) {
+    axis(1)
+  }
+  box(col = "grey60")
+  ctr <<- ctr + 1
+}
+mtext("Accuracy: median turker - median expert", side = 1, outer = TRUE, cex = 0.7, line = 2.2)
+mtext("Precision: log turker variance / expert variance", side = 2, outer = TRUE, cex = 0.7, line = 2.2)
+
 ```
 
-For nearly all landmarks, turkers only differ from the expert consensus by a few tens of pixels (Figure `r figr("turker_v_expert")`, Supplemental Figure S3).
+```{r pixel_conv, warning=FALSE}
+pixel_ratios <- read_csv("fish_reliability_images.csv") %>% transmute(Family, cm = TL / px * 10, frac = cm / TL)
+conv <- list(tenpx = range(round(pixel_ratios$cm * 10, 2)), frac = percent(range(round(pixel_ratios$frac, 3))))
+```
+
+For nearly 90% of the points measured, turkers differed from the expert consensus by less than 30 pixels, with half of all landmarks having less than 3 pixels of difference (10 px = `r conv$tenpx[1]`--`r conv$tenpx[2]` mm, `r conv$frac[1]`--`r conv$frac[2]` TL, Figure `r figr("turker_v_expert")`, Supplemental Figure S3, Supplemental Table S1).
 The most accurate and precise points are those that are related to the position of the eye (landmarks E1 and E2).
@@
 
+```{r turker_time, message=F, warning=FALSE}
+turker_time <- (only_good_points %>% group_by(role, family, sha1mac, sequence) %>% distinct %>% group_by(role, sequence) %>% summarise(time=median(duration)) %>% filter(role == "turker"))$time %>% round(2)
+```
+
 The inter-observer reliability of turkers and experts as measured by the ratio of among-individual and sum of the among-individual and measurement error ANOVA components was `r percent(reliability$turker)` and `r percent(reliability$expert)`, respectively.
 Although there is no current standard for acceptable levels of measurement reliability [@VonCramon-Taubadel2007], these percentages are not low enough to suggest weaknesses in the measurement protocol.
+Turkers were less consistent than the average expert (Supplemental Table S3); however, the overall difference in consistency between turkers and experts was generally quite small.
+We did not find evidence that turkers improved over time.
+Excluding the first three images did not markedly change turkers' performance compared to experts (Supplemental Table S4).
+Turkers took extra time to complete their first task, with a median completion time of `r turker_time[1]` minutes, compared to `r turker_time[5]` minutes on their fifth task.
 
@@
 The mean shape for all turkers and experts for a given family is the point outlined in black for each family, and connected with a black line to help emphasize the difference between turker and expert mean shapes.
-The convex hull for each family is drawn to show the amount of among-observer shape variation.", anchor="figure"}
+The convex hull for each family is drawn to show the amount of among-observer shape variation.
+The other PC axes are shown in Fig S4.", anchor="figure"}
 # Load data from the anonymized results file, remove students, drop bad marks
@@
 
+```{r tips_per_family}
+tips_per_family <- distinct(fish_families, tip) %>% group_by(family) %>% summarise(images=n()) %>% mutate(pretty_name = names(fish_phy), tips = lapply(fish_phy, function(x) length(x$tip.label)) %>% unlist, text = sprintf("%s: %d/%d", pretty_name, images, tips))
+```
+
+We were able to match `r nrow(distinct(fish_families, tip))` out of `r lapply(fish_phy, function(x) length(x$tip.label)) %>% unlist %>% sum` species to images in EOL's database (`r paste(tips_per_family$text, collapse=", ")`).
+Due to the low number of images matched for acanthurids, apogonids, and chaetodontids, we focused on the other four families with better taxon sampling for the comparative BAMM analysis.
+
 ```{r timing_plot, fig.cap="Line plot showing time to receive results for any given image (x axis) and the total fraction of the data set received (y axis).
-Landmarks were first received eight minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark.", anchor="figure"}
+Landmarks were first received eight minutes after creation of the Amazon MTurk task, and at least one replicate was received for every image at the 80 minute mark.", anchor="figure", fig.width = 5, fig.height = 4}
 # median time (in minutes) it takes an expert to measure these fishes
@@
 
-ggplot(timing_data, aes(time, frac)) + geom_line() + labs(x="Time taken (in minutes) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
+ggplot(timing_data, aes(time, frac)) + geom_line() + labs(x="Time taken (in min) to receive data for one unique replicate", y="Fraction of image set complete") + theme_minimal() + xlim(0, NA)
 
 total_turker_time <- max(fish_families$time_taken) %>% as.numeric %>% dminutes %>% round(1)
+
+worker_time_spent <- sum(distinct(fish_families, tip, time_taken)$worker_time_spent) %>% dminutes %>% round(1)
 ```
 
-Turkers in aggregate required `r as.character(total_turker_time)` to complete all `r nrow(distinct(fish_families, tip))` images at 5x replication `r figr("timing_plot")`.
+At 5x replication, `r as.character(total_turker_time)` elapsed between initial upload of the task to Amazon Mechanical Turk and submission of the last task by a turker (Figure `r figr("timing_plot")`).
 We estimate that a single expert would need `r as.character(morpho_per_image * nrow(distinct(fish_families, tip)))` to complete all images at 1x replication, extrapolated from a median expert time per image of `r as.character(morpho_per_image)`.
-At 5x replication, our projected expert would instead need `r as.character(morpho_per_image * nrow(fish_families))`.
+Our projected expert would need `r round(morpho_per_image * nrow(distinct(fish_families, tip)) * 5, 1) %>% as.character` were they to instead work at 5x replication.
 
@@
 The convex hull for each family is drawn to show area of morphospace occupied by each family.
-Figures for other PC axes are present in the Supplemental Material.", anchor="figure"}
+The other PC axes are shown in Supplemental Figures S5, S6.", anchor="figure"}
 name_map <- data_frame(family = c("acanthurid", "apogonidae", "balistoidae", "chaetodontidae", "labridae", "pomacentridae", "tetraodontidae"),
@@
 
-```{r}
-
-par_get_rtt_matrix <- function(ephy, start.time=NULL, end.time=NULL, nslices=100, node=NULL, nodetype = 'include', mc.cores = getOption("mc.cores", 2L)) {
-
-	if (!'bammdata' %in% class(ephy)) {
-		stop("Object ephy must be of class 'bammdata'\n");
-	}
-  if (ephy$type == 'diversification') {
-	  stop("'diversification' not supported")
-	}
-
-	if (is.null(node)) {
-		nodeset <- ephy$edge[,2];
-	} else if (!is.null(node) & nodetype == 'include') {
-		#nodeset <- getDesc(ephy, node)$desc_set;
-		nodeset <- unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set))
-	} else if (!is.null(node) & nodetype == 'exclude') {
-		nodeset <- setdiff( ephy$edge[,2], unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set)));
-	} else {
-		stop('error in getRateThroughTimeMatrix\n');
-	}
-
-	bt <- branching.times(as.phylo.bammdata(ephy));
-	maxpossible <- max(bt[as.character(intersect(nodeset, ephy$edge[,1]))]);
-
-	if (is.null(start.time)) {
-		start.time <- max(bt) - maxpossible;
-	}
-	if (is.null(end.time)) {
-		end.time <- max(bt);
-	}
-
-	decimals = function(x){
-	if(x%%1 != 0)	{
-		return(nchar(strsplit(as.character(x),".",fixed=TRUE)[[1]][[2]]));
-	}		else	{
-		return(10);
-	}
-}
-
-	tvec <- seq(start.time, end.time, length.out= nslices);
-	tol = 1*10^-decimals(ephy$eventBranchSegs[[1]][1,2]);
-	mm <- matrix(NA, nrow=length(ephy$eventBranchSegs), ncol=length(tvec));
-
-	# this is much faster than 'safeCompare'.
-	goodTime <- function (vec, val, tol) {
-		(vec[,2] - val <= tol) & (val - vec[,3] <= tol);
-	}
-
-	res <- mclapply(mc.cores = mc.cores, 1:length(ephy$eventBranchSegs), function(i) {
-	  	es <- ephy$eventBranchSegs[[i]];
-		events <- ephy$eventData[[i]];
-
-		isGoodNode <- rep(TRUE, nrow(es));
-
-		sapply(1:length(tvec), function(k) {
-			isGoodTime <- goodTime(es, tvec[k], tol=tol);
-			if (!(is.null(node))) { # only enter this if not the root. otherwise, only have to set once per i.
-				isGoodNode <- es[,1] %in% nodeset;	
-			}
-
-			estemp <- es[isGoodTime & isGoodNode, ];
-			tvv <- tvec[k] - events$time[estemp[,4]];
-			rates <- exponentialRate(tvv, events$lam1[estemp[,4]], events$lam2[estemp[,4]]);
-			mean(rates);
-		})
-	})
-
-	mm <- do.call(rbind, res)
-
-	obj <- list();
-	obj$beta <- mm;
-	obj$times <- tvec;
-
-	class(obj) <- 'bamm-ratematrix';
-	obj$type = 'trait';	
-	return(obj);
-}
-
-```
-
 ```{r load_bamm_data, message=FALSE, echo=FALSE, warning=FALSE, results="hide"}
@@
 base <- "../data-raw/bamm/"
-fams <- c("acanthurid", "apogonidae", "balistoidae", "chaetodontidae", "labridae", "pomacentridae", "tetraodontidae")
-pretty_names <- c("Acanthuridae", "Apogonidae", "Balistoidae", "Chaetodontidae", "Labridae", "Pomacentridae", "Tetraodontidae")
+fams <- c("balistoidae", "labridae", "pomacentridae", "tetraodontidae")
+pretty_names <- c("Balistoidae", "Labridae", "Pomacentridae", "Tetraodontidae")
 names(fams) <- pretty_names
 
-tre_trait <- lapply(paste0(base, "input_trees/", fams, ".tre"), read.tree)
+tre_trait <- fish_phy[pretty_names]
 names(tre_trait) <- fams
@@
 
-```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 across seven families of fishes.
+```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 across four families of fishes.
 (a) Phylorate plots color branch lengths by rates of shape evolution, where warmer colors indicate faster rates of evolution.
 Significant rate shift events (pp > 0.95) are indicated on the phylorate plot as a red circle on the corresponding branch.
-(b) Median log rates of shape evolution through time, where black lines indicate the background rate and red lines indicate the rate of phenotypic evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a).", fig.width = 6, fig.height = 3, out.height="3in", out.width="6in"}
-
-htext <- function(str) {
-  plot.new()
-  text(0.5, 0.5, str, cex=1)
-}
-vtext <- function(str) {
-  plot.new()
-  text(0.5, 0.5, str, cex=1, srt=90)
-}
+Black circles at the tips indicate the species that had shape data collected.
+(b) Median log rates of shape evolution through time, where black lines indicate the background rate and red lines indicate the rate of phenotypic evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a).
+The other three families are available in Supplemental Figure S7.", fig.width = 6, fig.height = 3, out.height="3in", out.width="6in"}
 
@@
 
-mantissa <- function(x) {
-  sapply(x, function(x) {
-    if (x == 0) 0
-    else {
-      log <- log10(abs(x))
-      10^(log - floor(log))
-    }
-  })
-}
-
-exponent <- function(x) {
-  sapply(x, function(x) {
-    if (x == 0) 0
-    else floor(log10(abs(x)))
-  })
-}
-
-plot_rtt <- function(rtt, shifts = list(), xlim=NULL, ylim=NULL, labels = TRUE) {
-  vartype <- if (rtt$type == "trait") "beta" else "lambda"
-  if (length(shifts) > 0) {
-    allshifts <- c(list(rtt), shifts)
-  } else {
-    allshifts <- list(rtt)
-  }
-  allshifts <- lapply(allshifts, function (x) {
-    x$avg <- apply(x[[vartype]], 2, median)
-    x$times <- rev(x$times)
-    x
-  })
-  xlim <- if (is.null(xlim)) c(0, max(sapply(allshifts, `[[`, "times"))) else xlim
-  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[[`, "avg")) else ylim
-  plot.new()
-  plot.window(rev(xlim), ylim, log = "y")
-  axis(at = axTicks(1), side = 1, labels = axTicks(1))
-  axis(at = c(1e-4, 1e-3, 1e-2, 1e-1), side = 2, las = 2, labels = labels)
-  lines(allshifts[[1]]$times, allshifts[[1]]$avg)
-  if (length(shifts) > 0) {
-    lapply(2:length(allshifts), function (x) {
-      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red", lwd=1.5)
-    })
-  }
-  return(invisible(NULL))
-}
-
 mar <- c(1.5,1,0,0)+0.1
@@
 layout_mat <- matrix(c(
-  8, 10, 12, 14, 16, 18, 20, 22, # shape phylorates
-  9, 11, 13, 15, 17, 19, 21, 23, # shape rate through time
-  0, 1, 2, 3, 4, 5, 6, 7
+  5, 7,  9, 11, 13, # shape phylorates
+  6, 8, 10, 12, 14, # shape rate through time
+  0, 1,  2,  3,  4
 ), nrow = 3, byrow = TRUE)
@@
   q <- plot.bammdata(evt_trait[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = interval)
+  tips <- unique(filter(fish_families, family == fam)$tip)
   best <- getBestShiftConfiguration(evt_trait[[fam]], prior = prior_trait[[fam]])
-  addBAMMshifts(best, cex = 2, par.reset=F)
+  tiplabels(tip = which(tre_trait[[fam]]$tip.label %in% tips), pch = 21, cex = 0.5, bg = "black", frame = "none")
+  addBAMMshifts(best, cex = 1.5, par.reset=F)
   #axis_phylo(label_ticks = FALSE)
@@
 Using the broken-stick method of determining a PCA stopping point, we analyzed PC 1 through PC `r stopping_pc`.
-We project per-species consensus shapes into Procrustes space (Figure `r figr("sevenfamily_tangent_plot")`, Supplemental Figure S5).
-The BAMMtools analysis uncovered heterogeneity in the rate of body shape evolution in each family (Figure `r figr("bamm_plots")`).
+We project per-species consensus shapes into Procrustes space (Figure `r figr("sevenfamily_tangent_plot")`, Supplemental Figure S5, S6).
+The BAMMtools analysis uncovered heterogeneity in the rate of body shape evolution in each family (Figure `r figr("bamm_plots")`, Supplemental Figure S7).
 Significant shifts in the rate of shape evolution were detected within two families: Labridae and Pomacentridae.
@@
 
-<!--
-Distal margin of the fin: would semilandmarks work better?
--->
-
 We have shown that crowdsourcing through Amazon Mechanical Turk is a tractable approach for generating reliable trait data at an unprecedented scale.
@@
 We found that for certain landmarks there was significant between and within group disagreement.
-Based on median average deviation, points belonging to the opercular series and those that locating the distal margin of the dorsal and anal fins were particularly challenging, compared to the experts.
+Points belonging to the opercular series and those that locating the distal margin of the dorsal and anal fins were particularly challenging, compared to the experts.
 Based on these results, nonspecialist turkers are unlikely to replace experts for all morphometric tasks.
@@
 
-## Reliability
+## Reliability of crowdsourced workers
 
 One advantage of the crowdsourced method we develop here is that inter-observer error can be readily assessed.
-Traditional geometric morphometric studies often rely on a single observer for practical reasons (the pool of trained geometric morphometricians is limited), and to avoid individually-driven systematic biases in data collection.
+Traditional geometric morphometric studies often rely on a single observer for practical reasons, as the pool of trained geometric morphometricians is limited, to ensure accurate comparisons of the same landmark across specimens, and to avoid individually-driven systematic biases in data collection.
 Although this common practice may reduce bias, it also precludes meaningful assessment of differences among observers.
@@
 For example, the landmarks that describe the shape of the fish's caudal fin asked workers to mark the distal tip of the first principal fin ray.
-Even when turkers are armed with a definition and a comparison between procurrent and principal fin rays, the experts' experience and training allow them to substantially outperform turkers in identifying this point.
+Even when turkers are armed with a definition and a comparison between procurrent and principal fin rays, the experts' experience and training allowed them to substantially outperform turkers in identifying this point.
 Furthermore, experts generally had lower disagreement in their landmark placement when compared to turkers, even for landmarks that turkers found especially difficult.
@@
 For example, @Brusatte2014 used a 853 character discrete character matrix for 150 taxa to estimate the rate of morphological evolution in the transition from theropod dinosaurs to modern birds.
-These data were collected over the course of 20 years as part of the Therapod Working Group [@dryad_84t75].
+These data were collected over the course of 20 years as part of the Therapod Working Group [@Brusatte2014].
 @OLeary2013 combined the work of MorphoBank contributors [@OLeary2011] with literature review to generate 4,541 characters for 86 species.
@@
 Natural language processing of the scientific literature could potentially be used for automatic extraction of morphological characters using DeepDive [@Peters2014; @Shin2015], but it may require impractically large corpus sizes [@Brill2003; @Halevy2009].
-Crowdsourcing can augment and enhance these statistical techniques.
+Instead of using any one method exclusively, crowdsourcing can augment and enhance these statistical techniques.
 For example, the algorithm in @Corney2012 occasionally captures non-leaf objects and systematically underestimates leaf sizes.
@@
 
-A third alternative to using expert morphologists and crowdsourced workers to collect data is through citizen science.
+A third alternative to using expert morphologists and crowdsourced workers is to collect data through citizen science.
 Citizen scientists are enthusiasts that volunteer to collect data or contribute annotations to a scientific endeavor.
@@
 For example, a study comparing citizen scientists and MTurk workers showed that for an image segmentation task MTurk workers had higher throughput and comparable accuracy to citizen scientists, but MTurk workers performed poorly when asked to identify birds to the species level [@VanHorn2015].
+Volunteer citizen scientists can be inexpensive to use, but the pool of available MTurk workers is likely much larger.
+This larger participant pool means that tasks can be completed much faster due to the ability of multiple individuals to work in parallel; the financial motivation additionally ensures that higher-paying tasks are completed more quickly [@Ipeirotis2010; @Mason2012a].
+Balancing the desired speed and quality of results, and the cost of data collection will be an important consideration for any future study using crowdsourcing.
 
@@
 Although there are limitations in the type and accuracy of data that can be collected through MTurk crowdsourcing, even a simplified protocol can produce meaningful biological results that are concordant with previous hypotheses in these groups.
-We detected a significant shift in the rate of body shape evolution in Labridae, restricted to the wrasse tribes Labrini, Cheilini, and Scarini.
+Despite our low sampling fraction, we detected a significant shift in the rate of body shape evolution in Labridae, restricted to the wrasse tribes Labrini, Cheilini, and Scarini.
 The scarines and cheilines are mostly reef-associated [@Froese2014], which has been proposed as an environment that drives diversification rate changes in marine teleosts [@Alfaro2007; @Cowman2011; @Price2011].
@@
 Although the example we present here was necessarily limited, extending this technique to generate new phenotypic datasets for existing large phylogenetic trees such as fishes [@Rabosky2013], birds [@Jetz2012], mammals [@Bininda-Emonds2007], and angiosperms [@Zanne2014] would be straightforward, especially for taxa where image data are already aggregated in a database such as FishBase [@Froese2014] or the Encyclopedia of Life [@Parr2014].
+
+## Future challenges for generating massive phenotypic datasets
 
@@
 Researchers do not have to spend time digitizing collections, workers need not generate biological hypotheses, and biologists will not have to solve open questions in the fields of machine vision and natural language processing in order to answer questions in comparative biology.
-The task of phenomic-scale data collection is split up and efficiently allocated according to the strengths of each role, without overly relying on any one axis to carry out the entire task.
+The task of phenomic-scale data collection is split up and efficiently allocated according to the strengths of each role, without overly relying on any single role to carry out the entire task.
+
+Though we have shown that crowdsourcing can increase these speed of data collection, we are still dependent on high quality image data sets, as evidenced by our low sampling fraction for three of the seven families analyzed.
+The problem of difficult-to-retrieve *dark data* is well-known [@Heidorn2008], but without either physical access to the collections or an image of the specimen, morphological data is impossible to acquire.
+The need to collect, identify, photograph, and publish specimen images remains as another obstacle to high-throughput phenotyping.
+Efforts are underway to digitize more biodiversity resources, such as the National Science Foundation's iDigBio initiative (<https://www.idigbio.org>) in the U.S.
+and the Natural History Museum's iCollections project (<http://www.nhm.ac.uk/our-science/our-work/digital-museum/digital-collections-programme.html>) in the U.K.
+Whole-drawer imaging of insect collections and scanning of herbarium pressings are already well underway, but one future direction would be to expand this to other avenues: skeletal imaging with radiographs, 3D morphometrics using laser or CT scanning, of both fossils and extant organisms.
+Much work and engineering expertise will be required to extend our framework into the physical world to further streamline data collection, but these efforts will likely result in a huge increase in the quality and quantity of phenotypic data.
 
@@
 The burden is now on experienced taxonomists and morphologists to create protocols that are simple enough to be understood by MTurk workers, but comprehensive enough to test hypotheses of interest across the tree of life.
-Additionally, museums and other institutions must increase their efforts to make their biodiversity collections available digitally, including images suitable for morphological research.
-The problem of difficult-to-retrieve *dark data* is well-known [@Heidorn2008], but without either physical access to the collections or an image of the specimen, morphological data is impossible to acquire.
 
 Our results suggest that, where possible, crowdsourcing should be an integral part of any large-scale morphological analysis.
-Crowdsourcing should play a key role in unlocking the "dark data" present in biodiversity collections by providing a high-throughput way to extract the phenotypic data present in specimens.
+Crowdsourcing can play a key role in unlocking the "dark data" present in biodiversity collections by providing a high-throughput way to extract the phenotypic data present in specimens.
 Furthermore, coordinating efforts from digitizing museum collections, natural language processing and machine vision software, citizen scientists, expert morphologists and taxonomists, and crowdsourced Mechanical Turk workers would result in an extremely powerful pipeline that could generate a "phenoscape" across the tree of life.
@@
 
-We thank XXX, YYY, and ZZZ for helpful comments on the manuscript, as well as T.
+We thank P.
+Chakrabarty and G.
+Thomas for helpful comments on the manuscript, as well as T.
 Marcroft, B.
@@
 
-All data are deposited online at the Encyclopedia of Life and Dryad.
+Data collected for this paper have been archived on Dryad doi:10.5061/dryad.gh4k7 `r citep(RefManageR::BibEntry(bibtype = "Online", title = "Data from: Crowdsourced geometric morphometrics enable rapid large-scale collection and analysis of phenotypic data", year = "2015", author = c(person("Jonathan", "Chang"), person("Michael E.", "Alfaro")), publisher = "Dryad Digital Repository", doi = "10.5061/dryad.gh4k7", url = "http://dx.doi.org/10.5061/dryad.gh4k7"))`.
+Source code is available on GitHub for the web interface (<https://github.com/jonchang/eol-mturk-landmark>), repeatability experiment (<https://github.com/jonchang/fake-mechanical-turk>), and this manuscript (<https://github.com/jonchang/fish.reliability>).
 
@@
 
-Conceived and designed the experiments: JC MEA. Performed the experiments: JC. Analyzed the data: JC. Contributed reagents/materials/analysis tools: JC MEA. Wrote the paper: JC MEA.
+Conceived and designed the experiments: JC MEA. Performed the experiments: JC. Analyzed the data: JC. Contributed reagents/materials/analysis tools: JC. Wrote the paper: JC MEA.
 
@@
 
-```{r knitcitations_write, cache=FALSE, message=FALSE}
+```{r knitcitations_write, cache=0, message=FALSE}
 write.bibtex(file="knitcitations.bib")

### 2016-08-05 manuscript/supplement.Rmd
Code drop

@@
 
+##
+
+```{r fish_images}
+read_csv("fish_reliability_images.csv") %>% select(Family, Species, Author, Rights, TL, px) %>% knitr::kable(caption = "Images digitized by turkers and experts to compare their performance.
+TL = total length of the specimen, in cm.
+PX = The total length of the specimen, in number of image pixels.")
+```
+
+```{r fish_images_urls}
+read_csv("fish_reliability_images.csv") %>% select(Family, URL) %>% knitr::kable(caption = "Online URLs of images from Supplemental Table S1")
+```
+
 ## Landmarks used
@@
 
-![Landmarks (J1) rostral tip of premaxilla (J2) ventral tip of premaxilla (J3) rostral tip of dentary (E1) anterior margin of midline through eye (E2) posterior margin of midline through eye (O1) dorsal end of preopercle (O2) ventral elbow of preopercle (O3) anterior end of preopercle (O4) dorsal end of opercle (O5) posterior end of opercle (D1) anterior insertion of dorsal fin (D2) distal tip of the anterior dorsal fin ray (D3) distal tip of the posterior dorsal fin ray (D4) posterior insertion of dorsal fin (P1) dorsal insertion of pectoral fin (P2) distal tip of the dorsal pectoral fin ray (P3) distal tip of the ventral pectoral fin ray (P4) ventral insertion of anal fin (A1) anterior insertion of anal fin (A2) distal tip of the anterior anal fin ray (A3) distal tip of the posterior anal fin ray (A4) posterior insertion of anal fin (C1) dorsal insertion of the caudal fin (C2) distal tip of the dorsal caudal fin ray (C3) distal tip of the the ventral caudal fin ray (C4) ventral insertion of the caudal fin (C5) midpoint of the caudal margin of the caudal peduncle.](landmark_figure/landmarks.pdf)
+![Description of landmarks used to digitize fish body shape.
+(J1) rostral tip of premaxilla (J2) ventral tip of premaxilla (J3) rostral tip of dentary (E1) anterior margin of midline through eye (E2) posterior margin of midline through eye (O1) dorsal end of preopercle (O2) ventral elbow of preopercle (O3) anterior end of preopercle (O4) dorsal end of opercle (O5) posterior end of opercle (D1) anterior insertion of dorsal fin (D2) distal tip of the anterior dorsal fin ray (D3) distal tip of the posterior dorsal fin ray (D4) posterior insertion of dorsal fin (P1) dorsal insertion of pectoral fin (P2) distal tip of the dorsal pectoral fin ray (P3) distal tip of the ventral pectoral fin ray (P4) ventral insertion of anal fin (A1) anterior insertion of anal fin (A2) distal tip of the anterior anal fin ray (A3) distal tip of the posterior anal fin ray (A4) posterior insertion of anal fin (C1) dorsal insertion of the caudal fin (C2) distal tip of the dorsal caudal fin ray (C3) distal tip of the the ventral caudal fin ray (C4) ventral insertion of the caudal fin (C5) midpoint of the caudal margin of the caudal peduncle.](landmark_figure/landmarks.pdf)
 
@@
 
-```{r geomorph_consistency, warning=FALSE}
+```{r geomorph_variance, warning=FALSE}
 only_good_points <- fish_reliability %>% filter(role != "student", !mark %in% c("A3", "A4", "D3", "D4", "O1", "O2", "O3","O4","O5", "P3", "P4"))
-res <- gpa_with_vectors(only_good_points, family, role, sha1mac, sequence)
+
+consist <- only_good_points %>% mutate(role = ifelse(role == "morphologist", "expert", role)) %>% group_by(family, role) %>% do(
+  (function(df) {
+    res <- gpa_with_vectors(df, family, role, sha1mac, sequence)
+    consistency <- morphol.disparity(res$coords, factor(res$vectors$sha1mac), iter=1)$Disp.obs[, 1]
+    fv <- structure(fivenum(consistency) * 10^6, .Names = c("minimum", "1st quartile", "median", "3rd quartile", "maximum"))
+    as.data.frame(t(fv))
+  })(.)
+)
+
+consist %>% knitr::kable(digits = 0, caption = "Five number summaries of turker and expert consistency. The summary statistics are multiplied by 1,000,000 to facilitate comparisons.")
+
+morph_better <- (consist %>% group_by(family) %>% select(role, median) %>% spread(role, median) %>% transmute(morph_better = expert < turker))$morph_better
+```
+
+## Do turkers improve with experience?
+
+```{r geomorph_lm, warning=FALSE}
+res <- gpa_with_vectors(filter(only_good_points, sequence > 2), family, role, sha1mac)
 
@@
 role <- res$vectors$role
-repno <- as.integer(res$vectors$sequence) %% 5
-sha1mac <- as.factor(res$vectors$sha1mac)
 
-lm_result <- advanced.procD.lm(res$coords ~ family*role*sha1mac, ~family*sha1mac, .parallel.cores = 24, .parallel.cluster = parallel::makeForkCluster(24))
-lm_p <- lm_result$P[2]
-lm_Z <- lm_result$Z[2]
-lm_F <- lm_result$F[2]
-rm(family, role, repno, sha1mac)
+result_no_seq0 <- advanced.procD.lm(res$coords ~ family * role, ~ family, groups = ~family + role, .parallel.cores = 24, iter=1)
+
+res <- gpa_with_vectors(only_good_points, family, role, sha1mac)
+
+family <- res$vectors$family
+role <- res$vectors$role
+
+result_with_seq0 <- advanced.procD.lm(res$coords ~ family * role, ~ family, groups = ~family + role, .parallel.cores = 24, iter=1)
+
+allX <- seq(1, 17, by=2)
+
+fams <- c("Acanthuridae","Apogonoidae","Balistidae","Chaetodontidae","Gobiidae","Labridae","Pomacanthidae","Scorpaenidae","Tetraodontidae")
+
+seq0 <- unlist(lapply(allX, function(ii) (result_with_seq0$Means.dist)[ii, ii+1, drop=FALSE]))
+no_seq0 <- unlist(lapply(allX, function(ii) (result_no_seq0$Means.dist)[ii, ii+1, drop=FALSE]))
+
+tabs <- data_frame(`Family` = fams, `Procrustes distance: full dataset` = seq0, `Reduced dataset` = no_seq0, `Ratio` = seq0 / no_seq0)
+
+tabs %>% knitr::kable(digits = 5, caption = "Comparison of the Procrustes distance between the mean turker shape and the mean expert shape, for a full dataset, and a dataset excluding the first three images that turkers worked on.
+The ratio is computed by dividing the full dataset's distance by the reduced dataset's distance, in order to compare the relative distance change among the different images digitized.")
+
 ```
-
-We also used a non-parametric MANOVA with RRPP to determine whether turkers were less consistent among individual replicates as compared to experts.
-The non-parametric MANOVA with RRPP failed to detect a significant difference between turker and expert shapes ($p = `r lm_p`$, $Z = `r lm_Z`$, $F=`r lm_F`$).
 
@@
 
-```{r turker_morph_procrustes_other_pc34, message=FALSE, warning=FALSE, fig.cap="PC3 and PC4 for turker / morphologist Procrustes space"}
+```{r turker_morph_procrustes_other_pc34, message=FALSE, warning=FALSE, fig.cap="Morphospace projection of PC3 and PC4 for each observer's mean shape.
+Blue points indicate experts, while red points indicate turkers.
+The mean shape for all turkers and experts for a given family is the point outlined in black for each family, and connected with a black line to help emphasize the difference between turker and expert mean shapes.
+The convex hull for each family is drawn to show the amount of among-observer shape variation.
+PC 1 and 2 are shown in Fig 2."}
 turk_morph_pca <- read_csv("turk_morph_pca_for_supplement.csv")
@@
 
-```{r sevenfamily_extra_pc, message=FALSE, warning=FALSE, fig.cap = "PC3-6 for seven families of ray finned fishes."}
+```{r sevenfamily_extra_pc, message=FALSE, warning=FALSE, fig.cap = "Morphospace of PC3 and PC4 for seven families of ray-finned fishes.
+Each point indicates a separate species; families are separated by colors.
+The convex hull for each family is drawn to show area of morphospace occupied by each family.
+The other PC axes are shown in Figs 4 and S6."}
 pcaed <- read_csv("seven_family_pca_for_supplement.csv")
 plot_tangent_space2(pcaed, "PC3", "PC4")
+```
+
+```{r sevenfamily_extra_pc2, message=FALSE, warning=FALSE, fig.cap = "Morphospace of PC5 and PC6 for seven families of ray-finned fishes.
+Each point indicates a separate species; families are separated by colors.
+The convex hull for each family is drawn to show area of morphospace occupied by each family.
+The other PC axes are shown in Figs 4 and S5."}
 plot_tangent_space2(pcaed, "PC5", "PC6")
+```
+
+```{r load_bamm_data, message=FALSE, echo=FALSE, warning=FALSE, results="hide"}
+#base <- "data-raw/bamm/"
+base <- "../data-raw/bamm/"
+fams <- c("acanthurid", "apogonidae", "chaetodontidae")
+pretty_names <- c("Acanthuridae", "Apogonidae", "Chaetodontidae")
+names(fams) <- pretty_names
+
+tre_trait <- fish_phy[pretty_names]
+names(tre_trait) <- fams
+
+cl <- makePSOCKcluster(detectCores(), useXDR=F)
+clusterExport(cl, list("base", "fams", "tre_trait"))
+evt_trait <- parLapplyLB(cl, fams, function(fam) {
+  tre <- tre_trait[[fam]]
+  BAMMtools::getEventData(tre, paste0(base, "trait_runs/", fam, "_PC1_event_data.txt"), burnin=0.1, type="trait")
+})
+names(evt_trait) <- fams
+clusterExport(cl, list("evt_trait"))
+rtt_trait <- parLapplyLB(cl, fams, function(fam) {
+  BAMMtools::getRateThroughTimeMatrix(evt_trait[[fam]])
+})
+prior_trait <- parLapplyLB(cl, fams, function(fam) {
+  BAMMtools::getBranchShiftPriors(tre_trait[[fam]], paste0(base, "trait_runs/", fam, "_PC1_prior_probs.txt"))
+})
+
+names(pretty_names) <- names(prior_trait) <- names(rtt_trait) <- fams
+stopCluster(cl)
+```
+
+```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 across three families of fishes.
+(a) Phylorate plots color branch lengths by rates of shape evolution, where warmer colors indicate faster rates of evolution.
+No significant rate shift events were detected within these families.
+(b) Median log rates of shape evolution through time.
+Analysis for the other four families are available in the main text, Figure 5.", fig.width = 6, fig.height = 3, out.height="3in", out.width="6in"}
+
+customintervals <- list(
+  pomacentridae = c(0.000, 0.0025)
+)
+
+mar <- c(1.5,1,0,0)+0.1
+par(mar = c(0,0,0,0))
+
+layout_mat <- matrix(c(
+  4, 6, 8, 10, # shape phylorates
+  5, 7, 9, 11, # shape rate through time
+  0, 1, 2, 3
+), nrow = 3, byrow = TRUE)
+
+layout(layout_mat, widths = c(1.3, 12, 9, 9, 9, 9, 9, 9), heights = c(5, 6, 1))
+lapply(pretty_names, htext) # family names
+vtext("(a) Shape Phylorates")
+vtext("(b) Shape Rate Through Time")
+par(mar = mar)
+
+ctr <- 1
+lapply(fams, function(fam) {
+  par(mar=mar)
+  if (ctr == 1) {
+    par(mar=c(1.5,4,0,0)+0.1)
+  }
+  interval <- if(is.null(customintervals[[fam]])) c(0, 0.001) else customintervals[[fam]] 
+  q <- plot.bammdata(evt_trait[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = interval)
+  tips <- unique(filter(fish_families, family == fam)$tip)
+  best <- getBestShiftConfiguration(evt_trait[[fam]], prior = prior_trait[[fam]])
+  tiplabels(tip = which(tre_trait[[fam]]$tip.label %in% tips), pch = 21, cex = 0.5, bg = "black", frame = "none")
+  addBAMMshifts(best, cex = 1.5, par.reset=F)
+  #axis_phylo(label_ticks = FALSE)
+  axisPhylo(col.axis = "transparent")
+  shift_node <- getShiftNodesFromIndex(best, 1)
+  if (length(shift_node) > 0) {
+    background_rates <- par_get_rtt_matrix(evt_trait[[fam]], node = shift_node, nodetype = "exclude", start.time=0)
+    clade_rates <- lapply(shift_node, function (x) {
+      phy <- extract.clade(tre_trait[[fam]], x)
+      write.tree(phy, file = paste0(fam, "_trait_shift_", x, ".tre"))
+      par_get_rtt_matrix(evt_trait[[fam]], node = x)
+      })
+    plot_rtt(background_rates, clade_rates, ylim=c(4e-5, 0.1), labels = ctr == 1)
+  } else {
+    plot_rtt(rtt_trait[[fam]], ylim=c(4e-5, 0.1), labels = ctr == 1)
+  }
+  ctr <<- ctr + 1
+})
+
 ```
@@
 fish_families_urls <- read_csv("fish_families_urls.csv")
-knitr::kable(select(fish_families_urls, -eol_guid), caption='Families, species names, and links to the images hosted on Encyclopedia of Life for the section "Example: a phenomic pipeline for comparative phylogenetic analysis".')
+knitr::kable(select(fish_families_urls, -eol_guid), caption='Families, species names, and URLs of the images hosted on Encyclopedia of Life for the section "Example: a phenomic pipeline for comparative phylogenetic analysis".')
 ```

### 2016-08-05 R/plotting.R

@@ -126,3 +126,53 @@ plotColorRamp <- function(q) {
   axis(2, round(seq(0, max(y), length.out = 3), 0), las = 1, yaxs = "i", cex.axis = 1, tcl = NA, mgp = c(0, 0.25, 0))
 }
 
+
+#' Plot BAMM rate through time
+#' @param rtt Rate through tiem matrix
+#' @param shifts list of other rate through time matrices corresponding to certain nodes
+#' @param xlim vector of length 2, x limits of plot
+#' @param ylim vector of length 2, y limits of plot
+#' @param labels whether to plot axis labels
+#' @export
+plot_rtt <- function(rtt, shifts = list(), xlim = NULL, ylim = NULL, labels = TRUE) {
+  vartype <- if (rtt$type == "trait") "beta" else "lambda"
+  if (length(shifts) > 0) {
+    allshifts <- c(list(rtt), shifts)
+  } else {
+    allshifts <- list(rtt)
+  }
+  allshifts <- lapply(allshifts, function (x) {
+    x$avg <- apply(x[[vartype]], 2, median)
+    x$times <- rev(x$times)
+    x
+  })
+  xlim <- if (is.null(xlim)) c(0, max(sapply(allshifts, `[[`, "times"))) else xlim
+  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[[`, "avg")) else ylim
+  plot.new()
+  plot.window(rev(xlim), ylim, log = "y")
+  axis(at = axTicks(1), side = 1, labels = axTicks(1))
+  axis(at = c(1e-4, 1e-3, 1e-2, 1e-1), side = 2, las = 2, labels = labels)
+  lines(allshifts[[1]]$times, allshifts[[1]]$avg)
+  if (length(shifts) > 0) {
+    lapply(2:length(allshifts), function (x) {
+      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red", lwd=1.5)
+    })
+  }
+  return(invisible(NULL))
+}
+
+#' Plot horizontal or vertical text
+#' @param str string to plot
+#' @export
+htext <- function(str) {
+  plot.new()
+  text(0.5, 0.5, str, cex=1)
+}
+
+#' @export
+#' @rdname htext
+vtext <- function(str) {
+  plot.new()
+  text(0.5, 0.5, str, cex=1, srt=90)
+}
+

### 2016-08-05 R/utils.R

@@ -84,3 +84,88 @@ make_sampling_file <- function(sampling, file="") {
   write.table(sampling$clade, file, append = TRUE, quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)
   invisible(NULL)
 }
+
+
+#' Parallel version of BAMMtools getRateThroughTimeMatrix
+#'
+#' @param ... other parameters, see \link[BAMMtools]{getRateThroughTimeMatrix}
+#' @param mc.cores number of cores to initialize a cluster with
+#' @export
+par_get_rtt_matrix <- function(ephy, start.time=NULL, end.time=NULL, nslices=100, node=NULL, nodetype = 'include', mc.cores = getOption("mc.cores", 2L)) {
+
+  if (!'bammdata' %in% class(ephy)) {
+    stop("Object ephy must be of class 'bammdata'\n");
+  }
+  if (ephy$type == 'diversification') {
+    stop("'diversification' not supported")
+  }
+
+
+  if (is.null(node)) {
+    nodeset <- ephy$edge[,2];
+  } else if (!is.null(node) & nodetype == 'include') {
+    #nodeset <- getDesc(ephy, node)$desc_set;
+    nodeset <- unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set))
+  } else if (!is.null(node) & nodetype == 'exclude') {
+    nodeset <- setdiff( ephy$edge[,2], unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set)));
+  } else {
+    stop('error in getRateThroughTimeMatrix\n');
+  }
+
+  bt <- branching.times(as.phylo.bammdata(ephy));
+  maxpossible <- max(bt[as.character(intersect(nodeset, ephy$edge[,1]))]);
+
+  if (is.null(start.time)) {
+    start.time <- max(bt) - maxpossible;
+  }
+  if (is.null(end.time)) {
+    end.time <- max(bt);
+  }
+
+  decimals = function(x){
+    if(x%%1 != 0)	{
+      return(nchar(strsplit(as.character(x),".",fixed=TRUE)[[1]][[2]]));
+    }		else	{
+      return(10);
+    }
+  }
+
+  tvec <- seq(start.time, end.time, length.out= nslices);
+  tol = 1*10^-decimals(ephy$eventBranchSegs[[1]][1,2]);
[…]
